---
title: "Setup Wizard: Non-Technical System Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup Wizard Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# ZZedc Setup Wizard: Getting Started in 5 Minutes

The ZZedc Setup Wizard makes it easy to configure your system **without programming knowledge**. This guide walks you through each step.

## What You'll Need

- Your study information (name, protocol ID, principal investigator)
- An administrator username and password
- Email address for the admin account
- Names and email addresses of your team members (optional)

**Time required: 5 minutes**

---

## Starting the Setup Wizard

```{r launch_wizard}
library(zzedc)

# Launch ZZedc with the setup wizard
launch_zzedc_with_wizard()

# Your browser opens with the Setup Wizard
```

You'll see a progress bar at the top showing "Step 1 of 5".

---

## Step 1: Basic Information (30 seconds)

Enter your research study details:

### Fields to Complete

- **Study Name**: The official name of your research project
  - Example: "Depression Treatment Trial v2.1"
  - This appears in reports and user-facing screens

- **Protocol ID**: Your institutional protocol number
  - Example: "DEPR-2024-001"
  - Used for regulatory documentation

- **Principal Investigator Name**: Full name of the study lead
  - Example: "Dr. Jane Smith"

- **PI Email**: Contact email for the principal investigator
  - Example: "jane.smith@university.edu"
  - Used for system notifications

- **Target Enrollment**: How many participants you plan to enroll
  - Example: 50
  - Used for progress tracking

- **Study Phase**: Select from the dropdown
  - Pilot, Phase 1, Phase 2, Phase 3, Phase 4, or Observational
  - Used for categorization

### Example Entry

```
Study Name: Depression Treatment Trial v2.1
Protocol ID: DEPR-2024-001
PI Name: Dr. Jane Smith
PI Email: jane.smith@university.edu
Target Enrollment: 50
Study Phase: Phase 2
```

Click **Next →** to continue.

---

## Step 2: Administrator Account (1 minute)

Create your administrator login credentials. **This is the account you'll use to manage ZZedc.**

### Fields to Complete

- **Username**: Your login name (no spaces, no special characters)
  - Example: "jane_smith" or "jsmith"
  - Keep it simple and memorable

- **Full Name**: Your full name
  - Example: "Dr. Jane Smith"
  - Used in audit logs and system messages

- **Email**: Your email address
  - Example: "jane.smith@university.edu"
  - Used for password recovery

- **Password**: Create a strong password
  - Minimum 8 characters
  - Mix letters, numbers, and symbols
  - Example: "UniquePass2024!"

- **Confirm Password**: Re-type your password

### Password Strength Tips

| ❌ Weak | ✅ Strong |
|----------|----------|
| "password" | "Pass2024Study!" |
| "12345678" | "Research&Safe99" |
| "MyStudy" | "TrialData#2024" |
| "admin123" | "ZZedc@Research2024" |

Click **Next →** to continue.

---

## Step 3: Security Configuration (2 minutes)

Configure security settings for your system. **You only do this once.**

### Security Salt

A unique security salt has been generated for your system. **Save this in a secure place** — it encrypts all passwords.

```
Generated Salt: a7f9e2c4d1b3f8a6e9d2c5b8f1a4e7d0
```

**Ways to save it:**
1. Copy to a password manager (recommended)
2. Write in a notebook kept in a secure location
3. Email to yourself in an encrypted message

**Important:** If you lose this salt, you cannot recover it. Your passwords will not work without it.

### Session Configuration

- **Session Timeout (minutes)**: How long before inactive users are logged out
  - Recommended: 30 minutes
  - Shorter = more secure, longer = less hassle

- **Enforce HTTPS**: Require secure connections (recommended for production)
  - Yes = Only secure connections work
  - No = Development use only

- **Max Failed Login Attempts**: How many wrong passwords before account lockout
  - Recommended: 3
  - Users must contact admin if locked out

### Example Settings

```
Session Timeout: 30 minutes
Enforce HTTPS: Yes (recommended)
Max Failed Login Attempts: 3
```

**Checkbox**: Confirm that you have saved the security salt ✓

Click **Next →** to continue.

---

## Step 4: Team Members (Optional, 1 minute)

Add your research team members. **You can skip this and add them later.**

### Adding Team Members

1. Enter username: "alex_r" or "maria_santos"
2. Enter full name: "Alex Rodriguez" or "Maria Santos"
3. Enter email: "arodriguez@university.edu"
4. Select role:
   - **Coordinator**: Data entry, basic reports
   - **PI**: Full access (usually reserved for study leads)
   - **Data Manager**: Data management and exports
   - **Monitor**: Read-only access for auditing

5. Click **Add Member**

### Example Team

| Username | Name | Email | Role |
|----------|------|-------|------|
| coord_sarah | Sarah Chen | schen@university.edu | Coordinator |
| dm_alex | Alex Rodriguez | arodriguez@university.edu | Data Manager |
| monitor_john | John Monitor | jmonitor@university.edu | Monitor |

**Note:** You'll set their passwords on first login. For now, temporary passwords are generated automatically.

Click **Next →** to continue.

---

## Step 5: Data Dictionary & Forms (1 minute)

Choose how to set up your data collection forms.

### Four Options

#### 1. **Start with Blank Forms** (Easiest for new users)
- Creates an empty system
- You design forms in the Data Dictionary
- Best for: Custom studies, complex workflows
- Setup time: 5 minutes
- Design time: 30 minutes to 2 hours

#### 2. **Use Example ADHD Forms** (Recommended for testing)
- Pre-built clinical trial forms
- Full validation rules included
- Best for: Learning ZZedc, testing features
- Ready immediately, then customize

#### 3. **Import from CSV File** (For spreadsheet templates)
- Upload your existing data dictionary as CSV
- Supports: Excel saved as CSV
- Format:
  ```
  variable_name,variable_label,variable_type,valid_values
  subject_id,Subject ID,text,required
  age,Age (years),integer,between 18 and 120
  ```

#### 4. **Import from Google Sheets** (For team collaboration)
- Share Google Sheets URL
- Multiple people can edit
- Auto-sync with ZZedc
- Best for: Teams, complex templates

### Our Recommendation

**Start with ADHD Examples** ← Click this option
- See a complete working system
- Learn how ZZedc works
- Import your own data later
- Can always start over

Click **Create System** to finish.

---

## Success Screen ✓

When setup is complete, you'll see:

```
✓ Success! Your ZZedc system has been created
```

You'll receive:
- **config.yml** - System configuration file (automatically created)
- **launch_app.R** - Script to start your system (automatically created)
- Database with all users and tables

---

## Next Steps

### 1. Launch Your System

Option A: Click **Launch ZZedc** button
```r
launch_zzedc()
```

Option B: Run from command line
```bash
Rscript launch_app.R
```

### 2. Login

Use the administrator account you created:
- Username: `jane_smith` (your username)
- Password: `UniquePass2024!` (your password)

### 3. Welcome!

You'll see:
- Dashboard with system status
- Data entry form (EDC tab)
- Reports (Reports tab)
- Admin dashboard (for user management, backups, audit logs)

### 4. Add Your Team

In the **Admin Dashboard** → **👤 User Management**:
1. Click "Add User"
2. Enter team member information
3. System generates temporary password
4. Share credentials with team member

### 5. Create Forms

In the **Admin Dashboard** → **⚙️ System Configuration**:
1. Access form builder
2. Define your data collection fields
3. Add validation rules
4. Assign to forms

### 6. Start Collecting Data

Team members log in and begin entering data. Validation happens in real-time.

---

## Troubleshooting

### "Database already exists"
- Your system has been set up before
- To start over: Delete database file in `data/` folder
- Run setup wizard again

### "Session timeout is too short"
- Edit `config.yml` and change `session_timeout_minutes`
- Or use Admin Dashboard → System Configuration

### "Lost the security salt"
- **Critical:** Cannot be recovered
- Only option: Delete database and start over
- Future: Store salt in password manager

### "Need to change admin password"
- Use Admin Dashboard → User Management
- Find admin account
- Click "Reset Password"
- System generates temporary password

### "Want to add/remove team members"
- Use Admin Dashboard → User Management
- Add: Click "Add User"
- Remove: Click "Deactivate"

---

## Security Reminders

| Item | Action |
|------|--------|
| Security Salt | Store in secure location (not email!) |
| Admin Password | Share with no one, only you use |
| Session Timeout | 30 minutes default (auto logout) |
| HTTPS | Enable for production (production=cloud, development=local) |
| Backups | Automatic daily backups (can disable if needed) |

---

## Keeping It Simple

✅ **Recommended Setup**
- Use ADHD example forms to start
- Set session timeout to 30 minutes
- Enable HTTPS if on the internet
- Keep 30 days of automatic backups
- Add team members as needed

❌ **Don't Worry About**
- Advanced compliance settings (GDPR/CFR) for now
- Custom validation rules initially
- Complex multi-site configurations
- API integrations yet

---

## Next Steps After Setup

1. **Read the Quick Start Guide** - 5-minute overview
2. **Check the User Guide** - Complete feature documentation
3. **Try the Data Entry** - Enter sample data, see validation
4. **Review Reports** - See auto-generated analytics
5. **Import Your Forms** - Replace ADHD forms with your study

---

## Getting Help

- **Email:** rgthomas@ucsd.edu
- **GitHub:** https://github.com/rgt47/zzedc
- **Documentation:** All vignettes available in help menu
- **Admin Dashboard:** Click "❓ Help & Documentation" tab

---

## Summary: What You Just Did

✅ Created a complete clinical trial data management system
✅ Set up administrator credentials
✅ Configured security settings
✅ Added team members
✅ Prepared for data collection

**Time invested: 5 minutes**
**System ready for:** Immediate data collection

---

## Common Questions

### Q: Can I change study information later?
**A:** Yes. Use Admin Dashboard → System Configuration to update study details.

### Q: What if my team grows?
**A:** Add new users anytime in User Management. No restart needed.

### Q: How do I backup my data?
**A:** Automatic backups run daily. Manual backups in Admin Dashboard → Backup & Restore.

### Q: Can I change passwords?
**A:** Yes. Users can change their own password. Admins can reset any user's password.

### Q: What if something goes wrong?
**A:** Restore from backup in Admin Dashboard → Backup & Restore. No data is lost.

---

## Ready?

You're all set! Your ZZedc system is ready for research data collection.

[Launch ZZedc Now →](https://github.com/rgt47/zzedc)

Questions? Email rgthomas@ucsd.edu or check the full documentation.

---

**ZZedc Setup Wizard v1.0**
*Making clinical trial data management non-technical*
