---
title: 'Promotion Review Tracker: Technical Setup Guide'
subtitle: 'Installation, configuration, and maintenance for the lab data scientist'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Promotion Review Tracker: Technical Setup Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for the lab data scientist who is
responsible for installing, configuring, and maintaining the
Promotion Review Tracker on behalf of the faculty member. The
system is a ZZedc installation configured to track academic
progress across research, teaching, and service domains for
an upcoming promotion review. The guide covers installation,
configuration, data dictionary authoring, import,
verification, and backup. Moderate R familiarity and
basic command-line comfort are sufficient.

The companion document for the faculty member's assistant,
who handles day-to-day data entry, is
`vignette('promotion-tracker-user-guide')`. After completing
setup, hand the assistant that guide along with the application
URL (or local access instructions) and their login credentials.

The recommended setup path is the initialisation script
(Section 4), which writes all configuration and data-dictionary
files automatically. Sections 5 and 6 describe the manual
alternative for reference and troubleshooting.

---

## 1. Prerequisites

- R 4.4 or later
- RStudio (recommended) or any terminal with R on `PATH`
- The `pak` package: `install.packages('pak')`
- A local working directory for the project, hereafter
  called `~/prj/promotion-tracker/`

No server, cloud account, or Google account is required for
a local SQLite deployment.

---

## 2. Installation

```{r}
pak::pak('zzedc')
```

Verify the installation:

```{r}
library(zzedc)
packageVersion('zzedc')
```

---

## 3. Project Directory

Create the project directory and change into it:

```{r}
dir.create('~/prj/promotion-tracker', recursive = TRUE)
setwd('~/prj/promotion-tracker')
```

ZZedc will write the SQLite database, credentials, and logs
into this directory. Keep it backed up; it is the single
source of truth.

---

## 4. Initialisation Script

The script `inst/scripts/edc-init.R` collects study and user
information, writes `zzedc_config.yml` and `Study_Users.csv`,
and optionally copies a pre-existing `Data_Dictionary.csv` and
`validation_rules.csv` from caller-supplied source paths. It
supports three interface modes suited to different levels of
technical comfort.

### 4.1 Choosing a mode

Run the script with no arguments to be prompted interactively:

```{bash}
Rscript inst/scripts/edc-init.R
```

If `gum` is installed (`brew install gum`), the script presents
a styled menu. Without `gum` it falls back to a plain text
prompt:

```
zzedc Project Initialisation

  1  TUI    -- guided prompts
  2  Config -- plain-text file
  3  Sheets -- Google Sheets

Enter 1, 2, or 3 [1]:
```

Enter a number, or press Enter to accept the default (TUI).
You may also specify the mode directly on the command line to
skip the selection prompt (see §4.2--4.4).

### 4.2 Mode 1: TUI

*Lowest technical barrier. No files to prepare.*

The TUI walks through a sequence of prompts. Each prompt shows
the field name, a brief description, and a default value in
brackets. Press Enter to accept the default or type a new value.

```
── Study information ──────────────────────────────────────
Study name [Promotion Review Tracker]:
Protocol ID [PROMO-2026]:

── Principal investigator ─────────────────────────────────
PI full name: Professor Smith
PI email: psmith@university.edu

── Admin account ──────────────────────────────────────────
Admin username (no spaces) [psmith]:
Admin full name [Professor Smith]:
Admin email [psmith@university.edu]:
Admin password (min 8 chars, 1 upper, 1 digit, 1 symbol):

── Assistant account ──────────────────────────────────────
Assistant username (blank to skip):
...

── Security settings ──────────────────────────────────────
Session timeout in minutes [60]:

── Output location ────────────────────────────────────────
Output directory [.]:
```

The script validates every field before writing any files.
Weak passwords and malformed email addresses are rejected
with a descriptive error.

To invoke TUI mode explicitly:

```{bash}
Rscript inst/scripts/edc-init.R --mode tui
```

### 4.3 Mode 2: Config file

*Mid-level. Edit a plain-text file; no code required.*

On first run without a `--file` argument, the script writes
a commented template called `study.conf` and exits:

```{bash}
Rscript inst/scripts/edc-init.R --mode config
```

Open `study.conf` in any text editor. It looks like this:

```
# zzedc Project Configuration
# Edit the values below. Lines starting with # are comments.

study_name         = Promotion Review Tracker
protocol_id        = PROMO-2026
target_enrollment  = 500
phase              = Observational
db_name            = promotion.db

pi_name            = Professor Smith           # REQUIRED
pi_email           = psmith@university.edu     # REQUIRED

admin_username     = psmith                    # REQUIRED
admin_fullname     = Professor Smith           # REQUIRED
admin_email        = psmith@university.edu     # REQUIRED
admin_password     = ChangeMeNow!2026          # REQUIRED
admin_site_id      = UCSD

coord_1_username   = assistant
coord_1_fullname   = Lab Assistant
coord_1_email      = assistant@university.edu
coord_1_password   = TempPass!2026
coord_1_site_id    = UCSD

session_timeout_minutes = 60
out_dir = .
```

Fill in the values, save the file, then re-run and supply the
data dictionary and validation rules:

```{bash}
Rscript inst/scripts/edc-init.R \
  --mode config --file study.conf \
  --dd-csv vignettes/promotion-tracker/Data_Dictionary.csv \
  --rules-csv vignettes/promotion-tracker/validation_rules.csv
```

The file format is `key = value`. Lines beginning with `#`
and inline comments after `#` are ignored. No YAML syntax,
no quoting, no indentation rules.

### 4.4 Mode 3: Google Sheets

*Highest technical barrier. Requires a Google account and
OAuth authentication in the browser.*

**First run -- create the workbook:**

```{bash}
Rscript inst/scripts/edc-init.R --mode sheets
```

The script opens a browser window for Google authentication,
creates a new Google Sheet in your Drive with a `Study_Info`
tab (study metadata key-value pairs) and a `Study_Users` tab
(one row per user). Both tabs are pre-populated with placeholder
values.

The script prints the new sheet URL and exits. Open the sheet,
edit both tabs with the correct values and any additional
coordinators, then re-run with the URL and supply the
data dictionary and validation rules:

```{bash}
Rscript inst/scripts/edc-init.R \
  --mode sheets --url 'https://docs.google.com/spreadsheets/d/...' \
  --dd-csv vignettes/promotion-tracker/Data_Dictionary.csv \
  --rules-csv vignettes/promotion-tracker/validation_rules.csv
```

**Second run -- read the workbook:**

The script reads `Study_Info` for study metadata and all
Coordinator rows from `Study_Users`, writes `zzedc_config.yml`
and `Study_Users.csv`, copies the supplied CSV files, and
prints the next-steps import commands.

### 4.5 Outputs (all modes)

Regardless of mode, the script produces:

```
<out_dir>/
  zzedc_config.yml      -- ZZedc study configuration
  Study_Users.csv       -- user roster with hashed initial passwords
  Data_Dictionary.csv   -- copied from --dd-csv (if supplied)
  validation_rules.csv  -- copied from --rules-csv (if supplied)
```

Set `out_dir` (TUI prompt or config-file key) to your project
directory (`~/prj/promotion-tracker`) so the files land
alongside the database. The script closes with a printed block
of copy-paste R commands for the import step (see Section 7).

---

## 5. Configuration File (Manual Alternative)

Create `~/prj/promotion-tracker/zzedc_config.yml` with the
following content. Replace the password before first use.

```yaml
study:
  name: "Promotion Review Tracker"
  protocol_id: "PROMO-2026"
  pi_name: "Professor Smith"
  pi_email: "psmith@university.edu"
  target_enrollment: 500
  phase: "Observational"

admin:
  username: "psmith"
  fullname: "Professor Smith"
  email: "psmith@university.edu"
  password: "ChangeMeToSecurePassword123!"

security:
  session_timeout_minutes: 60
  enforce_https: false
  max_failed_login_attempts: 10

compliance:
  gdpr_enabled: false
  cfr_part11_enabled: false
  audit_logging_enabled: true

database:
  type: "sqlite"
  pool_size: 2
```

Notes on these settings:

- `target_enrollment: 500` is a generous ceiling for tracking
  items; it does not constrain the number of records.
- `enforce_https: false` is appropriate for a local-only
  installation. If the application is ever deployed to a
  shared server, set this to `true`.
- GDPR and 21 CFR Part 11 modules are disabled. The audit log
  remains active regardless.

Initialise the database from this file:

```{r}
library(zzedc)
setwd('~/prj/promotion-tracker')
zzedc::init(mode = 'config', config_file = 'zzedc_config.yml')
```

---

## 6. Data Dictionary (Manual Alternative)

The data dictionary defines the seven tracking forms. You may
author it in Google Sheets (preferred) or as a local CSV file.
Both paths are described below.

### 6.1 Google Sheets path

Create a new Google Sheet. Rename the default tab and add
two additional tabs, named exactly as shown:

- `Study_Users`
- `Data_Dictionary`
- `validation_rules`

Capitalisation and spelling are exact; the importer performs
a literal string match.

#### Study_Users tab

| username | full_name | email | role | password_initial | site_id | active |
|---|---|---|---|---|---|---|
| psmith | Professor Smith | psmith@university.edu | Admin | TempPass!2026 | UCSD | TRUE |

Add the faculty member's assistant as `Coordinator` role at
site `UCSD` so they can enter and update records. Add yourself
(the lab data scientist) as `DataManager` or `Admin` to retain
administrative access.

#### Data_Dictionary tab

The table below documents all fields across the seven forms.
The `form_code` column groups fields by form. When authoring
manually in Google Sheets, enter one row per field with the
column headers shown. The init script (Section 4) generates
these files automatically; this table is provided for review
and reference.

**`manuscripts` -- Research publications**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | PUB-YYYY-NNN |
| title | Manuscript Title | text | yes | |
| target_journal | Target Journal | text | | |
| co_authors | Co-authors | text | | |
| status | Status | select | yes | ideation / drafting / submitted / under_review / revision / accepted / in_press / published |
| submission_date | Submission Date | date | | |
| acceptance_date | Acceptance Date | date | | |
| publication_date | Publication Date | date | | |
| doi | DOI | text | | |
| notes | Notes | text | | |

**`courses` -- Teaching activity**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | CRS-YYYY-NNN |
| course_code | Course Code | text | yes | |
| course_title | Course Title | text | yes | |
| academic_year | Academic Year | text | yes | |
| term | Term | radio | yes | Fall / Winter / Spring / Summer |
| role | Role | radio | yes | developed / taught / developed_and_taught |
| student_count | Student Count | numeric | yes | between 1 and 500 |
| contact_hours | Contact Hours | numeric | yes | between 1 and 500 |
| notes | Notes | text | | |

**`mentoring` -- One-on-one tutorials**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | MNT-YYYY-NNN |
| student_initials | Student Initials | text | yes | |
| student_level | Student Level | radio | yes | undergraduate / graduate / postdoc / other |
| start_date | Start Date | date | yes | |
| end_date | End Date | date | | >= start_date |
| hours_per_week | Hours per Week | numeric | | between 0 and 40 |
| topic_area | Topic Area | text | | |
| outcome | Outcome | text | | |
| notes | Notes | text | | |

**`grant_reviews` -- External grant panels**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | GNT-YYYY-NNN |
| agency | Agency | text | yes | |
| program_name | Program Name | text | | |
| panel_name | Panel Name | text | | |
| review_date | Review Date | date | yes | |
| role | Role | radio | yes | reviewer / chair / observer |
| notes | Notes | text | | |

**`committees` -- Committee memberships**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | COM-YYYY-NNN |
| committee_name | Committee Name | text | yes | |
| organization | Organization | text | yes | |
| scope | Scope | radio | | departmental / institutional / national / international |
| role | Role | text | yes | |
| start_date | Start Date | date | yes | |
| end_date | End Date | date | | >= start_date |
| meeting_frequency | Meeting Frequency | radio | | weekly / monthly / quarterly / semi_annual / annual / ad_hoc |
| active | Active | radio | yes | yes / no |
| notes | Notes | text | | |

**`journal_reviews` -- Peer review assignments**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | JRV-YYYY-NNN |
| journal_name | Journal Name | text | yes | |
| manuscript_id | Manuscript ID | text | | |
| review_date | Review Date | date | yes | |
| recommendation | Recommendation | radio | | accept / minor_revision / major_revision / reject / not_reviewed |
| notes | Notes | text | | |

**`dsmb` -- Data safety monitoring boards**

| field_name | Label | Type | Req | Validation |
|---|---|---|:---:|---|
| tracking_id | Tracking ID | text | yes | DSB-YYYY-NNN |
| trial_name | Trial Name | text | yes | |
| sponsor | Sponsor | text | | |
| role | Role | radio | yes | member / chair / statistician / other |
| start_date | Start Date | date | yes | |
| end_date | End Date | date | | >= start_date |
| meeting_frequency | Meeting Frequency | radio | | semi_annual / annual / ad_hoc / other |
| active | Active | radio | yes | yes / no |
| notes | Notes | text | | |

#### validation_rules tab

Eight rules enforce basic data integrity across the forms.

| Rule ID | Form | Field | Condition | Severity |
|---|---|---|---|:---:|
| ACCEPT_AFTER_SUBMIT | manuscripts | acceptance_date | >= submission_date | ERROR |
| PUB_AFTER_ACCEPT | manuscripts | publication_date | >= acceptance_date | ERROR |
| ENROLL_RANGE | courses | student_count | between 1 and 500 | ERROR |
| HOURS_RANGE | courses | contact_hours | between 1 and 500 | ERROR |
| MENTOR_END_AFTER_START | mentoring | end_date | >= start_date | ERROR |
| COMM_END_AFTER_START | committees | end_date | >= start_date | ERROR |
| DSMB_END_AFTER_START | dsmb | end_date | >= start_date | ERROR |
| MENTOR_HOURS | mentoring | hours_per_week | between 0 and 40 | ERROR |

All rules are conditional: they fire only when the relevant
field is non-empty, so optional date and numeric fields do
not block saves on incomplete records.

### 6.2 CSV path (alternative to Google Sheets)

Pre-built template CSV files ship with the package in the
same directory as this guide:

```
vignettes/promotion-tracker/
  Study_Users.csv
  Data_Dictionary.csv
  validation_rules.csv
```

Copy them to your project directory, edit `Study_Users.csv`
to replace the placeholder credentials, and leave the other
two files unchanged unless you need to customise the field
set:

```{r}
pkg_dir <- system.file(
  'vignettes/promotion-tracker',
  package = 'zzedc'
)
file.copy(
  list.files(pkg_dir, pattern = '\\.csv$', full.names = TRUE),
  '~/prj/promotion-tracker/'
)
```

The CSV import path in ZZedc reads these files directly
once they are in the project directory.

---

## 7. Importing the Data Dictionary

### From Google Sheets

Share the workbook with Editor permission and copy the URL.
Then run:

```{r}
library(zzedc)
setwd('~/prj/promotion-tracker')

setup_zzedc_from_gsheets(
  sheet_url = 'https://docs.google.com/spreadsheets/d/YOUR_SHEET_ID',
  db_path   = 'data/promotion.db'
)
```

Expected output: one confirmation line per tab successfully
read, then a summary of forms, fields, validation rules, and
users imported. Elapsed time is typically under five seconds
for this data dictionary.

### From CSV

The three CSV files must be in your project directory
(`~/prj/promotion-tracker/`) before running this command.
They are placed there either by the init script (Section 4)
or by copying the package templates (Section 6.2).

```{r}
library(zzedc)
setwd('~/prj/promotion-tracker')

setup_zzedc_from_csv(
  users_csv      = 'Study_Users.csv',
  dictionary_csv = 'Data_Dictionary.csv',
  rules_csv      = 'validation_rules.csv',
  db_path        = 'data/promotion.db'
)
```

### Checking the import log

The import prints any skipped rules or unrecognised field
types. Review this output before proceeding. Common issues:

| Message | Cause | Fix |
|---|---|---|
| `Skipped rule: <rule_id>` | DSL syntax error | Re-check the `rule_dsl` cell against the patterns in the Content-Author Guide |
| `Unknown field type: <type>` | Typo in `field_type` | Allowed types: `text`, `numeric`, `date`, `select`, `radio`, `checkbox` |
| `Tab not found: Study_Users` | Tab name mismatch | Rename to exactly `Study_Users` |
| `No rows in Data_Dictionary` | Empty row before end of data | Delete the empty row; re-import |

---

## 8. Verification

### Launch and inspect

```{r}
launch_zzedc()
```

Log in with the `psmith` admin account and the initial
password set in `zzedc_config.yml`. You will be forced to
change it on first login.

Navigate to **Admin > Forms** and confirm that all seven
forms appear: `manuscripts`, `courses`, `mentoring`,
`grant_reviews`, `committees`, `journal_reviews`, `dsmb`.

Click into each form and verify:

- Field labels match the `field_label` column in the
  data dictionary.
- Required fields show a red asterisk.
- Select and radio fields show the correct options.

### Dry run

Create one synthetic record in each form using the
appropriate tracking ID prefix. Attempt to save a record
with a deliberately wrong date (e.g., an acceptance date
before the submission date) and confirm that the validation
error fires with the expected message.

Delete all synthetic records after the dry run. Deletions
are recorded in the audit log; the records themselves are
removed from the data tables.

### Audit log integrity check

```{r}
library(zzedc)
setwd('~/prj/promotion-tracker')

verify_audit_log_integrity(db_path = 'data/promotion.db')
```

Expected output: `Audit chain intact. N entries verified.`
Run this check after the dry run and again after any bulk
data migration.

---

## 9. Backup

For a local SQLite installation, backup is straightforward.

### Manual backup

```{r}
file.copy(
  'data/promotion.db',
  paste0('backups/promotion_', format(Sys.time(), '%Y%m%d'), '.db')
)
```

Create a `backups/` subdirectory in the project folder and
add it to your cloud sync (Dropbox, iCloud, etc.) if the
project directory is not already synced.

### Automated daily backup via cron (macOS)

Add a line to your crontab (`crontab -e`):

```
0 22 * * * Rscript -e "file.copy('~/prj/promotion-tracker/data/promotion.db', paste0('~/prj/promotion-tracker/backups/promotion_', format(Sys.time(), '%Y%m%d'), '.db'))"
```

This runs at 22:00 each night. Adjust the path and time
as needed.

### Restore from backup

```{r}
file.copy(
  'backups/promotion_20260515.db',
  'data/promotion.db',
  overwrite = TRUE
)
```

Stop the Shiny session before restoring, then restart
`launch_zzedc()` afterward.

---

## 10. Updating the Data Dictionary After Go-Live

**Adding or changing validation rules:** edit the
`validation_rules` tab and re-run the import command. The
import is safe to re-run; it merges new rules and updates
changed rules without deleting existing subject data.

**Adding a new field to an existing form:** add the row to
`Data_Dictionary` and re-run the import. New fields appear
as optional by default; existing records will show them as
blank until updated.

**Changing a field type** after data have been entered is not
supported by the importer. Contact the ZZedc maintainer for
a migration path.

**Adding a new form:** add the rows to `Data_Dictionary` and
re-run the import. The new form tab appears in the
application on the next launch.

---

## 11. Directory Structure (Final State)

After setup, the project directory should contain:

```
~/prj/promotion-tracker/
  zzedc_config.yml        # Study configuration
  Study_Users.csv         # (if using CSV path)
  Data_Dictionary.csv     # (if using CSV path)
  validation_rules.csv    # (if using CSV path)
  data/
    promotion.db          # Live SQLite database
  backups/
    promotion_YYYYMMDD.db # Rolling daily backups
  logs/                   # Application logs (if enabled)
```

Keep `zzedc_config.yml` in version control with the password
field replaced by a placeholder. Do not commit the `.db`
file to a public repository; it contains your academic
activity records.

---

## 12. Troubleshooting

**Import fails with `googlesheets4` authentication error.**
Run `googlesheets4::gs4_auth()` interactively to authorise
access to your Google account, then re-run the import.

**Port 3838 already in use.** Pass an alternate port:
`launch_zzedc(port = 3939)`. The user will access the
application at `http://localhost:3939`.

**The database is locked.** A previous R session did not
close cleanly and left a write lock. Restart R, then check
whether any other R process is running against the same
database file before relaunching.

**A validation rule fires on every save, even correct
data.** The most common cause is a typo in `field_code`
that causes the rule to reference a field that does not
exist. Re-check the `field_code` column against the
`field_name` values in `Data_Dictionary`.

**User cannot log in.** Confirm the `active` column is
`TRUE` for that user in `Study_Users`. If the account is
locked after repeated failed attempts, reset it in the
application under **Admin > Users**.

---

## 13. Reference

For deeper detail on any of the following, consult the
corresponding vignette:

| Topic | Vignette |
|---|---|
| Google Sheets authoring in full | `vignette('content-author-guide')` |
| Server deployment, TLS, AWS | `vignette('technical-lead-guide')` |
| Database backend selection | `vignette('backend-quickstart')` |
| Setup wizard (GUI alternative to config YAML) | `vignette('setup-wizard-guide')` |
| Regulatory compliance modules | `vignette('zzedc-whitepaper')` |
