---
title: 'Promotion Review Tracker: User Guide'
subtitle: 'Day-to-day data entry reference for the lab assistant'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Promotion Review Tracker: User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for the faculty member's administrative
assistant, who is responsible for day-to-day data entry and
record maintenance in the Promotion Review Tracker. The tracker
is a ZZedc application configured to monitor the faculty
member's academic progress across research, teaching, and
service domains in preparation for an upcoming promotion review.
No knowledge of R, databases, or ZZedc internals is required. A
lab data scientist has installed and configured the system; this
guide covers everything the assistant will do day to day.

---

## 1. Launching the Application

Open a terminal or RStudio console and run:

```{r}
library(zzedc)
launch_zzedc()
```

The application opens in your default browser at
`http://localhost:3838`. Keep the console open while you work;
closing it stops the server.

If you prefer a persistent local installation that starts
automatically, ask the lab data scientist to configure
auto-launch on login.

---

## 2. Logging In

Enter your username and password on the login screen. On your
first login you will be prompted to change the initial password.
Choose a password you will remember; it is not recoverable
without the lab data scientist's help.

If the application shows a session-timeout warning, save your
current entry before clicking away. Unsaved form fields are lost
on timeout.

---

## 3. Application Layout

The left sidebar contains seven tabs, one per tracking domain:

| Tab | What it tracks |
|---|---|
| Publications | Manuscripts from idea to print |
| Courses | Courses developed or taught |
| Mentoring | One-on-one student tutorials |
| Grant Reviews | External grant panels served |
| Committees | Departmental and national committees |
| Journal Reviews | Peer review assignments completed |
| DSMB | Data safety monitoring board memberships |

The top navigation bar also provides access to the Data Explorer
(for browsing all your records at once), the Reports tab, and the
Export Center.

---

## 4. Entering Data

### How to create a new entry

1. Click the relevant tab in the left sidebar.
2. Click the **New Record** button (top right of the form panel).
3. Assign a tracking ID following the convention for that domain
   (see each section below).
4. Fill in the fields. Required fields are marked with a red
   asterisk and will block the save if left empty.
5. Click **Save**. A green confirmation banner appears if the
   save succeeded. A red banner describes any validation errors.

### How to edit an existing entry

Click the record in the table at the top of the tab, then edit
the fields that have changed and click **Save**.

---

### 4.1 Publications (`PUB-YYYY-NNN`)

Use this form to track every manuscript from initial concept
through publication. Create one record per manuscript.

**Tracking ID convention:** `PUB-2026-001`, `PUB-2026-002`, ...
Reset the counter each calendar year.

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier, format `PUB-YYYY-NNN` |
| Title | Working or final manuscript title |
| Target Journal | Journal of intended submission |
| Co-authors | Comma-separated list of co-author surnames |
| Status | Current stage (see status list below) |
| Submission Date | Date the manuscript was first submitted |
| Acceptance Date | Date of formal acceptance notice |
| Publication Date | Date of online or print publication |
| DOI | Digital object identifier once assigned |
| Notes | Any notes relevant to the promotion dossier |

**Status options** (choose one):

- `ideation` -- concept stage, not yet drafted
- `drafting` -- active writing in progress
- `submitted` -- submitted to a journal, awaiting desk decision
- `under_review` -- sent to external reviewers
- `revision` -- revision requested; manuscript in progress
- `accepted` -- accepted; in production queue
- `in_press` -- accepted and assigned a DOI; not yet in an issue
- `published` -- assigned to a volume and issue

**Validation:** the system will flag you if the acceptance date
precedes the submission date, or if the publication date precedes
the acceptance date.

**Tip:** enter a record at ideation stage and update the status
field as the manuscript progresses. The Data Explorer will show
you your full pipeline at a glance.

---

### 4.2 Courses (`CRS-YYYY-NNN`)

Use this form to record every course you developed or taught.
One record per course per term.

**Tracking ID convention:** `CRS-2026-001`, `CRS-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Course Code | Department code and number, e.g., `FMPH 101` |
| Course Title | Full course title |
| Academic Year | e.g., `2025-2026` |
| Term | Fall, Winter, Spring, or Summer |
| Role | `developed`, `taught`, or `developed_and_taught` |
| Student Count | Total enrolled students |
| Contact Hours | Total instructor contact hours for the term |
| Notes | Curriculum innovations, guest lectures, etc. |

**Contact hours** are the total hours you spent in scheduled
instructional contact with the class, not preparation time.
A three-credit lecture course meeting three times per week for
ten weeks is 30 contact hours.

---

### 4.3 Mentoring (`MNT-YYYY-NNN`)

Use this form to record individual tutorial relationships.
One record per student.

**Tracking ID convention:** `MNT-2026-001`, `MNT-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Student Initials | Initials only (e.g., `J.D.`) |
| Student Level | undergraduate, graduate, postdoc, or other |
| Start Date | Date the relationship began |
| End Date | Date the relationship ended (leave blank if ongoing) |
| Hours per Week | Approximate weekly tutorial hours |
| Topic Area | Research area or skill being mentored |
| Outcome | Degree earned, publication, employment, etc. |
| Notes | Any notes relevant to the dossier |

**Validation:** the end date, if entered, must fall on or after
the start date.

---

### 4.4 Grant Reviews (`GNT-YYYY-NNN`)

Use this form to record each external grant review service.
One record per review assignment.

**Tracking ID convention:** `GNT-2026-001`, `GNT-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Agency | Funding agency, e.g., `NIH`, `NSF`, `PCORI` |
| Program Name | Study section or program name |
| Panel Name | Panel or review group identifier |
| Review Date | Date of the review panel meeting or deadline |
| Role | `reviewer`, `chair`, or `observer` |
| Notes | Notes relevant to the dossier |

---

### 4.5 Committees (`COM-YYYY-NNN`)

Use this form to record standing or ad-hoc committee
memberships. One record per committee.

**Tracking ID convention:** `COM-2026-001`, `COM-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Committee Name | Official name of the committee |
| Organization | Institution or body, e.g., `UCSD`, `ASA`, `NIH` |
| Scope | departmental, institutional, national, or international |
| Role | Your role, e.g., `member`, `chair`, `secretary` |
| Start Date | Date your membership began |
| End Date | Date your membership ended (leave blank if current) |
| Meeting Frequency | Typical meeting cadence |
| Active | `yes` if membership is current, `no` if completed |
| Notes | Notes relevant to the dossier |

**Validation:** the end date, if entered, must fall on or after
the start date.

---

### 4.6 Journal Reviews (`JRV-YYYY-NNN`)

Use this form to record each peer review assignment.
One record per manuscript reviewed.

**Tracking ID convention:** `JRV-2026-001`, `JRV-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Journal Name | Name of the journal |
| Manuscript ID | Blinded identifier supplied by the journal |
| Review Date | Date you submitted your review |
| Recommendation | Your editorial recommendation |
| Notes | Notes relevant to the dossier (do not include manuscript details) |

**Recommendation options:** `accept`, `minor_revision`,
`major_revision`, `reject`, `not_reviewed`.

Use `not_reviewed` if you declined after seeing the abstract.
This maintains a complete record of all review requests.

---

### 4.7 DSMB Memberships (`DSB-YYYY-NNN`)

Use this form to record data safety monitoring board
memberships. One record per trial.

**Tracking ID convention:** `DSB-2026-001`, `DSB-2026-002`, ...

| Field | What to enter |
|---|---|
| Tracking ID | Unique identifier |
| Trial Name | Abbreviated or full trial name |
| Sponsor | Funding agency or industry sponsor |
| Role | `member`, `chair`, `statistician`, or `other` |
| Start Date | Date your membership began |
| End Date | Date your membership ended (leave blank if current) |
| Meeting Frequency | Typical meeting cadence |
| Active | `yes` if membership is current, `no` if completed |
| Notes | Notes relevant to the dossier |

---

## 5. Reviewing Your Data

The **Data Explorer** tab (top navigation bar) provides a
consolidated view of all records across all seven forms.

Useful features:

- **Filter by date range** to isolate the promotion review
  period (typically the last three to five years; confirm the
  exact window with your department).
- **Filter by domain** using the form-code selector.
- **Sort any column** by clicking the column header.
- **Search** for a keyword using the search box to locate a
  specific manuscript, committee, or journal.

For the publications pipeline, use the status filter to count
how many manuscripts are in each stage.

---

## 6. Reports

The **Reports** tab provides summary counts and timelines for
each domain. The following reports are most relevant to the
promotion review:

| Report | Use |
|---|---|
| Activity Summary | Counts per domain per academic year |
| Publications Pipeline | Status breakdown across manuscripts |
| Teaching Load | Student counts and contact hours by year |
| Service Profile | Committee and review counts by year |

Run the Activity Summary report at the start of each semester
and save a copy as confirmation of your progress.

---

## 7. Exporting Data for the Promotion Dossier

When you are ready to compile the formal dossier:

1. Open the **Export Center** tab.
2. Select the date range that matches your promotion review
   window.
3. Choose your preferred format: Excel for a working
   spreadsheet, CSV for import into other tools.
4. Click **Export**. The file downloads to your browser's
   default download folder.

Export each domain separately for the cleanest output, then
assemble the individual sheets into the dossier tables your
department requires.

The audit log records every export. If the promotion committee
asks for confirmation that the data were not altered after
export, your technical lead can provide a signed audit extract.

---

## 8. Recommended Data Entry Cadence

Keeping the tracker current takes roughly five to ten minutes per
entry if done at the time the activity occurs. Batch
reconstruction is error-prone.

| Activity | When to enter or update |
|---|---|
| New manuscript | At conception (status: `ideation`) |
| Manuscript submission | Same day (update status, add submission date) |
| Manuscript accepted | Same day (update status, add acceptance date) |
| Manuscript published | Same day (update status, add DOI, publication date) |
| Course taught | First week of each term |
| Mentoring relationship | At start and end |
| Grant review completed | Same day as review deadline or panel meeting |
| Committee meeting attended | End of each semester |
| Journal review submitted | Same day |
| DSMB meeting attended | Day of the meeting |

A monthly five-minute review of open records (manuscripts in
`under_review` or `revision`, committees with no end date,
mentoring with no outcome) is sufficient to catch anything
missed.

---

## 9. Troubleshooting

**The application will not launch.** Close the browser and
re-run `launch_zzedc()` in a fresh R session. If the port
3838 is occupied, your lab data scientist can configure an
alternate port.

**A save fails with a validation error.** Read the red banner
message; it identifies the field and what is wrong. The most
common causes are a date out of order (acceptance before
submission) or a required field left empty.

**A record is missing.** Use the Data Explorer search box.
Records are not deleted from the database; if it was saved, it
is there.

**You forgot your password.** Contact the lab data scientist.
Passwords are hashed and cannot be recovered, but an
administrator can reset the account.

---

## 10. Glossary

- **Tracking ID**: the unique identifier you assign to each
  entry. Follow the domain prefix conventions in Section 4.
- **Status**: the current lifecycle stage of a manuscript.
- **Contact hours**: total scheduled instructor-student contact
  time for a course, not including preparation.
- **Audit log**: a system-maintained record of every change to
  the database. You do not need to manage it; it runs
  automatically.
- **Export Center**: the application tab for downloading your
  data as Excel or CSV files.
