---
title: 'Prazosin Sleep Study (Small): Technical Setup Guide'
subtitle: 'Installation, configuration, and maintenance for the lab data scientist'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prazosin Sleep Study (Small): Technical Setup Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*2026-05-26 16:46 PDT*

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for the lab data scientist responsible for
installing, configuring, and maintaining the ZZedc instance for
the Prazosin Sleep Study (Small), protocol PRAZ-S-2026. The
study is a single-site, single-instance deployment: one SQLite
database at UCSD, one application process, two user accounts.
Setup is straightforward. The guide covers prerequisites,
installation, project directory structure, the initialisation
script, the manual CSV alternative, import, verification, backup,
and troubleshooting.

The companion document for the site coordinator, who handles
day-to-day data entry, is
`vignette('prazosin-small-user-guide')`. After completing setup,
hand the coordinator that guide, the application URL (or local
access instructions), and their initial credentials.

---

## 1. Prerequisites

- R 4.4 or later
- RStudio (recommended) or any terminal with R on `PATH`
- The `pak` package: `install.packages('pak')`
- A local project directory, hereafter
  `~/prj/praz-s-2026/`

No server, cloud account, or Google account is required for
a local SQLite deployment.

---

## 2. Installation

```{r}
pak::pak('zzedc')
```

Verify the installation:

```{r}
library(zzedc)
packageVersion('zzedc')
```

---

## 3. Project Directory

Create the project directory:

```{r}
dir.create('~/prj/praz-s-2026', recursive = TRUE)
setwd('~/prj/praz-s-2026')
```

ZZedc will write the SQLite database, credentials, and log files
into this directory. It is the single source of truth for the
trial; keep it backed up (see Section 9).

---

## 4. Initialisation Script

The package ships with a generalised init script at
`inst/scripts/edc-init.R`. It collects study and user
information, writes `zzedc_config.yml` and `Study_Users.csv`,
and optionally copies a pre-existing `Data_Dictionary.csv` and
`validation_rules.csv` from caller-supplied source paths.

The recommended path for a single-site clinical trial is
Mode 2 (config file), described in Section 4.1 below, because
it produces a plain-text audit trail of the configuration
decisions.

### 4.1 Mode 2: Config file (recommended)

On first run without a `--file` argument, the script writes a
commented template called `study.conf` and exits:

```{bash}
Rscript inst/scripts/edc-init.R --mode config
```

Open `study.conf` and edit the study-specific values:

```
study_name         = Prazosin Sleep Study (Small)
protocol_id        = PRAZ-S-2026
target_enrollment  = 50
phase              = Phase II
db_name            = praz-s-2026.db

pi_name            = Professor Smith
pi_email           = psmith@ucsd.edu

admin_username     = psmith                    # REQUIRED
admin_fullname     = Professor Smith           # REQUIRED
admin_email        = psmith@ucsd.edu           # REQUIRED
admin_password     = ChangeMeNow!2026          # REQUIRED
admin_site_id      = UCSD

coord_1_username   = assistant
coord_1_fullname   = Site Coordinator
coord_1_email      = assistant@ucsd.edu
coord_1_password   = TempPass!2026
coord_1_site_id    = UCSD

session_timeout_minutes = 60
out_dir = ~/prj/praz-s-2026
```

Save the file, then re-run with the `--file` flag and supply
the pre-existing data dictionary and validation rules:

```{bash}
Rscript inst/scripts/edc-init.R \
  --mode config --file study.conf \
  --dd-csv vignettes/prazosin-small/Data_Dictionary.csv \
  --rules-csv vignettes/prazosin-small/validation_rules.csv
```

The script validates every field before writing any files.
Weak passwords and malformed email addresses are rejected
with a descriptive error.

### 4.2 Script outputs

Regardless of mode, the script produces:

```
~/prj/praz-s-2026/
  zzedc_config.yml      -- ZZedc study configuration
  Study_Users.csv       -- user roster with hashed initial passwords
  Data_Dictionary.csv   -- field definitions across 7 forms
  validation_rules.csv  -- cross-field and range validation rules
```

The script closes with a printed block of copy-paste R commands
for the import step (see Section 7).

---

## 5. Configuration File (Manual Alternative)

If you prefer to skip the init script, create
`~/prj/praz-s-2026/zzedc_config.yml` directly. Replace the
password before first use.

```yaml
study:
  name: "Prazosin Sleep Study (Small)"
  protocol_id: "PRAZ-S-2026"
  pi_name: "Professor Smith"
  pi_email: "psmith@ucsd.edu"
  target_enrollment: 50
  phase: "Phase II"

admin:
  username: "psmith"
  fullname: "Professor Smith"
  email: "psmith@ucsd.edu"
  password: "ChangeMeToSecurePassword123!"

security:
  session_timeout_minutes: 60
  enforce_https: false
  max_failed_login_attempts: 10

compliance:
  gdpr_enabled: false
  cfr_part11_enabled: false
  audit_logging_enabled: true

database:
  type: "sqlite"
  pool_size: 2
```

Notes on these settings:

- `target_enrollment: 50` matches the protocol-specified
  enrollment ceiling. This value informs summary reports but
  does not prevent additional records from being created.
- `enforce_https: false` is appropriate for a local-only
  installation at a single workstation. If the application is
  ever accessed over a network, set this to `true` and consult
  the ZZedc technical-lead guide for TLS configuration.
- GDPR and 21 CFR Part 11 modules are disabled. The audit log
  remains active regardless of these settings.
- `pool_size: 2` is sufficient for a two-user installation with
  no concurrent heavy queries.

Initialise the database from this file:

```{r}
library(zzedc)
setwd('~/prj/praz-s-2026')
zzedc::init(mode = 'config', config_file = 'zzedc_config.yml')
```

---

## 6. CSV Configuration Files (Manual Alternative)

The three CSV files define users, the data dictionary, and
validation rules. Pre-built templates for this trial ship with
the package in the same directory as this guide:

```
vignettes/prazosin-small/
  Study_Users.csv
  Data_Dictionary.csv
  validation_rules.csv
```

Copy them to your project directory:

```{r}
pkg_dir <- system.file(
  'vignettes/prazosin-small',
  package = 'zzedc'
)
file.copy(
  list.files(pkg_dir, pattern = '\\.csv$', full.names = TRUE),
  '~/prj/praz-s-2026/'
)
```

**Study_Users.csv** contains placeholder credentials for
`psmith` and `assistant`. Edit the `password_initial` column
before importing; do not leave the default `TempPass!2026` in
production unless you instruct both users to change it on first
login (the application enforces a password-change prompt on first
login when using initial credentials).

**Data_Dictionary.csv** defines all fields for the seven forms:
`enrollment`, `randomization`, `study_drug`, `sleep_psqi`,
`adverse_events`, `concomitant_meds`, and `visits`. The column
structure is:

| Column | Description |
|---|---|
| `form_code` | Internal form identifier; must match exactly in all files |
| `field_name` | Internal field identifier; used in validation rule expressions |
| `field_label` | Human-readable label shown in the application UI |
| `field_type` | One of `text`, `numeric`, `date`, `select`, `radio`, `checkbox` |
| `required` | `TRUE` if the field blocks save when empty; `FALSE` otherwise |
| `validation` | Field-level constraint expression (e.g., `between 0 and 21`) |
| `description` | Tooltip text shown in the application |
| `field_order` | Numeric sort order within the form |

**validation_rules.csv** defines 12 cross-field and range rules
covering PSQI component and global score ranges, date ordering
across forms, and compliance percentage bounds. The column
structure is:

| Column | Description |
|---|---|
| `rule_id` | Unique rule identifier; uppercase, no spaces |
| `form_code` | Form to which the rule applies |
| `rule_type` | `FIELD` for single-field range rules; `CROSS_FIELD` for multi-field rules |
| `expression` | Rule DSL expression evaluated at save time |
| `error_message` | Message shown to the user when the rule fires |
| `is_active` | `TRUE` to enable; `FALSE` to disable without deleting |

Do not modify `Data_Dictionary.csv` or `validation_rules.csv`
unless you need to customise the field set or rules. Any
structural change (adding or removing fields, altering field
names) must be made before the first import; changes after
go-live require a migration path.

---

## 7. Importing with `setup_zzedc_from_csv()`

With the three CSV files in the project directory, run:

```{r}
library(zzedc)
setwd('~/prj/praz-s-2026')

setup_zzedc_from_csv(
  users_csv      = 'Study_Users.csv',
  dictionary_csv = 'Data_Dictionary.csv',
  rules_csv      = 'validation_rules.csv',
  db_path        = 'data/praz-s-2026.db'
)
```

Expected output: one confirmation line per tab successfully
read, then a summary of forms, fields, validation rules, and
users imported. Elapsed time is typically under five seconds.

### Checking the import log

Review the printed import log before proceeding. Common issues
and their resolution:

| Message | Cause | Fix |
|---|---|---|
| `Skipped rule: <rule_id>` | DSL syntax error in `expression` | Re-check against Content-Author Guide patterns |
| `Unknown field type: <type>` | Typo in `field_type` column | Allowed: `text`, `numeric`, `date`, `select`, `radio`, `checkbox` |
| `No rows in Data_Dictionary` | Empty row before end of file | Delete trailing empty row; re-import |
| `Duplicate field_name in form` | Same `field_name` appears twice in one form | Remove or rename the duplicate row |

---

## 8. Verification

### Launch and inspect

```{r}
launch_zzedc()
```

Log in with the `psmith` admin account and the initial password
set in `zzedc_config.yml`. You will be prompted to change the
password on first login.

Navigate to **Admin > Forms** and confirm that all seven forms
appear: `enrollment`, `randomization`, `study_drug`,
`sleep_psqi`, `adverse_events`, `concomitant_meds`, `visits`.

For each form, verify:

- Field labels match the `field_label` column in
  `Data_Dictionary.csv`.
- Required fields display a red asterisk.
- Radio and select fields display the correct option lists.

### Dry run

Create one synthetic record in each form. Use subject ID
`PRAZ-S-000` to distinguish test data from real records.
Attempt at least the following validation checks:

- Enter a randomization date before the consent date; confirm
  the rule `RAND_AFTER_CONSENT` fires.
- Enter a PSQI component score of 4; confirm the relevant
  component range rule fires.
- Enter a PSQI global score of 22; confirm `PSQI_GLOBAL_RANGE`
  fires.
- Enter an AE resolution date before the onset date; confirm
  `AE_RESOLUTION_AFTER_ONSET` fires.
- Enter a concomitant medication stop date before the start
  date; confirm `MED_STOP_AFTER_START` fires.
- Enter a compliance percentage of 110; confirm
  `COMPLIANCE_RANGE` fires.

Delete all synthetic records after the dry run. Deletions are
recorded in the audit log; the records themselves are removed
from the data tables.

### Audit log integrity check

```{r}
library(zzedc)
setwd('~/prj/praz-s-2026')

verify_audit_log_integrity(db_path = 'data/praz-s-2026.db')
```

Expected output: `Audit chain intact. N entries verified.`
Run this after the dry run and again after any bulk data
migration.

---

## 9. Backup

For a local SQLite installation, backup is a file copy.

### Manual backup

```{r}
dir.create('~/prj/praz-s-2026/backups', showWarnings = FALSE)
file.copy(
  'data/praz-s-2026.db',
  paste0(
    'backups/praz-s-2026_',
    format(Sys.time(), '%Y%m%d'),
    '.db'
  )
)
```

### Automated daily backup via cron (macOS)

Add a line to your crontab (`crontab -e`):

```
0 22 * * * Rscript -e "file.copy('~/prj/praz-s-2026/data/praz-s-2026.db', paste0('~/prj/praz-s-2026/backups/praz-s-2026_', format(Sys.time(), '%Y%m%d'), '.db'))"
```

This runs at 22:00 nightly. If the project directory is inside
a Dropbox or cloud-synced folder, the backup files are
additionally covered by the cloud provider's version history.

### Restore from backup

```{r}
file.copy(
  'backups/praz-s-2026_20260601.db',
  'data/praz-s-2026.db',
  overwrite = TRUE
)
```

Stop the Shiny session before restoring, then restart
`launch_zzedc()` afterward.

---

## 10. Directory Structure (Final State)

After setup, the project directory should contain:

```
~/prj/praz-s-2026/
  zzedc_config.yml          # Study configuration
  study.conf              # Init-script config (keep for audit)
  Study_Users.csv           # User roster (if using CSV path)
  Data_Dictionary.csv       # Field definitions (if using CSV path)
  validation_rules.csv      # Validation rules (if using CSV path)
  data/
    praz-s-2026.db          # Live SQLite database
  backups/
    praz-s-2026_YYYYMMDD.db # Rolling daily backups
  logs/                     # Application logs (if enabled)
```

Keep `zzedc_config.yml` and `study.conf` in version control
with the password fields replaced by placeholders. Do not commit
the `.db` file to any repository; it contains subject data.

---

## 11. Updating the Configuration After Go-Live

**Adding or modifying a validation rule:** edit
`validation_rules.csv` and re-run `setup_zzedc_from_csv()`.
The import merges new and updated rules without deleting
subject data.

**Adding a new optional field:** add the row to
`Data_Dictionary.csv` and re-run the import. The new field
appears in the form as optional; existing records show it
as blank.

**Changing a field type** after data have been entered is not
supported by the importer. Contact the ZZedc maintainer for
a migration path if this is necessary.

**Adding a new form:** add all rows for the new form to
`Data_Dictionary.csv` and re-run the import. The new form tab
appears on the next application launch.

---

## 12. Troubleshooting

**Port 3838 already in use.** Pass an alternate port:
`launch_zzedc(port = 3939)`. Update the coordinator's access
instructions accordingly.

**The database is locked.** A previous R session did not close
cleanly. Restart R, verify that no other R process is running
against the same file, then relaunch.

**Import fails with duplicate key error.** A re-run of the
import is usually safe, but if the `rule_id` or `field_name`
values were edited between runs, conflicts can arise. Drop the
database and re-import from the CSV files: rename the existing
`.db` file to `.db.bak`, then re-run `setup_zzedc_from_csv()`.

**A validation rule fires on every save, even correct data.**
The most common cause is a `field_name` mismatch between
`Data_Dictionary.csv` and the `expression` column in
`validation_rules.csv`. Confirm the names match exactly,
including case.

**User cannot log in.** Check that the `active` column is
`TRUE` for that user in `Study_Users.csv` and that the value
was imported correctly via **Admin > Users** in the application.
An administrator can reset the account password from that panel.

**Session timeouts are too aggressive.** Increase
`session_timeout_minutes` in `zzedc_config.yml` and restart
the application. The default is 60 minutes; 120 is reasonable
for a coordinator who enters data during a clinic session.

---

## 13. Reference

For deeper detail on any of the following, consult the
corresponding vignette:

| Topic | Vignette |
|---|---|
| Server deployment, TLS, AWS | `vignette('technical-lead-guide')` |
| Database backend selection | `vignette('backend-quickstart')` |
| Regulatory compliance modules | `vignette('zzedc-whitepaper')` |
| Full content-author DSL reference | `vignette('content-author-guide')` |
| Setup wizard (GUI alternative to config YAML) | `vignette('setup-wizard-guide')` |
