---
title: 'Prazosin Sleep Study (Small): User Guide'
subtitle: 'Day-to-day data entry reference for the site coordinator'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prazosin Sleep Study (Small): User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*2026-05-26 16:46 PDT*

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for the site coordinator at UCSD, who is
responsible for day-to-day data entry in the Prazosin Sleep Study
(Small), protocol PRAZ-S-2026. The study is a randomised
double-blind placebo-controlled parallel-group trial comparing
prazosin to placebo for sleep quality in up to 50 subjects. The
ZZedc application captures all study data across seven forms
corresponding to the study schedule and data collection
instruments. No knowledge of R, databases, or ZZedc internals is
required to use the system. This guide covers everything you will
do day to day.

---

## 1. Launching the Application

Open a terminal or RStudio console and run:

```{r}
library(zzedc)
launch_zzedc()
```

The application opens in your default browser at
`http://localhost:3838`. Keep the console open while you work;
closing it stops the server.

If the port is already in use (error message: 'port 3838 is in
use'), contact the lab data scientist to configure an alternate
port or to identify the conflicting process.

---

## 2. Logging In

Enter your username (`assistant`) and your current password on
the login screen. On first login you will be prompted to set a
new password. Choose a password you will remember; it cannot be
recovered without the lab data scientist's assistance.

If the application shows a session-timeout warning, save your
current entry before clicking away. Unsaved field values are
lost on timeout.

---

## 3. Application Layout

The left sidebar contains seven tabs, one per data collection
form:

| Tab | What it captures |
|---|---|
| Enrollment | Screening, eligibility, and consent |
| Randomization | Treatment allocation and randomization code |
| Study Drug | Dose dispensing, returns, and compliance |
| Sleep PSQI | Pittsburgh Sleep Quality Index assessments |
| Adverse Events | Adverse event onset, characterisation, and outcome |
| Concomitant Meds | Concurrent medications during study participation |
| Visits | Visit schedule, completion, and study status |

The top navigation bar also provides access to the Data Explorer
(for browsing all records at once), the Reports tab, and the
Export Center.

---

## 4. Tracking ID Conventions

Consistent tracking IDs are essential for linking records across
forms and for generating clean data exports. Apply the following
conventions without exception.

**Subjects:** `PRAZ-S-NNN`

Assign the next available three-digit number at each new
enrolment. The counter resets at the start of each calendar year.
Examples: `PRAZ-S-001`, `PRAZ-S-002`, ... `PRAZ-S-050`.

**Adverse event records:** `AE-NNN`

Each AE record receives its own sequential identifier,
independent of subject ID. The counter does not reset by year
within a study. Examples: `AE-001`, `AE-002`, `AE-003`.

**Concomitant medication records:** `MED-NNN`

Each medication record receives its own sequential identifier.
Examples: `MED-001`, `MED-002`, `MED-003`.

---

## 5. Entering Data

### 5.1 How to create a new record

1. Click the relevant tab in the left sidebar.
2. Click the **New Record** button (top right of the form panel).
3. Enter the subject ID or record ID following the conventions
   in Section 4.
4. Fill in the fields. Required fields are marked with a red
   asterisk and will block the save if left empty.
5. Click **Save**. A green confirmation banner appears if the
   save succeeded. A red banner identifies any validation errors
   and the field that caused them.

### 5.2 How to edit an existing record

Click the record in the table at the top of the tab to load it
into the edit panel. Change the relevant fields and click
**Save**. Every save is recorded in the audit log with your
username and a timestamp.

---

### 5.3 Enrollment Form (`PRAZ-S-NNN`)

Create one enrollment record per screened subject, including
subjects who do not ultimately enrol.

| Field | What to enter |
|---|---|
| Subject ID | `PRAZ-S-NNN` per the convention in Section 4 |
| Site ID | `UCSD` (pre-filled; do not change) |
| Consent Date | Date the subject signed the informed consent document |
| Date of Birth | Subject date of birth |
| Sex | Select `male`, `female`, or `other_or_prefer_not` |
| Diagnosis | Primary sleep complaint diagnosis as stated in the medical record |
| IC1: Age 18 or Older | `yes` if subject is 18 years of age or older |
| IC2: Sleep Complaint | `yes` if subject reports a clinically significant sleep complaint |
| IC3: Stable Medications | `yes` if all medications have been stable for at least 30 days |
| EC1: Alpha-Blocker Use | `yes` if subject is currently taking any alpha-blocker medication |
| EC2: Pregnancy | `yes` if subject is currently pregnant or intends to become pregnant |
| Eligibility Determination | `yes` if all ICs are met and no ECs are triggered; `no` if the subject fails on clinical grounds but is not a formal screen failure; `screen_failure` if ICs or ECs result in formal screen failure |
| Screen Failure Reason | Required if Eligibility is `screen_failure`; describe the specific criterion not met |
| Notes | Any additional clinical notes |

**Tip:** create the enrollment record at first contact, before
consent, if you need to track pre-consent screening. Update the
consent date field once consent is obtained.

---

### 5.4 Randomization Form

Create one randomization record per subject after eligibility is
confirmed and the randomization code is retrieved from the
pharmacy.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` from the enrollment record |
| Site ID | `UCSD` |
| Randomization Date | Date the subject was randomized |
| Treatment Arm | `prazosin` or `placebo` as assigned by the code |
| Randomization Code | The sealed code provided by the pharmacy or IWRS |
| Randomized By | Your username or the username of the staff member who completed randomization |
| Notes | Any notes relevant to the randomization event |

**Validation:** the randomization date must fall on or after the
consent date. The system will flag a mismatch if these dates are
out of order.

Do not enter a randomization record for subjects who screen fail
before reaching the eligibility determination step.

---

### 5.5 Study Drug Form

Create one study drug record per visit at which drug is
dispensed or a compliance review is conducted. A subject will
typically have records at baseline, week 4, week 8, and week 12.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` |
| Visit Date | Date of this dispensing or review visit |
| Dose (mg) | Prescribed dose in milligrams: `1`, `2`, `4`, or `6` |
| Doses Dispensed | Number of doses dispensed to the subject at this visit |
| Doses Returned | Number of doses returned by the subject at this visit |
| Compliance Percentage | Calculated compliance: ((dispensed - returned) / expected) x 100; enter as a whole number between 0 and 100 |
| Dose Change | `yes` if the dose was adjusted at this visit; `no` otherwise |
| Dose Change Reason | Required if Dose Change is `yes`; describe the clinical reason |
| Notes | Any additional notes |

**Validation:** the compliance percentage must be between 0 and
100. The system will flag values outside this range.

---

### 5.6 Sleep PSQI Form

Create one PSQI record per visit at which the Pittsburgh Sleep
Quality Index is administered. For this protocol the PSQI is
collected at baseline, week 4, week 8, week 12, and end of study.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` |
| Visit Date | Date the PSQI was administered |
| Visit Number | Study visit number: 1 through 6 |
| PSQI 1 through 7 | Each component score: 0 (best) through 3 (worst) |
| PSQI Global Score | Sum of the seven component scores: 0 through 21 |
| Notes | Any clinically relevant notes |

**Scoring reminder:** the PSQI global score is the arithmetic
sum of the seven component scores. A global score above 5
indicates poor sleep quality. Verify that the global score you
enter matches the sum of the seven components before saving.

**Validation:** each component score must be between 0 and 3;
the global score must be between 0 and 21. The system enforces
these constraints.

---

### 5.7 Adverse Events Form (`AE-NNN`)

Create one record per adverse event. If the same event recurs
after full resolution, create a second record.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` |
| AE Record ID | `AE-NNN` per the convention in Section 4 |
| Onset Date | Date the adverse event began |
| Resolution Date | Date the adverse event resolved; leave blank if ongoing |
| AE Term | Standardised adverse event term (use MedDRA preferred term if available) |
| Severity | `mild`, `moderate`, `severe`, or `life_threatening` |
| Serious AE | `yes` if the event meets protocol-defined SAE criteria; `no` otherwise |
| Relatedness | `yes`, `no`, `possibly`, or `probably` per investigator assessment |
| Action Taken | `none`, `dose_reduced`, `dose_interrupted`, `drug_discontinued`, or `other` |
| Outcome | `resolved`, `ongoing`, `resolved_with_sequelae`, or `fatal` |
| Notes | Additional clinical detail |

**Validation:** the resolution date, if entered, must fall on or
after the onset date.

**SAE reporting:** if Serious AE is `yes`, immediately notify
Professor Smith and complete the institutional SAE reporting form
per the study protocol. The ZZedc record documents the event but
does not substitute for regulatory SAE notification.

---

### 5.8 Concomitant Meds Form (`MED-NNN`)

Create one record per concomitant medication per subject. Update
the stop date when a medication is discontinued.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` |
| Medication Record ID | `MED-NNN` per the convention in Section 4 |
| Medication Name | Generic or brand name |
| Indication | Clinical reason the subject is taking this medication |
| Start Date | Date the medication was initiated |
| Stop Date | Date the medication was discontinued; leave blank if ongoing |
| Dose | Dose and unit, e.g., `10 mg` |
| Route | `oral`, `topical`, `iv`, `im`, or `other` |
| Ongoing | `yes` if the medication is currently being taken; `no` if discontinued |
| Notes | Additional notes |

**Validation:** the stop date, if entered, must fall on or after
the start date.

Collect a complete concomitant medication list at screening and
update it at each visit. Pay particular attention to any
alpha-blocker medications, as these represent an exclusion
criterion and should have been captured at enrollment.

---

### 5.9 Visits Form

Create one visit record per scheduled study contact. Update the
record at the time of the visit or within 24 hours.

| Field | What to enter |
|---|---|
| Subject ID | Matching `PRAZ-S-NNN` |
| Visit Number | 1 through 6 per the study schedule |
| Visit Date | Date the visit occurred or was attempted |
| Visit Type | `screening`, `baseline`, `week4`, `week8`, `week12`, `end_of_study`, or `early_termination` |
| Completed | `yes` if the visit was completed as planned; `no` if not |
| Reason Not Completed | Required if Completed is `no`; describe why |
| Study Status | `ongoing`, `completed`, `withdrawn`, or `lost_to_followup` |
| Withdrawal Reason | Required if Study Status is `withdrawn`; describe the reason |
| Notes | Additional notes |

**Study schedule summary:**

| Visit Number | Visit Type | Approximate Timing |
|---|---|---|
| 1 | screening | Day -14 to Day -1 |
| 2 | baseline | Day 0 (randomization) |
| 3 | week4 | Week 4 |
| 4 | week8 | Week 8 |
| 5 | week12 | Week 12 |
| 6 | end_of_study | Week 12 or at early termination |

---

## 6. Recommended Data Entry Cadence

Entering data at the time of the activity prevents recall errors
and reduces the burden of batch reconstruction.

| Activity | When to enter or update |
|---|---|
| Subject screening contact | Day of screening, before consent if applicable |
| Informed consent obtained | Same day; update consent date on enrollment form |
| Eligibility determination | Same day as eligibility review |
| Randomization | Same day; enter code immediately after pharmacy contact |
| Drug dispensed | Same day as dispensing visit |
| PSQI administered | Same day as visit |
| AE identified | Same day; update outcome and resolution when known |
| Concomitant med change | Same day as change is identified or reported |
| Visit completed or missed | Within 24 hours of the visit date |
| Subject withdrawal | Same day as withdrawal discussion |

A weekly five-minute review of open records (AEs without a
resolution date, ongoing concomitant meds, visits with
incomplete status) is sufficient to keep the database current.

---

## 7. Reviewing Data

The **Data Explorer** tab in the top navigation bar provides a
consolidated view across all seven forms.

Useful features:

- **Filter by subject ID** to review all data for a single
  subject across forms.
- **Filter by form** using the form-code selector.
- **Sort any column** by clicking the column header.
- **Search** by keyword to locate a specific adverse event term
  or medication name.

Use the Data Explorer at each visit to review a subject's prior
records before entering new data. This reduces duplicate entries
and ensures dates are consistent.

---

## 8. Reports

The **Reports** tab provides summary counts and timelines. The
following reports are most relevant to study operations:

| Report | Use |
|---|---|
| Enrollment Summary | Screened vs enrolled vs screen-failed counts |
| Compliance Summary | Mean compliance by subject and visit |
| AE Summary | AE counts by severity and relatedness |
| Visit Completion | Visit completion rates by visit number |

Run the Enrollment Summary weekly and share with Professor Smith
during study team meetings.

---

## 9. Exporting Data

When exporting data for a data safety monitoring review, protocol
deviation report, or sponsor query:

1. Open the **Export Center** tab.
2. Select the form or forms to export.
3. Choose your preferred format: Excel for working review,
   CSV for import into statistical software.
4. Click **Export**. The file downloads to your browser's
   default download folder.

The audit log records every export with your username and
timestamp. Do not share raw exports outside the study team
without PI approval.

---

## 10. Troubleshooting

**The application will not launch.** Close the browser, stop any
existing R session, and re-run `launch_zzedc()` in a fresh R
session. If port 3838 is occupied, contact the lab data
scientist.

**A save fails with a validation error.** Read the red banner
message; it identifies the field and the rule that was violated.
The most common causes are a date out of order (randomization
before consent, AE resolution before onset) or a required field
left empty.

**A record appears to be missing.** Use the Data Explorer
search box. Records are not deleted from the database; if the
record was saved, it is present.

**The PSQI global score does not match the sum of components.**
The system stores the value you entered, not a calculated
total. If you detect a discrepancy during review, edit the
record to correct either the components or the global score,
whichever is erroneous, and document the correction in the
notes field.

**You forgot your password.** Contact the lab data scientist.
Passwords are hashed and cannot be recovered, but an
administrator can reset the account credentials.

**An SAE needs to be reported.** Complete the ZZedc record as
described in Section 5.7. Then follow the institutional SAE
reporting procedure and notify Professor Smith immediately.
The ZZedc system is not a regulatory reporting tool.

---

## 11. Glossary

- **Subject ID**: the unique identifier assigned at first
  contact with format `PRAZ-S-NNN`.
- **AE Record ID**: the unique identifier for each adverse
  event record, format `AE-NNN`.
- **Medication Record ID**: the unique identifier for each
  concomitant medication record, format `MED-NNN`.
- **Inclusion criterion (IC)**: a condition the subject must
  meet to be eligible for enrolment.
- **Exclusion criterion (EC)**: a condition that disqualifies a
  subject from enrolment.
- **Screen failure**: formal determination that a subject does
  not meet eligibility criteria after screening procedures.
- **PSQI**: Pittsburgh Sleep Quality Index; a validated
  19-item self-report questionnaire assessing sleep quality
  over the past month. Global scores range from 0 to 21;
  scores above 5 indicate clinically poor sleep.
- **SAE**: serious adverse event; an AE meeting one or more
  regulatory criteria for seriousness (hospitalisation, life
  threat, disability, death, etc.).
- **Audit log**: a system-maintained record of every change to
  the database, including user, timestamp, and previous value.
  You do not need to manage it; it runs automatically.
- **Compliance percentage**: calculated as
  ((doses dispensed - doses returned) / doses expected) x 100.
