---
title: 'Quick Start: Database Backends'
subtitle: 'A unified setup guide for the five supported ZZedc backends'
author: 'ZZedc Team'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start: Database Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  eval = FALSE
)
```

## Scope

ZZedc runs against five database backends through a single adapter
interface: SQLite, DuckDB, MySQL or MariaDB, PostgreSQL, and ClickHouse.
The application layer is backend-agnostic. Once a connection is
configured, study setup, form authoring, data entry, and reporting are
identical across backends. This vignette consolidates what was
previously five parallel quick-start guides into one document.

The body covers the workflow that is the same for every backend: pick
a backend, install ZZedc, write a `config.yml`, initialise the
database, and launch the application. Backend-specific mechanics
(server install, driver selection, dialect features, backup tooling)
are isolated in five short appendices. Read the body first, then read
only the appendix that matches your chosen backend.

---

## 1. Picking a backend

### Decision tree

The backend choice is a function of study size, deployment context,
and whether analytics or transactions dominate the workload.

- **SQLite** for prototypes, single-investigator studies, pilots, and
  any deployment with fewer than approximately 200 subjects and one
  to three concurrent users. It is the default and requires nothing
  beyond an R installation.
- **PostgreSQL** for multi-site production deployments, regulated
  studies (21 CFR Part 11, GDPR Article 32), and any context where
  the institution already operates a PostgreSQL server. It offers
  true multi-user concurrency, streaming replication, and
  point-in-time recovery.
- **MySQL or MariaDB** for the same multi-site profile when the
  institution standardises on MySQL or MariaDB rather than
  PostgreSQL, or when an existing REDCap deployment is the source of
  data. The two server families share a wire protocol and a single R
  driver (`RMariaDB`).
- **DuckDB** for analytics-heavy studies on data already in hand:
  interim analyses, secondary analyses of existing trial data, and
  any workflow where Parquet export to Python or SAS is on the
  critical path. DuckDB is file-based like SQLite but uses columnar
  storage, so aggregation queries are roughly an order of magnitude
  faster.
- **ClickHouse** for very-high-volume telemetry-style data: hundreds
  of millions of rows of sensor or device output, real-time
  dashboards, and time-series analytics. ClickHouse is not
  recommended for typical CRF data entry, where transactional
  guarantees matter more than aggregation speed.

A more detailed comparison, including a backend cheat sheet keyed to
specific study scenarios, lives in Chapter 2 of the
[Technical Lead Guide](technical-lead-guide.html). Refer there before
committing to a non-default backend on a production study.

### Comparison at a glance

The following table summarises the dimensions that drive the choice.
It is intentionally coarse; treat it as orientation, not as a scoring
rubric.

| Dimension          | SQLite    | DuckDB    | MySQL/MariaDB | PostgreSQL | ClickHouse |
|--------------------|-----------|-----------|---------------|------------|------------|
| Architecture       | File      | File      | Server        | Server     | Server     |
| Server install     | None      | None      | Required      | Required   | Required   |
| Concurrent writers | 1-3       | 1-3       | Many          | Many       | Many       |
| Best workload      | OLTP      | OLAP      | OLTP          | OLTP       | OLAP       |
| Best dataset size  | <1M rows  | <100M     | <1B rows      | <1B rows   | Unlimited  |
| Parquet export     | Via R pkg | Native    | Via R pkg     | Via R pkg  | Native     |
| Replication / HA   | No        | No        | Yes           | Yes        | Yes        |
| Compliance posture | Basic     | Basic     | Full          | Full       | Full       |
| Setup complexity   | None      | Minimal   | Moderate      | Moderate   | High       |

`OLTP` denotes online transaction processing, the typical CRF
data-entry pattern. `OLAP` denotes online analytical processing, the
aggregation-heavy pattern statisticians use during analysis.

---

## 2. Prerequisites

The following prerequisites are common to every backend. Backend-specific
server prerequisites appear in the corresponding appendix.

- R version 4.1 or higher (the codebase uses the native pipe `|>`).
- An R environment with package installation available (RStudio is
  the supported IDE but not required).
- Internet access for the initial `devtools::install_github()` call.
- Write access to a working directory for the study artefacts (the
  `config.yml`, the `logs/` and `exports/` subdirectories, and, for
  the file-based backends, the database file itself).

Verify the R version:

```{r check_r}
R.version.string
# Should print 'R version 4.1.0' or higher.
```

---

## 3. Install ZZedc

### From GitHub

```{r install_github}
if (!requireNamespace('devtools', quietly = TRUE)) {
  install.packages('devtools')
}
devtools::install_github('rgt47/zzedc')
```

### From a local clone

```{r install_local}
# Clone first:
#   git clone https://github.com/rgt47/zzedc.git
#   cd zzedc
devtools::install('.')
```

### Verify

```{r verify}
library(zzedc)
packageVersion('zzedc')
```

The R driver for the chosen backend (`RSQLite`, `duckdb`, `RMariaDB`,
`RPostgres`, or `RClickhouse`) is installed separately; see the
relevant appendix.

---

## 4. Create the project layout

Every study, regardless of backend, uses the same on-disk layout:

```
my_study/
  config.yml          # study and database configuration
  data/               # SQLite or DuckDB file lives here
  logs/               # application logs
  exports/            # Parquet or CSV exports
  backups/            # database backups
```

```{r project_layout}
project_dir <- 'my_study'
dir.create(project_dir, showWarnings = FALSE)
for (sub in c('data', 'logs', 'exports', 'backups')) {
  dir.create(file.path(project_dir, sub), showWarnings = FALSE)
}
setwd(project_dir)
```

For the client-server backends the `data/` directory is unused; the
database lives on the server. The other three directories are still
useful for application logs, exports, and locally cached backups.

---

## 5. Write `config.yml`

ZZedc reads its database configuration from a `config.yml` in the
working directory. The shape is the same for every backend; only the
`database` block differs. Below is the SQLite default. Substitute the
appropriate block from the relevant appendix when targeting a different
backend.

```yaml
default:
  study_name:  'My Clinical Trial'
  protocol_id: 'STUDY-2026-001'

  database:
    backend: sqlite
    sqlite:
      path: 'data/my_study.db'

  session_timeout_minutes: 30
  max_login_attempts: 3

  app_title:  'My Clinical Trial EDC'
  debug_mode: false
```

Two conventions are worth flagging:

- The `${ZZEDC_DB_PASSWORD}` syntax interpolates from the process
  environment. Use it for any client-server backend so the password is
  not committed alongside the configuration. Set the variable in your
  shell or in a `.env` file the runtime sources.
- A `production:` section can inherit from `default:` and override
  individual keys (e.g., a shorter `session_timeout_minutes`). This is
  the recommended pattern when the same `config.yml` is used in both
  development and production.

---

## 6. Initialise the database

The `create_db_adapter()` factory dispatches on `db_backend` and
returns an R6 adapter object whose `connect()`, `query()`, `execute()`,
and `disconnect()` methods are identical across backends. Study setup
runs through `create_wizard_database()`, which creates the canonical
ZZedc tables (`edc_users`, `subjects`, `data_entries`, `audit_trail`,
and the supporting tables) and seeds the admin account.

```{r init_db}
library(zzedc)

config <- list(
  db_backend = 'sqlite',
  sqlite = list(path = 'data/my_study.db'),

  study_name        = 'My Clinical Trial',
  protocol_id       = 'STUDY-2026-001',
  pi_name           = 'Dr. Investigator',
  pi_email          = 'investigator@example.org',
  study_phase       = 'Phase 2',
  target_enrollment = 50,

  admin_username = 'admin',
  admin_password = 'ChangeMe!2026',
  admin_fullname = 'Study Administrator',
  admin_email    = 'admin@example.org',

  security_salt = paste0(
    sample(c(letters, LETTERS, 0:9), 32, replace = TRUE),
    collapse = ''
  ),
  session_timeout    = 30,
  max_login_attempts = 3
)

adapter <- create_db_adapter(config)
conn    <- adapter$connect()
on.exit(adapter$disconnect(conn), add = TRUE)

result <- create_wizard_database(
  config,
  config$sqlite$path,
  backend = config$db_backend
)

if (!isTRUE(result$success)) {
  stop('Database initialisation failed: ', result$message)
}
```

The `default_db_config()` helper reduces the boilerplate when the
study-level fields are inherited from elsewhere:

```{r default_config}
config <- default_db_config(
  'postgresql',
  host     = 'db.example.org',
  name     = 'my_clinical_trial',
  user     = 'zzedc_user',
  password = Sys.getenv('ZZEDC_DB_PASSWORD')
)
adapter <- create_db_adapter(config)
```

For non-default backends, replace the `db_backend` and the
corresponding nested list with the values shown in the appendix.

### Setup wizard alternative

The interactive wizard accepts the same configuration and is suitable
for non-programmatic study setup:

```{r wizard}
library(zzedc)
launch_setup_wizard()
```

The wizard prompts for study information, the admin account, security
parameters, the database backend, and the initial data dictionary.

---

## 7. Launch the application

Once the database is initialised, launching the Shiny application is
backend-independent:

```{r launch}
library(zzedc)

Sys.setenv(ZZEDC_DB_BACKEND = 'sqlite')
Sys.setenv(ZZEDC_DB_PATH    = 'data/my_study.db')

launch_zzedc(port = 3838)
```

For client-server backends, set the connection variables instead:

```{r launch_server}
Sys.setenv(ZZEDC_DB_BACKEND = 'postgresql')
Sys.setenv(ZZEDC_DB_HOST    = 'localhost')
Sys.setenv(ZZEDC_DB_PORT    = '5432')
Sys.setenv(ZZEDC_DB_NAME    = 'my_clinical_trial')
Sys.setenv(ZZEDC_DB_USER    = 'zzedc_user')
Sys.setenv(ZZEDC_DB_PASSWORD = '...')

launch_zzedc(port = 3838)
```

The browser opens to `http://localhost:3838`. Log in with the admin
credentials seeded during setup, then proceed to user management, form
design, and subject enrollment.

---

## 8. Migrating between backends

A study that begins on SQLite during the pilot can be promoted to
PostgreSQL or MySQL when it goes multi-site. The
`migrate_between_backends()` helper copies all canonical tables
between any pair of supported backends and translates the dialect
differences (auto-increment columns, `JSON` versus `JSONB`, `DateTime`
versus `TIMESTAMP`, and so on).

```{r migrate}
library(zzedc)

source_config <- default_db_config(
  'sqlite',
  path = 'data/my_study.db'
)
dest_config <- default_db_config(
  'postgresql',
  host     = 'db.example.org',
  name     = 'my_clinical_trial',
  user     = 'zzedc_user',
  password = Sys.getenv('ZZEDC_DB_PASSWORD')
)

result <- migrate_between_backends(source_config, dest_config)
result$tables  # per-table row counts and timing
```

The reverse direction (e.g., PostgreSQL to SQLite for an offline
archive) works by swapping the two configuration objects. Migration
preserves the audit-trail hash chain; the integrity check in
`verify_audit_chain()` should pass on the destination database after
migration.

---

## 9. Backup, recovery, and maintenance

A consistent backup posture has two pieces: a regular full backup and,
for client-server backends, a write-ahead-log archive that supports
point-in-time recovery. The mechanics are backend-specific and live in
the appendix. The principles below apply universally:

- Schedule full backups daily, retained for at least thirty days.
- Test the restore path quarterly. An untested backup is a hypothesis,
  not a guarantee.
- Store backups off the database host. A backup co-located with the
  database it is intended to protect does not protect against host
  loss.
- For regulated studies, encrypt backups at rest and document the key
  custody chain.

The application-level audit trail in `audit_trail` is a separate
artefact from the database backup and complies with 21 CFR Part 11
ALCOA-C requirements. It is preserved by every backup mechanism the
appendices describe, but its hash-chained integrity should be
verified after any restore.

---

## 10. Troubleshooting (common patterns)

The five backends share a small set of failure modes. Backend-specific
errors appear in the appendix. The patterns below are universal.

- **'Could not connect'.** For file-based backends, check the path
  resolves and the directory is writable. For client-server
  backends, check the server is reachable
  (`pg_isready`, `mysqladmin ping`, `clickhouse-client --query
  'SELECT 1'`), the port is open, and the user is permitted from the
  client IP.
- **'Authentication failed'.** Check the credential value (a stale
  `ZZEDC_DB_PASSWORD` is the most common cause), then check the
  authentication method on the server side. PostgreSQL's
  `pg_hba.conf` and MySQL's `'user'@'host'` grant pattern are both
  strict.
- **'Driver not found'.** Install the driver listed in the
  appendix (`install.packages('RPostgres')`, etc.). The error message
  usually names the missing package.
- **'Session expired'.** The default timeout is 30 minutes. Adjust
  `session_timeout_minutes` in `config.yml` if the workflow requires
  a longer window.
- **'Database is locked' (SQLite only).** Close the other R session
  or application that holds the file. SQLite serialises writers.

---

## 11. Summary

The unified workflow is: pick a backend (Section 1), install ZZedc
(Section 3), assemble the project layout and `config.yml` (Sections
4-5), initialise the database (Section 6), launch (Section 7). Sections
8-10 cover the operational lifecycle. The five appendices document the
backend-specific mechanics: server install, R driver, dialect features,
and backup tooling.

For a complete worked example end-to-end, see the
[Demonstration Trial Setup](demonstration-trial-setup.html) vignette,
which walks through study setup with synthetic data on the SQLite
backend.

---

## Appendix A. SQLite

### Server install

None. SQLite ships with the `RSQLite` package, which is a
recommended dependency of ZZedc. No server process runs. The database
is a single file (`data/my_study.db` by default).

### R driver

```{r sqlite_driver}
install.packages('RSQLite')
```

### `config.yml` block

```yaml
database:
  backend: sqlite
  sqlite:
    path: 'data/my_study.db'
```

### Backup and maintenance

A SQLite backup is a file copy of a database that is not currently
being written. Use `sqlite3 ... .backup` for online backups, or stop
the application briefly for a `file.copy()`.

```{r sqlite_backup}
backup_path <- sprintf(
  'backups/my_study_%s.db',
  format(Sys.time(), '%Y%m%d_%H%M%S')
)
file.copy('data/my_study.db', backup_path)
```

Periodic `VACUUM` reclaims space from deleted rows; `ANALYZE` updates
the query planner's statistics. Both are cheap to run during a
maintenance window:

```{r sqlite_vacuum}
library(DBI)
library(RSQLite)
conn <- dbConnect(SQLite(), 'data/my_study.db')
dbExecute(conn, 'VACUUM')
dbExecute(conn, 'ANALYZE')
dbDisconnect(conn)
```

### Backend-specific notes

- SQLite serialises writers. A second R session attempting to write
  will block until the first releases its lock. This is the expected
  cost of zero-configuration concurrency.
- For studies that outgrow SQLite, the most common promotion path is
  to PostgreSQL via `migrate_between_backends()`.
- Encryption at rest is supported via the encrypted-database helpers
  documented in
  [Technical-Lead Guide Chapter 5](technical-lead-guide.html).
  The other backends configure encryption at the server level instead.

---

## Appendix B. DuckDB

### Server install

None. DuckDB is in-process like SQLite, but uses columnar storage.

### R driver

```{r duckdb_driver}
install.packages('duckdb')
library(duckdb)
packageVersion('duckdb')
```

### `config.yml` block

```yaml
database:
  backend: duckdb
  duckdb:
    path: 'data/my_study.duckdb'
```

### Native Parquet export

DuckDB's defining feature for clinical trial work is direct Parquet
export. Statisticians can read the resulting files in Python, R, or
SAS without an intermediate CSV step:

```{r duckdb_parquet}
library(duckdb)
conn <- dbConnect(duckdb(), 'data/my_study.duckdb')

# Export a single table.
dbExecute(conn, "
  COPY subjects
  TO 'exports/subjects.parquet'
  (FORMAT PARQUET)
")

# Export a join result with ZSTD compression.
dbExecute(conn, "
  COPY (
    SELECT s.*, r.treatment_arm
    FROM subjects s
    LEFT JOIN randomization r ON s.subject_id = r.subject_id
  ) TO 'exports/subjects_with_treatment.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD)
")

dbDisconnect(conn)
```

For a full archive, iterate over `SHOW TABLES` and emit one Parquet
file per table.

### Backup and maintenance

DuckDB databases are single files. A backup is a file copy taken with
the application stopped or with all writers disconnected. The Parquet
export above is also a portable archive format and is recommended in
addition to file-level backup for long-term retention.

### Backend-specific notes

- Memory usage is higher than SQLite for the same dataset because
  DuckDB streams columns into memory for analytical queries. Cap it
  with `config = list(memory_limit = '2GB')` on the connection if the
  default exceeds available RAM.
- Indexes are less critical than on row-oriented databases because of
  columnar layout, but `CREATE INDEX` is supported for selective
  filters.
- Concurrent writers are limited (similar to SQLite). DuckDB is best
  treated as a single-writer analytical store.

---

## Appendix C. MySQL or MariaDB

### Server versions supported

| Server             | Versions                       | Notes                |
|--------------------|--------------------------------|----------------------|
| MariaDB 11.4 LTS   | June 2024 to June 2029         | Headline target      |
| MariaDB 10.11 LTS  | February 2023 to February 2028 | Documented floor     |
| MySQL 8.4 LTS      | April 2024 to April 2032       | Headline MySQL target|

End-of-life server versions (MySQL 8.0 EOL April 2026, MariaDB 10.5
EOL June 2025) are not supported. The adapter requires features
present from MySQL 8.0 / MariaDB 10.5 onward.

### Server install (Docker)

```bash
docker run -d --name zzedc_mariadb \
  -e MARIADB_ROOT_PASSWORD='RootPwd!' \
  -e MARIADB_DATABASE=zzedc_study \
  -e MARIADB_USER=zzedc_user \
  -e MARIADB_PASSWORD='StudyPassword!' \
  -p 3306:3306 \
  mariadb:11.4
```

For a managed institutional MySQL or MariaDB server, the DBA performs
the equivalent setup; no Docker is involved.

### Initial server setup (SQL)

```sql
CREATE DATABASE zzedc_study
  CHARACTER SET = 'utf8mb4'
  COLLATE       = 'utf8mb4_unicode_ci';

CREATE USER 'zzedc_user'@'%'
  IDENTIFIED BY 'your-strong-password';

GRANT SELECT, INSERT, UPDATE, DELETE,
      CREATE, DROP, ALTER, INDEX, REFERENCES
  ON zzedc_study.*
  TO 'zzedc_user'@'%';

FLUSH PRIVILEGES;
```

The `utf8mb4` character set is required (not `utf8`) so that subject
names and free-text fields handle the full Unicode range.

### R driver

```{r mysql_driver}
install.packages('RMariaDB')
# Optional, for multi-user Shiny deployments:
install.packages('pool')
```

`RMariaDB` is the maintained DBI driver for both MySQL and MariaDB
servers. The older `RMySQL` package is in maintenance mode and is not
recommended.

### `config.yml` block

```yaml
database:
  backend: mysql
  mysql:
    host:     'db.example.org'
    port:     3306
    name:     'zzedc_study'
    user:     'zzedc_user'
    password: '${ZZEDC_DB_PASSWORD}'
    ssl_mode: 'REQUIRED'
    pool_min: 1
    pool_max: 5
```

`backend: mariadb` is accepted as an alias and produces an identical
adapter; the wire protocol is shared.

### TLS

For any deployment crossing a network boundary, configure
`ssl-cert`, `ssl-key`, and `ssl-ca` on the server, then set
`ssl_mode: REQUIRED` (or `VERIFY_IDENTITY` for full peer-certificate
validation) in `config.yml`. The adapter exposes
`adapter$is_ssl_connected(conn)` to verify the negotiated state.

### Backup

```bash
mysqldump --single-transaction --routines --triggers \
  -h db.example.org -u zzedc_user -p \
  zzedc_study | gzip > backups/zzedc_$(date +%Y%m%d).sql.gz
```

For continuous protection, configure binary-log replication to a
standby server. The binary log is also the basis for point-in-time
recovery (`mysqlbinlog --start-datetime=...`).

### REDCap migration

REDCap stores its data in MySQL. A common reason to choose this
backend is the prospect of ingesting a REDCap export. The current
supported path is to export REDCap data as CSV plus the metadata
dictionary and author the corresponding ZZedc CRFs from there. A
direct REDCap-to-ZZedc importer is on the roadmap; see
[the roadmap vignette](mysql-redcap-migration-roadmap.html).

### Backend-specific notes

- Encryption at rest is configured at the server level (InnoDB
  tablespace encryption), not through the adapter. The `key` argument
  used by the SQLite encrypted-database helpers is ignored on this
  backend.
- The most common error is the host pattern in the `CREATE USER`
  grant: `'zzedc_user'@'%'` allows any host;
  `'zzedc_user'@'10.0.0.0/8'` restricts to a subnet. Mismatches
  produce 'Access denied for user'.

---

## Appendix D. PostgreSQL

### Server install (Docker)

```bash
cat > docker-compose.yml << 'EOF'
version: '3.8'
services:
  postgres:
    image: postgres:16
    container_name: zzedc_postgres
    environment:
      POSTGRES_USER:     zzedc_admin
      POSTGRES_PASSWORD: SecurePassword123!
      POSTGRES_DB:       zzedc_trial
    ports:
      - '5432:5432'
    volumes:
      - pgdata:/var/lib/postgresql/data
    restart: unless-stopped
volumes:
  pgdata:
EOF

docker-compose up -d
```

### Server install (Homebrew, macOS)

```bash
brew install postgresql@16
brew services start postgresql@16
psql --version
```

### Server install (Ubuntu / Debian)

```bash
sudo apt update
sudo apt install postgresql-16
sudo systemctl enable --now postgresql
psql --version
```

### Initial server setup (SQL)

Connect as the superuser (`psql -U postgres`) and run:

```sql
CREATE USER zzedc_user WITH PASSWORD 'StudyPassword123!';

CREATE DATABASE my_clinical_trial
  OWNER      = zzedc_user
  ENCODING   = 'UTF8'
  LC_COLLATE = 'en_US.UTF-8'
  LC_CTYPE   = 'en_US.UTF-8';

GRANT ALL PRIVILEGES ON DATABASE my_clinical_trial TO zzedc_user;

\c my_clinical_trial
GRANT ALL ON SCHEMA public TO zzedc_user;
```

### R driver

```{r postgres_driver}
install.packages('RPostgres')
library(RPostgres)
packageVersion('RPostgres')
```

### `config.yml` block

```yaml
database:
  backend: postgresql
  postgresql:
    host:     'localhost'
    port:     5432
    name:     'my_clinical_trial'
    user:     'zzedc_user'
    password: '${ZZEDC_DB_PASSWORD}'
    sslmode:  'verify-full'
```

### TLS and `pg_hba.conf`

For production, enable TLS in `postgresql.conf`:

```
ssl              = on
ssl_cert_file    = '/path/to/server.crt'
ssl_key_file     = '/path/to/server.key'
ssl_ca_file      = '/path/to/ca.crt'
```

Restrict authentication in `pg_hba.conf` so non-TLS connections are
rejected:

```
# TYPE      DATABASE         USER         ADDRESS         METHOD
local       all              postgres                     peer
hostssl     my_clinical_trial zzedc_user  10.0.0.0/8      scram-sha-256
hostssl     my_clinical_trial zzedc_user  192.168.0.0/16  scram-sha-256
hostnossl   all              all          0.0.0.0/0       reject
```

### Backup and recovery

For full backups, schedule `pg_dump` via cron:

```bash
#!/bin/bash
BACKUP_DIR='/var/backups/postgresql'
DB_NAME='my_clinical_trial'
TIMESTAMP=$(date +%Y%m%d_%H%M%S)
mkdir -p "$BACKUP_DIR"
pg_dump -h localhost -U zzedc_user "$DB_NAME" \
  | gzip > "${BACKUP_DIR}/${DB_NAME}_${TIMESTAMP}.sql.gz"
find "$BACKUP_DIR" -name '*.sql.gz' -mtime +30 -delete
```

For point-in-time recovery, enable WAL archiving:

```
wal_level       = replica
archive_mode    = on
archive_command = 'cp %p /var/lib/postgresql/wal_archive/%f'
```

For high availability, set up streaming replication. On the primary:

```
wal_level       = replica
max_wal_senders = 3
wal_keep_size   = 1GB
```

```sql
CREATE ROLE replicator WITH REPLICATION LOGIN
  PASSWORD 'ReplicaPass123!';
```

On the replica, stop PostgreSQL, clear the data directory, and run
`pg_basebackup -h primary -U replicator -D /var/lib/postgresql/16/main
-Fp -Xs -P -R`.

### Connection pooling

For multi-user Shiny deployments, place PgBouncer in front of
PostgreSQL. The `pool` R package provides session-level pooling on
top, which is complementary.

### Backend-specific notes

- PostgreSQL is the recommended backend for any deployment that needs
  to satisfy 21 CFR Part 11 audit and access-control requirements.
- The maintenance commands worth scheduling are `ANALYZE` (cheap, run
  weekly) and `VACUUM FULL ANALYZE` (requires an exclusive lock, run
  during a maintenance window only when bloat is significant).

---

## Appendix E. ClickHouse

### When ClickHouse is and is not appropriate

ClickHouse is a column-oriented analytical database. It excels at
aggregating hundreds of millions to billions of rows with sub-second
query times. It is **not** designed for transactional workloads.
Typical CRF data entry is transactional, so for a routine clinical
trial PostgreSQL or SQLite is usually the better choice. ClickHouse is
appropriate when the study generates very-high-volume telemetry
(continuous device output, sensor streams, real-time monitoring) and
real-time dashboards are a primary deliverable.

### Server install (Docker)

```bash
cat > docker-compose.yml << 'EOF'
version: '3.8'
services:
  clickhouse:
    image: clickhouse/clickhouse-server:latest
    container_name: zzedc_clickhouse
    ports:
      - '8123:8123'   # HTTP
      - '9000:9000'   # Native TCP
    environment:
      CLICKHOUSE_USER:     zzedc_user
      CLICKHOUSE_PASSWORD: SecurePassword123!
      CLICKHOUSE_DEFAULT_ACCESS_MANAGEMENT: 1
    volumes:
      - clickhouse_data:/var/lib/clickhouse
      - clickhouse_logs:/var/log/clickhouse-server
    ulimits:
      nofile: { soft: 262144, hard: 262144 }
    restart: unless-stopped
volumes:
  clickhouse_data:
  clickhouse_logs:
EOF

docker-compose up -d
```

### Server install (Ubuntu / Debian)

```bash
sudo apt-get install -y apt-transport-https ca-certificates dirmngr
sudo apt-key adv --keyserver hkp://keyserver.ubuntu.com:80 \
  --recv 8919F6BD2B48D754
echo 'deb https://packages.clickhouse.com/deb stable main' \
  | sudo tee /etc/apt/sources.list.d/clickhouse.list
sudo apt-get update
sudo apt-get install -y clickhouse-server clickhouse-client
sudo systemctl enable --now clickhouse-server
```

### Initial server setup (SQL)

```sql
CREATE DATABASE IF NOT EXISTS my_clinical_trial;

CREATE USER IF NOT EXISTS zzedc_user
  IDENTIFIED WITH sha256_password BY 'StudyPassword123!';

GRANT ALL ON my_clinical_trial.* TO zzedc_user;
```

### R driver

```{r clickhouse_driver}
install.packages('RClickhouse')
library(RClickhouse)
packageVersion('RClickhouse')
```

### `config.yml` block

```yaml
database:
  backend: clickhouse
  clickhouse:
    host:     'localhost'
    port:     9000
    name:     'my_clinical_trial'
    user:     'zzedc_user'
    password: '${ZZEDC_DB_PASSWORD}'
```

### MergeTree engine and `ORDER BY`

ClickHouse tables require an explicit storage engine and an
`ORDER BY` clause that defines the primary sort order on disk. The
`MergeTree` family is the workhorse:

```{r ch_create}
library(RClickhouse)
conn <- dbConnect(
  RClickhouse::clickhouse(),
  host = 'localhost', port = 9000,
  db = 'my_clinical_trial',
  user = 'zzedc_user',
  password = Sys.getenv('ZZEDC_DB_PASSWORD')
)

dbExecute(conn, "
  CREATE TABLE IF NOT EXISTS subjects (
    subject_id      UInt32,
    study_id        String,
    site_id         String,
    enrollment_date Date,
    status          String,
    treatment_arm   String,
    created_at      DateTime DEFAULT now(),
    updated_at      DateTime DEFAULT now()
  ) ENGINE = MergeTree()
  ORDER BY (site_id, subject_id)
")

dbExecute(conn, "
  CREATE TABLE IF NOT EXISTS audit_trail (
    audit_id     UInt64,
    user_id      UInt32,
    action       String,
    table_name   String,
    record_id    String,
    old_value    String,
    new_value    String,
    action_date  DateTime DEFAULT now(),
    ip_address   String
  ) ENGINE = MergeTree()
  ORDER BY (action_date, user_id)
  TTL action_date + INTERVAL 7 YEAR
")

dbDisconnect(conn)
```

The `TTL` clause on `audit_trail` retains audit records for seven
years (a common regulatory floor) and discards them automatically
afterward.

### Materialised views

Pre-aggregated views support real-time dashboards:

```sql
CREATE MATERIALIZED VIEW IF NOT EXISTS mv_enrollment_summary
ENGINE = SummingMergeTree()
ORDER BY (site_id, enrollment_date)
AS SELECT
  site_id,
  enrollment_date,
  count() AS enrolled_count,
  countIf(status = 'Active') AS active_count
FROM subjects
GROUP BY site_id, enrollment_date;
```

### Backup

Use `clickhouse-backup` for table-level backups:

```bash
clickhouse-backup create my_trial_backup
clickhouse-backup list
clickhouse-backup restore my_trial_backup
```

For Docker deployments, a tar of the data volume is a workable
fallback, taken with the container stopped.

### Backend-specific notes

- ClickHouse's transactional guarantees are limited; do not use it as
  the system of record for transactional CRF entry.
- For genuinely large deployments, the cluster configuration in
  `/etc/clickhouse-server/config.d/cluster.xml` distributes a study
  across shards and replicas. This is out of scope for the quick
  start; consult the
  [ClickHouse documentation](https://clickhouse.com/docs/) before
  committing to a clustered deployment.
- Periodic `OPTIMIZE TABLE ... FINAL` merges small parts and reduces
  the on-disk part count. Schedule it during a maintenance window;
  it is I/O-intensive.

---

## Resources

- [Technical Lead Guide, Chapter 2 (Picking the right backend)](technical-lead-guide.html)
- [Technical Lead Guide, Chapter 5 (Encryption)](technical-lead-guide.html)
- [Technical Lead Guide, Chapter 9 (Disaster recovery)](technical-lead-guide.html)
- [Demonstration Trial Setup](demonstration-trial-setup.html)
- [Demonstration Trial via Google Sheets](demonstration-trial-via-gsheets.html)
- `?create_db_adapter`, `?default_db_config`,
  `?create_wizard_database`, `?migrate_between_backends`
- Driver documentation:
  [RSQLite](https://rsqlite.r-dbi.org),
  [duckdb](https://duckdb.org/docs/api/r),
  [RMariaDB](https://rmariadb.r-dbi.org),
  [RPostgres](https://rpostgres.r-dbi.org),
  [RClickhouse](https://github.com/IMSMWU/RClickhouse)

**Support:**

- Email: rgthomas@ucsd.edu
- GitHub: <https://github.com/rgt47/zzedc/issues>
