---
title: "Demonstration Clinical Trial via Google Sheets"
subtitle: "Active vs. Placebo Memory Enhancement Study, authored in Sheets and imported into ZZedc"
author: "ZZedc Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demonstration Clinical Trial via Google Sheets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette is the Google-Sheets-driven companion to the
CSV-based [Demonstration Clinical Trial Setup](demonstration-trial-setup.html) walk-
through. The study is the same — *Active vs. Placebo Memory
Enhancement Study with 20 subjects* — but the data dictionary,
validation rules, and user roster are authored in Google Sheets
rather than in CSV files. A single import call materialises them
into the ZZedc database.

This path is intended for clinical teams that prefer a familiar
spreadsheet interface for ongoing CRF and validation-rule
maintenance:

- Study coordinators without R fluency can edit forms and rules
  directly in a browser
- Multi-author edits and revision history are handled by Google
  Sheets natively
- Bidirectional sync allows Sheets and the ZZedc database to stay
  aligned through the life of the trial

The CSV path remains available and is preferred for studies that
will live entirely on a single laptop, in air-gapped environments,
or anywhere Google Workspace is not an option.

---

## Prerequisites

ZZedc must be installed and at least the SQLite backend
configured. Follow [Getting Started](getting-started.html) first
if you have not done so.

### Install the optional Google Sheets dependency

`googlesheets4` is in `Suggests`, not `Imports`. Install it
explicitly the first time you take this path:

```{r prereq_install}
install.packages("googlesheets4")
```

### Choose an authentication mode

ZZedc supports both authentication modes that `googlesheets4`
provides:

- **Interactive OAuth** — recommended for development and small
  teams. The first call opens a browser, prompts for consent, and
  caches a refresh token at `~/.cache/gargle/`.
- **Service account** — required for unattended jobs (CI, cron,
  containerised deployments). Create a service-account key in
  Google Cloud Console, share the target Sheet with the service
  account's email address, then point `googlesheets4` at the JSON
  key.

```{r auth_interactive}
library(googlesheets4)

# Interactive mode (one-time browser consent):
gs4_auth(email = "your-google-account@example.org")
```

```{r auth_service_account}
# Service-account mode (unattended):
gs4_auth(
  path = "/path/to/zzedc-service-account.json",
  scopes = "https://www.googleapis.com/auth/spreadsheets"
)
```

---

## Trial structure (recap)

Identical to the CSV vignette:

```
Timeline:
  Visit 1 (Baseline) -> Randomization -> Visit 2 (Week 4) -> Visit 3 (Week 8)

Forms:
  1. Enrollment    (subject_id, enrollment_date, age, gender,
                    education_years, medical_history)
  2. Randomization (subject_id, randomization_date,
                    treatment_assignment, study_coordinator)
  3. MMSE          (subject_id, visit_label, assessment_date,
                    7 component scores, mmse_total,
                    adverse_events, ae_description)

Population:
  - 20 subjects, 1:1 randomization (10 active / 10 placebo)
```

---

## Sheet structure

A single Google Workbook with five tabs holds everything the
demonstration trial needs. The tab names and column schemas below match
what ZZedc's importers expect.

### Tab 1: `Study_Users`

Authentication / role roster. Plain-text passwords are hashed by
ZZedc on import; never store production credentials in a Sheet
intended to be shared.

| Column          | Required | Example                  |
|-----------------|----------|--------------------------|
| username        | yes      | `jane_smith`             |
| full_name       | yes      | Jane Smith               |
| email           | yes      | jane@example.org         |
| role            | yes      | Coordinator              |
| password_initial| yes      | jane123                  |
| site_id         | no       | 001                      |
| active          | no       | TRUE                     |

Roles must match values returned by `zzedc::list_user_roles()`
(`Admin`, `PI`, `Coordinator`, `Researcher`, `DataManager`,
`Monitor` in the default schema).

### Tab 2: `Data_Dictionary`

CRF field definitions, one row per field across all forms. The
column shape mirrors the CSV templates in the
[CSV vignette](demonstration-trial-setup.html#part-3-data-dictionary-csv-files):

| Column        | Required | Example                              |
|---------------|----------|--------------------------------------|
| form_code     | yes      | `mmse`                               |
| field_name    | yes      | `attention`                          |
| field_label   | yes      | Attention and Calculation            |
| field_type    | yes      | numeric                              |
| required      | yes      | TRUE                                 |
| validation    | no       | `between 0 and 5`                    |
| description   | no       | Score 0-5 points                     |
| field_order   | no       | 4                                    |

### Tab 3: `validation_rules`

DSL rules in plain-English syntax. The schema matches what
`import_validation_rules_from_gsheets()` expects (see
`?import_validation_rules_from_gsheets` for the full column list).

| Column        | Required | Example                              |
|---------------|----------|--------------------------------------|
| rule_id       | yes      | `MMSE_TOTAL_RANGE`                   |
| field_code    | yes      | `mmse_total`                         |
| rule_dsl      | yes      | `mmse_total between 0 and 30`        |
| form_code     | no       | `mmse`                               |
| rule_name     | no       | MMSE total in valid range            |
| error_message | no       | MMSE total must be 0-30              |
| severity      | no       | ERROR (default)                      |
| rule_category | no       | FIELD (default)                      |
| is_active     | no       | TRUE (default)                       |

A small starter set for the demonstration trial:

| rule_id            | field_code         | rule_dsl                                          |
|--------------------|--------------------|---------------------------------------------------|
| AGE_RANGE          | age                | `age between 50 and 90`                           |
| MMSE_TOTAL         | mmse_total         | `mmse_total between 0 and 30`                     |
| MMSE_SUM_CONSISTENT| mmse_total         | `mmse_total = orientation_time + orientation_place + registration + attention + recall + language + visual_spatial` |
| RAND_AFTER_ENROL   | randomization_date | `randomization_date >= enrollment_date`           |
| VISIT_WINDOW       | assessment_date    | `assessment_date - randomization_date between -1 and 70` |

### Tab 4: `Visit_Schedule` (optional)

Visit windows and form assignments per visit. ZZedc derives
defaults if absent.

| visit_code  | visit_name | nominal_day | window_minus | window_plus | forms              |
|-------------|------------|-------------|--------------|-------------|--------------------|
| BASELINE    | Baseline   | 0           | 0            | 0           | enrollment,mmse    |
| WK4         | Week 4     | 28          | 3            | 3           | mmse               |
| WK8         | Week 8     | 56          | 3            | 7           | mmse               |

### Tab 5: `Subjects` (optional, for seeded data)

Used only if you want to pre-populate subject records, as the demonstration
trial does. In a real study this tab is omitted; subjects are
entered through the EDC interface as they enrol.

Identical schema to `subjects.csv` in the CSV vignette.

---

## Generating a template

ZZedc provides a helper that creates a Google Sheet pre-populated
with the validation-rules schema and example rows:

```{r template_validation}
library(zzedc)

template <- create_validation_rules_template(
  sheet_name       = "ToyTrial_ValidationRules",
  include_examples = TRUE
)
template$sheet_url
```

The helper covers the `validation_rules` tab specifically. For the
`Study_Users` and `Data_Dictionary` tabs, use the column lists in
the section above as a manual template, or copy from the CSV files
in `vignettes/demonstration-trial/csv_templates/` and paste into your Sheet.

---

## Importing a populated workbook

Once the workbook has the four required tabs filled in
(`Study_Users`, `Data_Dictionary`, `validation_rules`, optionally
`Visit_Schedule` and `Subjects`), import it into a fresh ZZedc
database with a single call:

```{r import_one_call}
library(zzedc)

db_path <- "~/zzedc_demonstration_trial_gsheets/data/demonstration_trial.db"
dir.create(dirname(db_path), recursive = TRUE,
           showWarnings = FALSE)

sheet_id <- "1AbC...your-sheet-id..."  # from the Sheet URL

result <- setup_zzedc_from_gsheets(
  sheet_id     = sheet_id,
  db_path      = db_path,
  imported_by  = "data_manager",
  overwrite    = TRUE
)

# Per-stage status
result$success                   # overall pass/fail
result$users$imported            # number of users imported
result$forms$imported_forms      # number of CRFs created
result$forms$imported_fields     # number of fields across all CRFs
result$validation_rules$imported # number of validation rules imported

# The encryption key is needed for any subsequent operations in
# this R session.
Sys.setenv(DB_ENCRYPTION_KEY = result$key)
```

`setup_zzedc_from_gsheets()` runs four stages in order:

1. Initialise the encrypted SQLite database at `db_path`.
2. Read `Study_Users` and create user accounts.
3. Read `Data_Dictionary`, create one CRF per `form_code` value,
   and add each field as defined.
4. Read `validation_rules` and import the DSL rules through
   [import_validation_rules_from_gsheets()] (which itself
   validates DSL syntax row-by-row).

If any stage fails, the others still run and the failures are
collected in stage-specific `errors` lists; the top-level
`success` flag is `TRUE` only if every stage passed cleanly.

### Validating before importing

Use `dry_run = TRUE` to parse and validate the workbook without
writing to the database. The dry-run pass exercises the DSL
parser on every validation rule and reports counts the same way:

```{r dryrun}
preview <- setup_zzedc_from_gsheets(
  sheet_id  = sheet_id,
  db_path   = db_path,
  dry_run   = TRUE
)
preview$validation_rules$skipped  # rule rows with syntax errors
preview$validation_rules$errors   # named list of error messages
```

### Importing only the validation rules

If you already have an initialised database and only need to
refresh the validation-rule tab, call the lower-level importer
directly:

```{r import_rules_only}
result <- import_validation_rules_from_gsheets(
  sheet_id        = sheet_id,
  sheet_name      = "validation_rules",
  imported_by     = "data_manager",
  validate_syntax = TRUE,
  dry_run         = FALSE
)
```

This is the underlying call `setup_zzedc_from_gsheets()` makes in
its fourth stage.

---

## Bidirectional sync

For studies that maintain CRFs and validation rules in Google
Sheets as the source of truth, ZZedc supports periodic re-import:

```{r sync}
sync_result <- sync_dsl_rules_from_gsheets(
  sheet_id     = sheet_id,
  imported_by  = "data_manager",
  delete_missing = FALSE
)

sync_result$added
sync_result$updated
sync_result$deleted
```

`delete_missing = FALSE` (default) treats Sheet rows as additive:
rules absent from the Sheet are left alone in the database. Set
to `TRUE` when the Sheet is the canonical authority and missing
rules should be retired.

For audit purposes, every sync run writes a record to the
`audit_trail` table tagged with the `sheet_id`, the timestamp,
and the user who ran it.

---

## Verification

```{r verify}
conn <- connect_encrypted_db(db_path = db_path,
                              key = init_result$key)
on.exit(DBI::dbDisconnect(conn), add = TRUE)

# Roster
DBI::dbGetQuery(conn, "SELECT username, role FROM edc_users")

# Form / field counts
DBI::dbGetQuery(conn, "
  SELECT crf_code, COUNT(*) AS n_fields
  FROM crf_definitions cd
  JOIN designer_fields df ON df.design_id = cd.crf_id
  GROUP BY crf_code
")

# Validation rules
DBI::dbGetQuery(conn, "
  SELECT rule_id, field_code, severity FROM validation_rules
")
```

---

## Launching the application

Once the database is populated, the launch path is identical to
the CSV vignette:

```{r launch}
library(zzedc)
Sys.setenv(ZZEDC_DB_PATH = db_path)
launch_zzedc()
```

The Shiny session opens in your default browser. Authenticate as
one of the users you imported and proceed to data entry. The
audit log records every action, the validation rules fire as
defined in the Sheet, and exports are available through the
secure-export module.

---

## Operating model

In practice, the Google-Sheets path is most useful when:

- The data manager and study coordinator share the Sheet; the
  data manager owns validation rules and the coordinator owns
  CRF text and the user roster.
- A non-trivial CRF revision is anticipated mid-study; the Sheet
  provides the revision history and conflict-resolution surface
  Google Workspace already maintains.
- The deployment is multi-site and at least one site cannot run R
  locally; that site can still author CRFs in the Sheet, and the
  central data manager imports.

For a strictly single-PI laptop deployment, the CSV path
(`vignette('demonstration-trial-setup')`) is simpler.

---

## Limitations

The Google-Sheets path has reached parity with the CSV path for
study setup, but a few constraints remain:

- Bidirectional sync currently handles the validation-rules tab
  only; users and form definitions are import-only. To revise
  CRFs or the user roster after initial setup, edit the database
  directly through the ZZedc admin interface or rerun
  `setup_zzedc_from_gsheets()` against a fresh `db_path`.
- Service-account quotas apply: at the time of writing,
  `googlesheets4` defaults to the standard Google Sheets API
  rate limits (300 read requests per minute per user). Large
  deployments should batch reads via `range_read()` rather than
  per-tab calls.
- Stage-level errors surface in the result object's per-stage
  `errors` lists rather than aborting the import. Inspect
  `result$success` and the per-stage objects before assuming a
  successful import.

---

## Troubleshooting

**Authentication fails with `gs4_auth()`**: clear the cached
token (`gargle::gargle_oauth_cache()` shows the location) and
re-authenticate. If running unattended, switch to a service
account.

**`Missing required columns`**: the importer requires `rule_id`,
`field_code`, `rule_dsl` for the validation tab. Match column
names exactly (case-sensitive).

**`Syntax error` in DSL**: rerun the import with
`dry_run = TRUE` and inspect `result$errors` to identify the
offending row. The DSL parser accepts plain-English forms like
`age between 18 and 120`, `enrollment_date <= randomization_date`,
`mmse_total = orientation_time + ...`. See the validation-rules
chapter of the
[Content-Author Guide](content-author-guide.html) (Chapter 3)
for the full grammar.

**Sheet not found**: verify the Sheet has been shared with the
service account email (when using a service account) or with
the user account that authenticated (when using interactive
OAuth).

---

## Resources

- [Demonstration Clinical Trial Setup (CSV path)](demonstration-trial-setup.html)
- [Content-Author Guide, Chapter 3 (Authoring the workbook)](content-author-guide.html)
- `?setup_zzedc_from_gsheets`
- `?import_validation_rules_from_gsheets`
- `?sync_dsl_rules_from_gsheets`
- `?create_validation_rules_template`
- [googlesheets4 documentation](https://googlesheets4.tidyverse.org)
