---
title: "Getting Started with ZZedc"
subtitle: "A role-based router to the right runbook for your job"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ZZedc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Welcome

ZZedc is an Electronic Data Capture (EDC) system for clinical
trials and research studies. It provides:

- AES-256 encryption at rest with SQLCipher (and TDE on
  server backends)
- A hash-chained audit log that supports 21 CFR Part 11
  continuity-of-record claims
- Role-based access control with site-aware siloing for
  multi-site trials
- A choice of five database backends (SQLite, DuckDB,
  PostgreSQL, MySQL/MariaDB, ClickHouse)
- Google Sheets and CSV authoring paths so the trial's
  content owner does not need to write code
- A one-call REDCap migration importer

This page is a **router**, not a tutorial. It points you to
the canonical guide for your role, then to reference docs you
will need along the way.

---

# Pick your role

## I am setting up the trial content

You design the case-report forms, write the validation rules,
build the user roster, and train site staff. Skip everything
else on this page and go to:

> **[Content-Author Guide](content-author-guide.html)**

You will also use the
[Demonstration Trial via Google Sheets](demonstration-trial-via-gsheets.html)
example as a worked sample.

## I am deploying and maintaining the application

You install the package, provision the server, configure
encryption, run backups, and respond to operational
escalations. Skip everything else and go to:

> **[Technical-Lead Guide](technical-lead-guide.html)**

The technical-lead guide cross-references everything else on
this page; you do not have to consult them in any particular
order.

## I am both (solo investigator)

For a single-investigator trial running locally on a laptop,
the same person plays both roles. Read both guides in order:
[Content-Author Guide](content-author-guide.html) first to
understand what content you will produce, then
[Technical-Lead Guide](technical-lead-guide.html) (especially
Chapters 3, 5, 6, 8, 9) for the deployment.

---

# Reference docs

These reference vignettes are cross-linked from the canonical
guides. Read them when the canonical guide tells you to.

## Backend quickstarts

One per supported database. Pick the one matching your
deployment:

- [`vignette('backend-quickstart')`](backend-quickstart.html) —
  decision tree across SQLite, DuckDB, MySQL/MariaDB,
  PostgreSQL, and ClickHouse, with per-backend appendices

## Specific topics

- [`vignette('mysql-redcap-migration-roadmap')`](mysql-redcap-migration-roadmap.html)
  — REDCap import phases, REST API + MySQL paths, audit replay

Encryption, disaster recovery, and AWS deployment are covered
in the Technical-Lead Guide (Chapters 5, 9, and 8 respectively).
Custom Shiny modules and REST API patterns live in the
contributor docs at `docs/development/custom-modules-and-api.md`.

## Worked examples

- [`vignette('demonstration-trial-setup')`](demonstration-trial-setup.html)
  — a synthetic memory study, set up via the CSV path with
  multiple backend examples
- [`vignette('demonstration-trial-via-gsheets')`](demonstration-trial-via-gsheets.html)
  — the same study, set up via Google Sheets

## White papers

- [`vignette('zzedc-whitepaper')`](zzedc-whitepaper.html)
  — architectural and design rationale, regulatory framing

---

# Installing from CRAN

```{r}
install.packages('zzedc')
```

This pulls all required dependencies. Optional database
drivers are installed separately:

```{r}
install.packages('RPostgres')   # PostgreSQL
install.packages('RMariaDB')    # MySQL / MariaDB
install.packages('duckdb')      # DuckDB
```

For full installation guidance including renv setup, see
[Technical-Lead Guide](technical-lead-guide.html) Chapter 3.

---

# Where to ask for help

- **GitHub issues**: bug reports, feature requests, and
  technical questions at the project's repository.
- **CRAN page**: latest version, NEWS, downloads.
- **Email the maintainer**: for security findings or
  regulatory questions you do not want public.

---

# Where this guide used to live

If you arrived here looking for a quickstart, the content has
been reorganised into role-based runbooks (above). Five older
vignettes were retired in favour of the two canonical guides:

- `quick-start-solo-researcher`, `small-project-guide`, and
  `medium-project-guide` are subsumed by
  [Content-Author Guide](content-author-guide.html) and
  [Technical-Lead Guide](technical-lead-guide.html).
- `gsheets-authoring-guide` is now Chapter 3 of the
  Content-Author Guide.
- The setup-wizard reference (the UI-driven alternative to
  the Google Sheets path) is at
  [`vignette('setup-wizard-guide')`](setup-wizard-guide.html).

The retired vignettes overlapped in ways that made it hard for
new users to know where to start.
The new structure is two canonical role-based runbooks plus
focused reference docs.
