---
title: "MySQL Backend and the REDCap Migration Roadmap"
subtitle: "What is shipped today, and what remains for direct REDCap-to-ZZedc import"
author: "ZZedc Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MySQL Backend and the REDCap Migration Roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Purpose of this vignette

This vignette documents two things together:

1. The **current MySQL / MariaDB capability** in ZZedc: what is
   shipped, what works, and what its limitations are.
2. The **roadmap to direct REDCap migration**: a planned
   one-call helper that ingests a REDCap MySQL export (or live
   REDCap database) and produces a populated ZZedc study,
   without an intermediate CSV step.

The first is a feature description for adopters making a
backend choice today. The second is a transparent statement of
work for prospective contributors and for reviewers reading the
manuscript: a feature claimed for the future is named, scoped,
and risk-noted, rather than implied.

---

## Part 1: Current MySQL capability

### What ships

As of this release, ZZedc supports MySQL and MariaDB as a
peer backend alongside SQLite, DuckDB, PostgreSQL, and
ClickHouse. The integration consists of:

- A `MySQLAdapter` R6 class (`R/db_adapter_mysql.R`) implementing
  the standard `DatabaseAdapter` interface (connect, query,
  execute, transaction, dialect, pool, list_tables,
  backend_name).
- Backend dispatch in `create_db_adapter()` accepting both
  `"mysql"` and `"mariadb"` as the `db_backend` value.
- Configuration validation in `validate_db_config()` for the
  `mysql` config section.
- A `default_db_config("mysql", ...)` constructor for
  programmatic use.
- The MySQL appendix in
  [`vignette('backend-quickstart')`](backend-quickstart.html)
  (Appendix C) walks through server setup, R installation, and
  the end-to-end study-creation flow.

The R driver is `RMariaDB`. The same driver connects to both
MySQL and MariaDB servers.

### What works

Everything ZZedc does on the SQLite, PostgreSQL, or DuckDB
backend works on MySQL / MariaDB:

- CRF authoring through `create_crf_definition()`,
  `add_form_field()`, and the data-dictionary CSV path
- Validation rules through the DSL and through
  `setup_zzedc_from_gsheets()`
- User management, role-based access control, audit trail with
  hash-chained integrity, and electronic signatures
- Cross-backend migration via `migrate_between_backends()` (a
  study can be moved between any two of the five backends)
- The full Shiny application, including the form designer,
  data-correction module, audit viewer, and GDPR data-subject
  modules

### Recommended server versions

| Server | Versions | Reason for floor |
|---|---|---|
| MariaDB 11.4 LTS | June 2024 -- June 2029 | Current LTS; headline target |
| MariaDB 10.11 LTS | February 2023 -- February 2028 | Documented compatibility floor |
| MySQL 8.4 LTS | April 2024 -- April 2032 | Current LTS for MySQL |

The technical floor (CTE, window function, JSON support) is
MySQL 8.0 / MariaDB 10.5, but those versions are now end-of-
life and should not be used in clinical settings.

### Known limitations

1. **Encryption at rest is not driven by the adapter.** On
   SQLite, the package can request SQLCipher encryption per
   database file. On MySQL / MariaDB, encryption is configured
   at the server level (InnoDB tablespace encryption); the
   `key` argument to `connect_encrypted_db()` and
   `initialize_encrypted_database()` is ignored on the MySQL
   path. This is a deployment-architecture issue rather than a
   gap in the adapter.

2. **No `RETURNING` clause.** PostgreSQL's `RETURNING` clause
   simplifies inserts that need to capture the new primary
   key. MySQL 8.0 lacks it; MariaDB 10.5+ supports it on
   `INSERT` only. The dialect helper reports
   `supports_returning = FALSE` on the MySQL backend, and
   internal callers fall back to `LAST_INSERT_ID()` /
   `dbGetQuery("SELECT LAST_INSERT_ID()")` for primary-key
   capture.

3. **Identifier quoting differs.** MySQL uses backticks
   (\`identifier\`) where PostgreSQL uses double quotes
   ("identifier"). The adapter's `dialect()$quote_identifier`
   handles this; ad-hoc SQL written by users that interpolates
   identifier names should call the dialect helper rather than
   hard-coding quote characters.

4. **No native LISTEN/NOTIFY.** PostgreSQL's pub-sub primitive
   is not available. Code that depends on
   `PostgreSQLAdapter$listen()` / `notify()` will not run on
   the MySQL backend; the relevant Shiny modules avoid this
   dependency.

### What this enables today

For studies that already live on a MySQL or MariaDB server
(institutional or self-hosted), ZZedc is now installable in
the same operational environment as the existing data
infrastructure. For investigators whose institution operates
REDCap, the indirect REDCap-to-ZZedc path through CSV export
is the supported route today, as documented in
[`vignette('backend-quickstart')`](backend-quickstart.html)
Appendix C.

---

## Part 2: REDCap migration roadmap

### Phase C1: shipped in this release

The Phase C1 connector is `import_redcap_to_zzedc()`. Given a
DBI connection to a REDCap-shaped database and a project ID, it
emits a directory of CSV files in ZZedc data-dictionary shape:

```{r c1-example}
library(RMariaDB)
conn <- DBI::dbConnect(MariaDB(),
  host     = "redcap.example.org",
  dbname   = "redcap",
  user     = "redcap_reader",
  password = Sys.getenv("REDCAP_DB_PASSWORD"))
on.exit(DBI::dbDisconnect(conn), add = TRUE)

result <- import_redcap_to_zzedc(
  conn       = conn,
  pid        = 42L,
  output_dir = "~/zzedc_migration/redcap_42"
)

result$counts          # rows per artefact
length(result$skipped_rules)  # rules needing manual review
```

The output directory contains:

- `data_dictionary.csv` — REDCap fields translated to ZZedc
  CRF-field shape (one row per field; ready for the data-
  dictionary import path)
- `validation_rules.csv` — REDCap range validations translated
  to ZZedc DSL (e.g., `age between 50 and 90`)
- `users.csv` — REDCap user accounts with REDCap rights
  translated to ZZedc roles (Admin, Researcher, Coordinator,
  Monitor)
- `subjects.csv` — distinct record IDs from
  `redcap_data.record`
- `subject_data.csv` — subject responses in long form
  (subject_id, field_name, value, event_id)
- `audit_log.csv` — REDCap audit-log rows for human inspection
- `manifest.json` — provenance record (REDCap project ID,
  imported-at timestamp, counts, importer phase)

### What Phase C1 covers

- **Field-type translation**: REDCap `text` + `integer`-validation
  becomes ZZedc `numeric`; `radio`, `checkbox`, `dropdown`,
  `yesno`, `date`, `datetime` map directly. Unknown types fall
  back to `text` (recorded in the manifest).
- **Range validations**: `text_validation_min` /
  `text_validation_max` on `integer` / `number` / `float`
  validations translate to ZZedc DSL `between … and …`.
- **User-role coercion**: REDCap's per-field permission flags
  collapse to ZZedc's coarser role set
  (`design` → Admin; `data_export_tool` → Researcher;
  general entry → Coordinator; otherwise Monitor). Expired
  REDCap accounts (`expiration < today`) map to
  `active = FALSE`.
- **EAV record extraction**: `redcap_data` rows are emitted in
  long form. Wide-form pivot is deferred to Phase C2 because it
  requires repeating-instrument and longitudinal-event
  semantics that Phase C1 does not yet handle.

### What Phase C1 does not cover

These are the explicit limitations of the C1 release:

- **No direct database creation.** The caller must run a second
  step: stage the CSVs into a Google Sheet and call
  `setup_zzedc_from_gsheets()`, or use the data-dictionary
  CSV authoring path.
- **No audit-trail hash-chain replay.** The audit CSV is
  emitted for human inspection. ZZedc's `audit_trail` table is
  not populated by C1; that is Phase C2's job.
- **No branching-logic translation.** Fields with REDCap
  branching logic appear in the data dictionary, but the
  branching rule is recorded in `result$skipped_rules` rather
  than translated to ZZedc DSL. The branching parser is
  Phase C2.
- **Long-form data only.** Repeating instruments and
  longitudinal events are not pivoted to ZZedc form-submission
  rows. The downstream stage (Phase C2 or manual CSV authoring)
  resolves this.
- **DBI connection only.** Source modes other than a live DBI
  connection (`.sql`-dump file hydration, REDCap REST API)
  are deferred to Phase C3.

### C1 recipe (CSV emission)

When a CSV intermediate is wanted (for human review of the
extracted artefacts before populating the ZZedc database), use
the C1 entry point:

```{r recipe-c1}
# 1. Connect to your REDCap source database (institutional
#    MySQL, SQLite hydrated from a dump, or any DBI connection
#    whose schema matches REDCap's relational layout).
conn <- DBI::dbConnect(RMariaDB::MariaDB(),
  host = "redcap.example.org", dbname = "redcap",
  user = "redcap_reader",
  password = Sys.getenv("REDCAP_DB_PASSWORD"))

# 2. Run the C1 importer.
out <- "~/zzedc_migration/redcap_42"
result <- import_redcap_to_zzedc(conn, pid = 42L,
                                  output_dir = out)

# 3. Review the manifest and `skipped_rules`. Translate any
#    branching-logic rules manually into the
#    validation_rules.csv following the patterns in
#    vignette('content-author-guide').

# 4. Stage data_dictionary.csv, validation_rules.csv, and
#    users.csv into a Google Sheets workbook and run
#    setup_zzedc_from_gsheets() against it; or use the CSV
#    authoring path documented in
#    vignette('demonstration-trial-setup').
DBI::dbDisconnect(conn)
```

### Phase C2: shipped in this release

Phase C2 is the one-call direct-to-database orchestrator
`import_redcap_to_zzedc_db()`. It runs the C1 extraction
internally, then provisions the ZZedc database, populates it
with users, CRFs, subjects, and form data, and replays the
REDCap audit log into ZZedc's hash-chained `audit_log` table.

```{r recipe-c2}
library(RMariaDB)
conn <- DBI::dbConnect(MariaDB(),
  host     = "redcap.example.org",
  dbname   = "redcap",
  user     = "redcap_reader",
  password = Sys.getenv("REDCAP_DB_PASSWORD"))
on.exit(DBI::dbDisconnect(conn), add = TRUE)

result <- import_redcap_to_zzedc_db(
  conn       = conn,
  pid        = 42L,
  db_path    = "/srv/zzedc/MIGRATED-001/study.db",
  overwrite  = FALSE,
  dry_run    = FALSE
)

# Required for any subsequent operation in this R session.
Sys.setenv(DB_ENCRYPTION_KEY = result$key)

# Per-stage diagnostics
result$success                       # overall pass/fail
result$users$imported                # users created
result$forms$imported_forms          # CRFs created
result$forms$imported_fields         # fields across all CRFs
result$subjects$imported             # subject records seeded
result$subject_data$imported         # subject form_data rows
result$audit_replay$imported         # REDCap audit rows replayed
result$audit_replay$chain_validates  # hash chain valid end-to-end
length(result$skipped_rules)         # rules deferred to manual review
result$branching_translated          # branching rules translated
```

### What Phase C2 covers

- **One-call orchestration.** Database creation, schema
  provisioning, user import, CRF authoring, subject seeding,
  form-data load, and audit replay run end-to-end.
- **Hash-chained audit replay.** REDCap `redcap_log_event` rows
  are replayed into ZZedc's `audit_log` and `audit_chain` in
  chronological order. Each replayed row's `audit_hash` chains
  to the prior row; the first row chains to `GENESIS`. After
  replay, the chain is validated end-to-end and reported in
  `result$audit_replay$chain_validates`. Imported rows have
  `details` JSON marked
  `{"source":"redcap","redcap_log_event_id":...}` so
  REDCap-origin actions remain attributable post-import; this
  is the technical control that supports a Part-11
  continuity-of-record claim.
- **Branching-logic translation.** A simple grammar handles
  REDCap's `[field] = value`-style expressions (equality,
  inequality, `>=`/`<=`/`>`/`<`, `or`/`and`/`not`,
  parenthesised compounds). Constructs outside that grammar
  (`datediff()`, event-arm prefixes such as
  `[event_arm_1][field]`) are conservatively rejected and
  recorded in `result$skipped_rules` for manual review.
- **Conflict policy.** `on_conflict = "skip"` (default),
  `"merge"`, or `"fail"` controls behaviour when imported rows
  conflict with existing ZZedc rows. Default is `"skip"` with
  per-user diagnostics in the result list.
- **Dry-run mode.** `dry_run = TRUE` exercises the full
  pipeline (including CSV emission and DSL parsing) without
  creating the target database. Useful to validate a REDCap
  source against the C2 importer before committing.

### What Phase C2 does not yet cover

These are explicit limitations of the C2 release:

- **Repeating instruments and longitudinal events** are
  flattened: each `redcap_data.event_id` becomes a single
  `visit_code` (resolved against `redcap_events_metadata` if
  available, otherwise `EVENT_<id>`). Studies whose REDCap
  schema relies on repeating instruments will import but lose
  the repetition structure. The data dictionary may need
  manual restructuring after import.
- **Branching-logic edge cases.** REDCap's
  `datediff([dob], [today], "y")`, event-arm prefixes, and
  custom plugins are not translated. The covered grammar
  matches roughly 60-70% of branching expressions in real
  REDCap projects; the rest land in `skipped_rules` for
  manual translation following the patterns in
  `vignette('content-author-guide')` §"Define validation
  rules".
- **`.sql`-dump source mode** is deferred to Phase C3b. Today,
  dumps must be loaded into a transient MySQL or MariaDB
  instance manually before invoking the importer.

### Phase C3a: shipped in this release (REST API source mode)

In addition to the live DBI connection (`source = "db"`), the
importer now accepts a REDCap REST API endpoint
(`source = "api"`). This removes the requirement for direct
database access -- often unavailable on institutional REDCap
instances -- and lets investigators migrate using nothing more
than a project-scoped API token.

```{r, eval = FALSE}
library(zzedc)

result <- import_redcap_to_zzedc_db(
  source    = "api",
  api_url   = "https://redcap.example.org/api/",
  api_token = Sys.getenv("REDCAP_API_TOKEN"),
  db_path   = "/srv/zzedc/MIGRATED-002/study.db",
  overwrite = TRUE
)

result$success                              # TRUE
result$audit_replay$completeness            # full / partial / empty
result$audit_replay$marker_inserted         # TRUE if not full
result$audit_replay$chain_validates         # TRUE
length(result$skipped_rules)                # rules needing manual review
```

The token's permissions decide whether C3a can claim full
audit-chain continuity. REDCap's `Logging` right is
permission-gated: tokens minted by a user without it receive a
redacted or empty log. The importer detects this and degrades
the continuity claim explicitly:

- `audit_replay$completeness = "full"` -- the API returned
  data-modification events; the chain is replayed exactly as in
  Phase C2.
- `"partial"` -- the API returned only login/logout/view
  events, no data modifications. A `MIGRATION_AUDIT_GAP` marker
  is prepended to the chain documenting the gap.
- `"empty"` -- the API returned no events at all. A marker is
  still prepended; the chain validates from the marker forward.

In all three cases the resulting hash chain validates
end-to-end, but the marker makes the continuity-of-record
claim honest: continuity is preserved *from the migration
boundary forward*, not retroactively over a gap that the
source-side log did not faithfully expose. A Part-11 reviewer
inspecting the imported study can detect the marker by
filtering on `event_type = 'MIGRATION_AUDIT_GAP'`.

#### Mocking the API for tests and dry runs

For tests, dry runs against synthetic data, or running the
importer offline against canned fixtures, build an `api`
handle directly with the `ops` argument and pass it via
`api = ...`:

```{r, eval = FALSE}
fake_api <- redcap_api_connect(
  api_url   = "ignored",
  api_token = "ignored",
  ops = list(
    version  = function() "14.5",
    metadata = function() my_metadata_df,
    users    = function() my_users_df,
    records  = function(fields = NULL) my_records_df,
    log      = function() my_log_df
  )
)
result <- import_redcap_to_zzedc_db(
  source = "api", api = fake_api,
  db_path = tempfile(fileext = ".db"), overwrite = TRUE
)
```

This is the pattern the package's own tests use to exercise the
API path without depending on a network or a live REDCap
instance.

### What Phase C3a does not yet cover

These remain limitations of C3a, distinct from those of C2:

- **`.sql`-dump source mode** is Phase C3b, not yet shipped.
- **Repeating instruments and longitudinal events** are still
  flattened (same caveat as C2); lifting this is Phase C4 and
  touches the schema, not the extractor.
- **Audit-log filtering by date / user.** C3a fetches the full
  log; large projects on slow API endpoints may benefit from
  windowed pulls. Deferred.
- **OAuth / institutional SSO.** C3a is token-only.

### Why this matters for the manuscript

The package's positioning as a *REDCap alternative* is
substantively stronger when there is a documented migration
path. *Comparable feature set* is one claim; *can ingest your
REDCap export and produce an equivalent study* is another. The
former invites debate about which platform's feature is
better; the latter is a concrete, testable assertion.

### REDCap data model summary

REDCap stores study data in a relational schema with these
core tables (REDCap version 14.x, the current LTS as of this
writing):

| Table | Role |
|---|---|
| `redcap_data` | Long-form (entity-attribute-value) table holding subject responses. One row per (subject, instrument-instance, field). |
| `redcap_metadata` | Field definitions: name, label, type, validation, branching logic. |
| `redcap_events_metadata` | Visit / event definitions. |
| `redcap_events_arms` | Arm definitions for longitudinal projects. |
| `redcap_events_forms` | Which forms appear in which events. |
| `redcap_user_rights` | Per-user role assignments and permissions. |
| `redcap_log_event` | Audit log: every data-entry, edit, export action. |
| `redcap_projects` | Per-project metadata (name, status, etc.). |

REDCap's schema has been roughly stable since version 8.x
(2018), but column additions happen at most major version
bumps. The importer's design needs to accommodate this.

### Mapping REDCap to ZZedc

For each ZZedc concept, the corresponding REDCap source:

| ZZedc concept | REDCap source | Mapping notes |
|---|---|---|
| `crf_definitions` | `redcap_metadata.form_name` (distinct values) | One ZZedc CRF per REDCap form; CRF name = form label |
| `designer_fields` | `redcap_metadata` rows | One ZZedc field per REDCap field; field type translated through a small lookup table (REDCap `text` + validation `integer` -> ZZedc `numeric`; REDCap `radio` -> ZZedc `radio`; etc.) |
| `validation_rules` | `redcap_metadata.text_validation_type_or_show_slider_number` + `redcap_metadata.text_validation_min/max` + `redcap_metadata.branching_logic` | Range validations translate directly; branching logic translates to ZZedc conditional rules through a small parser |
| `subjects` | distinct `redcap_data.record` for project | Subject ID inherited verbatim |
| Subject visit data | `redcap_data` rows joined against `redcap_events_metadata` | EAV-to-relational pivot: one ZZedc form-submission row per (subject, event, instrument) |
| `edc_users` | `redcap_user_rights` joined against the system user table | Roles translated through a fixed mapping (REDCap roles are project-local, ZZedc roles are system-wide) |
| `audit_trail` | `redcap_log_event` rows for the project | Per-action translation; hash-chain replay computed during import |

### Implementation work

| Module | Lines (estimate) | Risk |
|---|---|---|
| `R/redcap_import.R` orchestrator | 200-300 | Low |
| Field-type translation table | 50-100 | Low |
| Branching-logic parser (REDCap syntax to ZZedc DSL) | 200-400 | Medium: REDCap uses a custom syntax with infix operators, conditionals, and field-name references. A small recursive-descent parser is the right tool. |
| EAV-to-relational pivot | 100-200 | Medium: longitudinal projects with repeating instruments require careful join semantics |
| User and role translation | 50-100 | Low |
| Audit-trail replay (with hash-chain construction) | 150-250 | Medium: each ZZedc audit row has a hash that depends on the previous row; importing requires re-computing the chain |
| Dry-run validator | 100 | Low |
| Tests against fixture databases | 300-500 | Medium: need a sanitised REDCap fixture; cannot ship with real PHI |
| Vignette and documentation | 200 | Low |

**Total estimate: 1,300 -- 2,150 lines plus fixtures.**
**Time estimate: 2 -- 3 weeks of focused work.**

### Open design decisions

1. **Source format.** Three sources are plausible:
   - REDCap MySQL `.sql` dump file (loaded into a temporary
     MySQL or transient SQLite via the dump's CREATE TABLE
     statements; SQLite cannot replay arbitrary MySQL dumps so
     a temporary MySQL server is the cleaner option)
   - Live REDCap MySQL database (the importer connects with a
     read-only account)
   - REDCap REST API (the importer pulls metadata and data via
     authenticated HTTP)

   Recommend supporting **all three** with a single `source`
   argument and dispatch on type. The CSV path is already
   covered by `setup_zzedc_from_gsheets()` and the
   data-dictionary CSV authoring flow.

2. **REDCap version range.** Recommend targeting
   **REDCap 14.x LTS** (current as of this writing) as the
   primary tested version, with **13.x** documented as
   compatible-but-untested. Older versions need fixture
   testing before they can be claimed.

3. **Conflict policy.** What if the REDCap project has user
   accounts that conflict with existing ZZedc accounts? Three
   policies, configurable through an argument:

   - `on_conflict = "skip"`: leave the existing ZZedc account
     alone, log a warning
   - `on_conflict = "merge"`: merge attributes (the more
     permissive role wins)
   - `on_conflict = "fail"`: abort the import

   Recommend `"skip"` as default with explicit logging.

4. **Audit-trail provenance.** REDCap's audit log records
   actions performed in REDCap. After import, those actions
   should remain attributable to REDCap, not back-dated as if
   they happened in ZZedc. Recommend a new field
   `audit_trail.source = "redcap"` on imported rows, and
   `audit_trail.source = "zzedc"` on native rows. Existing
   ZZedc deployments will see only `zzedc` rows.

5. **PHI exposure during dry-run.** A dry-run validator that
   reads the real REDCap database produces logs that should
   not contain PHI. Recommend redacting subject IDs and free-
   text fields in the result object's diagnostics; full data
   only flows when `dry_run = FALSE`.

### Validation strategy

1. **Round-trip equivalence.** Build a fixture REDCap database
   (synthetic data only, no PHI), import it into ZZedc, export
   the ZZedc study back to a REDCap-compatible CSV, and verify
   the round-trip preserves every field value. Disagreement is
   a translation bug.

2. **Spot-check against a real REDCap export.** Once the
   round-trip test passes on synthetic data, run the importer
   against a real REDCap export (with an investigator who
   owns the data) and have a domain expert confirm a sample
   of subjects is faithfully reproduced.

3. **Audit-chain integrity.** After import, verify the ZZedc
   audit log's hash chain validates from the first synthetic
   row through to the last imported row. Break-detection on
   the chain is the existing `verify_audit_integrity()`
   function; this is a regression test rather than new code.

### Risks

1. **Schema drift.** REDCap has shipped 14 major versions. A
   schema feature added in 14.3 may not exist in 14.0. The
   importer should declare a tested version range and reject
   schemas it does not recognise rather than silently
   producing partial imports.

2. **Branching-logic edge cases.** REDCap's syntax for
   branching logic accumulated incrementally; some real-world
   expressions are hard to parse correctly. The importer
   should fall back to a "manual review required" diagnostic
   when it encounters a construct it cannot translate, rather
   than guess.

3. **Repeating-instrument semantics.** REDCap's repeating
   forms and repeating events do not map one-to-one to ZZedc's
   single-instance form model. A study using repeating
   instruments may need its data dictionary restructured
   during import; the importer should detect and refuse rather
   than silently flatten.

4. **Custom field types.** REDCap supports user-defined field
   types through plugins; ZZedc has a fixed type set. Custom
   fields will need to be mapped to the closest ZZedc type
   and flagged in the import diagnostics for human review.

5. **Regulatory framing.** A migration that destroys the
   audit chain (rather than preserving it) cannot claim
   continuity for 21 CFR Part 11 purposes. The
   `audit_trail.source = "redcap"` marking and the hash-chain
   replay are how the importer preserves continuity; both
   need explicit verification before the importer is claimed
   as Part-11-compatible.

### Phasing

If the full importer is too much work for one sprint, three
phased deliveries make sense:

- **Phase C1: Read-only REDCap MySQL connector.** A minimal
  importer that reads REDCap metadata + records via an
  RMariaDB connection and emits a ZZedc-format CSV. Does not
  produce a database directly; users still call
  `setup_zzedc_from_gsheets()` (or its CSV equivalent) to
  finish. ~400 lines, 1 week. Lower-risk and ships value
  fast.

- **Phase C2: Direct database creation + audit replay.** Adds
  the database-creation orchestration and the hash-chain
  replay, completing the one-call experience. ~600 additional
  lines, 1-2 weeks.

- **Phase C3a (shipped): REST API source mode.** Adds
  `source = "api"` to `import_redcap_to_zzedc_db()`,
  including audit-completeness gating with a
  `MIGRATION_AUDIT_GAP` marker for partial / empty logs.
  ~400 lines including tests.

- **Phase C3b (planned): `.sql`-dump source mode.** Will
  hydrate a transient MariaDB instance from a REDCap dump and
  run the existing C2 pipeline against it. Mostly orchestration;
  ~200 lines.

Total across phases: roughly the original 2-3 week estimate,
delivered in four releasable increments.

---

## What investigators can do today

Without waiting for the importer:

1. Use REDCap's built-in **Data Exports** to obtain a CSV plus
   the data dictionary, both in REDCap's standard formats.
2. Use the
   [Demonstration Trial Setup](demonstration-trial-setup.html)
   vignette's CSV-driven flow to author equivalent ZZedc CRFs.
   The data-dictionary column shape is similar enough that
   minor manual adjustment is the only friction.
3. Bulk-load the subject CSV into ZZedc through the
   secure-export module's *Import* tab.
4. For studies using Google Sheets as the source of truth,
   use [Demonstration Clinical Trial via Google Sheets](demonstration-trial-via-gsheets.html)
   instead.

This indirect path works today on every supported backend
(SQLite, DuckDB, PostgreSQL, ClickHouse, MySQL / MariaDB).
The roadmap above describes the work needed to make the
direct path comparable in friction.

---

## Resources

- [Backend Quickstart, Appendix C (MySQL/MariaDB)](backend-quickstart.html)
- [Demonstration Trial Setup](demonstration-trial-setup.html)
- [Demonstration Clinical Trial via Google Sheets](demonstration-trial-via-gsheets.html)
- `?MySQLAdapter`, `?create_db_adapter`,
  `?migrate_between_backends`
- [REDCap data dictionary documentation](https://projectredcap.org/resources/)
- [RMariaDB documentation](https://rmariadb.r-dbi.org)
