---
title: "Regulatory Compliance Framework in ZZedc"
subtitle: "A Comprehensive Analysis of GDPR, GCP, and FDA 21 CFR Part 11 Implementation"
author: "ZZedc Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Regulatory Compliance Framework in ZZedc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Executive Summary

ZZedc is a modern Electronic Data Capture (EDC) system designed for clinical
research with comprehensive regulatory compliance capabilities. This white
paper provides a detailed analysis of the system's implementation of three
critical regulatory frameworks: the European Union's General Data Protection
Regulation (GDPR), Good Clinical Practice (GCP) guidelines, and the United
States Food and Drug Administration's 21 CFR Part 11 regulations.

The system implements 32 distinct compliance features across these frameworks,
with over 3,000 automated tests validating regulatory requirements. This
document describes each feature's regulatory basis, technical implementation,
and practical application in clinical research settings.

# Part I: GDPR Compliance Features

The General Data Protection Regulation (EU) 2016/679 establishes comprehensive
data protection requirements for organizations processing personal data of EU
residents. ZZedc implements the following GDPR-mandated features.

## 1. Data Encryption at Rest (Article 32)

### Regulatory Requirement

Article 32 of the GDPR mandates that data controllers implement "appropriate
technical and organisational measures to ensure a level of security appropriate
to the risk" including "the pseudonymisation and encryption of personal data."

### Technical Implementation

ZZedc implements transparent database encryption using SQLCipher, an open-source
extension to SQLite that provides AES-256 encryption. The implementation
includes:

- **Key Management**: Integration with AWS Key Management Service (KMS) for
  enterprise deployments, with environment variable fallback for development
- **Encryption Algorithm**: AES-256 in CBC mode with HMAC-SHA512 for
  authentication
- **Key Derivation**: PBKDF2 with 256,000 iterations for passphrase-based keys

```r
# Example: Initialize encrypted database
initialize_encrypted_database(
  db_path = "study_data.db",
  key_source = "aws_kms"
)
```

### Compliance Verification

The system includes automated verification of encryption status through header
inspection and key validation routines. Audit logs record all key access events
for compliance documentation.

### References

European Parliament. (2016). Regulation (EU) 2016/679 (General Data Protection
Regulation). Article 32: Security of processing.

## 2. Data Subject Access Request (Article 15)

### Regulatory Requirement

Article 15 establishes the right of data subjects to obtain confirmation of
whether personal data concerning them is being processed, and access to that
data along with supplementary information about processing purposes, categories,
and recipients.

### Technical Implementation

The DSAR module provides a complete workflow for handling access requests:

- **Request Intake**: Structured capture of request details with automatic
  deadline calculation (30 days, extendable to 90 days for complex requests)
- **Identity Verification**: Multi-factor verification workflow to prevent
  unauthorized access
- **Data Collection**: Systematic identification and extraction of personal
  data across all system tables
- **Response Generation**: Automated compilation of data packages in
  machine-readable formats

```r
# Example: Create and process DSAR
request <- create_dsar_request(
  subject_email = "participant@example.com",
  subject_name = "John Doe",
  request_type = "ACCESS",
  requested_by = "dpo"
)

# Verify identity before processing
verify_subject_identity(
  request_id = request$request_id,
  verification_method = "DOCUMENT",
  verified_by = "admin"
)
```

### Compliance Verification

All DSAR activities are logged with timestamps, creating an immutable audit
trail demonstrating compliance with response deadlines.

## 3. Right to Rectification (Article 16)

### Regulatory Requirement

Article 16 provides data subjects the right to obtain rectification of
inaccurate personal data and completion of incomplete data.

### Technical Implementation

The rectification system implements a controlled correction workflow that
maintains data integrity while enabling corrections:

- **Request Processing**: Structured intake with specification of incorrect
  data and proposed corrections
- **Verification**: Review process to validate correction requests
- **Audit Trail**: Complete history of original values, corrections, and
  authorizations
- **Third-Party Notification**: Tracking of data recipients requiring
  notification of corrections

The system distinguishes between clinical data corrections (which follow
FDA-mandated workflows) and administrative data corrections (which follow
GDPR-only workflows).

### Clinical Trial Considerations

In clinical research contexts, rectification must be balanced against data
integrity requirements. The system implements a dual-approval workflow for
clinical data corrections while allowing streamlined processing for
administrative corrections.

## 4. Right to Erasure (Article 17)

### Regulatory Requirement

Article 17, commonly known as the "right to be forgotten," provides data
subjects the right to erasure of personal data under specified circumstances,
subject to exceptions for legal obligations and public health research.

### Technical Implementation

The erasure module implements a sophisticated workflow accommodating both GDPR
rights and regulatory retention requirements:

- **Legal Hold Management**: Ability to place data under legal hold when
  regulatory requirements supersede erasure rights
- **Selective Erasure**: Granular control over erasure scope, enabling
  compliance with Article 17(3) exceptions
- **Anonymization Alternative**: When complete erasure conflicts with research
  integrity, the system supports irreversible anonymization as an alternative
- **Third-Party Coordination**: Tracking and notification of downstream
  recipients

```r
# Example: Create legal hold before erasure
hold <- create_legal_hold(
  subject_id = "SUBJ-001",
  hold_type = "REGULATORY",
  hold_reason = "FDA retention requirement - 2 years post-study",
  held_by = "compliance_officer"
)

# Erasure request blocked by legal hold
request <- create_erasure_request(
  subject_email = "participant@example.com",
  erasure_grounds = "WITHDRAWAL",
  requested_by = "dpo"
)
# Returns: status = "LEGAL_HOLD"
```

### Regulatory Conflict Resolution

The system implements automatic detection of conflicts between GDPR erasure
rights and FDA/ICH retention requirements, placing data under regulatory hold
and documenting the legal basis for retention.

## 5. Right to Restriction of Processing (Article 18)

### Regulatory Requirement

Article 18 provides data subjects the right to obtain restriction of processing
in circumstances including accuracy disputes, unlawful processing, and pending
objection reviews.

### Technical Implementation

The restriction module implements processing controls at the field level:

- **Scope Definition**: Restrictions can apply to specific data categories,
  processing purposes, or entire subject records
- **Processing Blocks**: Technical controls preventing restricted data from
  being included in analyses, exports, or reports
- **Allowed Processing**: Configuration of permitted processing activities
  during restriction (storage, legal claims, subject consent)
- **Lift Procedures**: Controlled workflow for removing restrictions with
  subject notification

### Processing Attempt Logging

All attempts to process restricted data are logged, enabling demonstration of
compliance and identification of system components requiring modification.

## 6. Right to Data Portability (Article 20)

### Regulatory Requirement

Article 20 provides data subjects the right to receive their personal data in
a structured, commonly used, machine-readable format and to transmit that data
to another controller.

### Technical Implementation

The portability module supports multiple export formats and transfer mechanisms:

- **Export Formats**: JSON, XML, CSV, and CDISC ODM for clinical data
- **Data Scope**: Configurable inclusion of provided data versus derived data
- **Direct Transfer**: Controller-to-controller transfer with verification
- **Security**: Encrypted exports with integrity verification

```r
# Example: Generate portable data package
export <- generate_portability_export(
  request_id = request$request_id,
  format = "JSON",
  include_metadata = TRUE
)

# Initiate controller transfer
transfer <- initiate_controller_transfer(
  request_id = request$request_id,
  recipient_controller = "New Research Institution",
  transfer_method = "SECURE_API"
)
```

## 7. Right to Object (Article 21)

### Regulatory Requirement

Article 21 provides data subjects the right to object to processing based on
legitimate interests or public interest, and an absolute right to object to
direct marketing.

### Technical Implementation

The objection module manages processing objections with configurable scope:

- **Objection Types**: General processing, profiling, direct marketing, and
  research objections
- **Processing Cessation**: Immediate cessation of objected processing
  activities
- **Override Capability**: Documented override for compelling legitimate
  grounds with supervisory authority notification
- **Marketing Preferences**: Granular channel-level marketing opt-out

### Research Exception Handling

For scientific research processing, the system implements the Article 21(6)
exception, requiring documentation of necessity for research objectives.

## 8. Consent Management (Articles 6-7)

### Regulatory Requirement

Articles 6 and 7 establish consent as a lawful basis for processing and specify
requirements for valid consent including freely given, specific, informed, and
unambiguous indication.

### Technical Implementation

The consent management system provides comprehensive consent lifecycle
management:

- **Purpose-Specific Consent**: Granular consent collection for distinct
  processing purposes
- **Consent Records**: Immutable records of consent including timestamp,
  method, and consent text version
- **Withdrawal Processing**: Immediate effect of withdrawal with downstream
  processing cessation
- **Consent Refresh**: Automated identification of aging consents requiring
  refresh

```r
# Example: Record granular consent
record_consent(
  subject_id = "SUBJ-001",
  purpose_id = purpose$purpose_id,
  consent_given = TRUE,
  consent_method = "ELECTRONIC",
  consent_text_version = "1.2",
  recorded_by = "coordinator"
)

# Check consent before processing
check <- check_consent(
  subject_id = "SUBJ-001",
  purpose_code = "RESEARCH_PRIMARY"
)
```

## 9. Data Retention Enforcement (Article 5)

### Regulatory Requirement

Article 5(1)(e) establishes the storage limitation principle, requiring that
personal data be kept no longer than necessary for processing purposes.

### Technical Implementation

The retention module implements policy-based retention management:

- **Retention Policies**: Configurable policies by data category with legal
  basis documentation
- **Automated Review**: Scheduled identification of data reaching retention
  limits
- **Disposal Actions**: Configurable disposal methods (deletion, anonymization,
  archival)
- **Legal Hold Integration**: Automatic suspension of disposal for data under
  legal hold

### Retention Conflict Resolution

When GDPR retention limits conflict with regulatory requirements (e.g., FDA
15-year retention for clinical data), the system documents the legal basis
and applies the longer retention period.

## 10. Privacy Impact Assessment (Article 35)

### Regulatory Requirement

Article 35 mandates Data Protection Impact Assessments for processing
operations likely to result in high risk to data subjects' rights and freedoms.

### Technical Implementation

The PIA module provides a structured assessment framework:

- **Risk Categories**: Systematic identification of processing risks across
  confidentiality, integrity, and availability
- **Risk Scoring**: Quantitative risk assessment with likelihood and impact
  ratings
- **Mitigation Tracking**: Documentation of risk mitigation measures
- **DPO Consultation**: Workflow for mandatory DPO review
- **Supervisory Authority Consultation**: Tracking for Article 36 prior
  consultation requirements

```r
# Example: Create and assess PIA
pia <- create_pia_assessment(
  assessment_name = "Phase III Trial Data Processing",
  processing_description = "Collection of sensitive health data",
  legal_basis = "EXPLICIT_CONSENT",
  created_by = "dpo"
)

# Add risk assessment
add_risk_assessment(
  pia_id = pia$pia_id,
  risk_category = "DATA_BREACH",
  risk_description = "Unauthorized access to trial data",
  likelihood = "LOW",
  impact = "HIGH",
  mitigation_measures = "Encryption, access controls, audit logging",
  residual_risk = "LOW",
  assessed_by = "security_officer"
)
```

## 11. Breach Notification (Articles 33-34)

### Regulatory Requirement

Article 33 requires notification to supervisory authorities within 72 hours of
becoming aware of a personal data breach. Article 34 requires communication to
affected data subjects when the breach is likely to result in high risk.

### Technical Implementation

The breach notification module implements a complete incident response workflow:

- **Incident Recording**: Structured capture of breach details including
  discovery time, scope, and data categories affected
- **Timeline Tracking**: Automatic monitoring of the 72-hour deadline with
  escalation alerts
- **Risk Assessment**: Structured assessment of breach severity and subject
  impact
- **Notification Generation**: Template-based notifications for authorities
  and subjects
- **Remediation Tracking**: Documentation of containment and remediation
  actions

```r
# Example: Report and assess breach
incident <- report_breach_incident(
  breach_type = "UNAUTHORIZED_ACCESS",
  breach_description = "Unauthorized login detected",
  discovery_datetime = Sys.time(),
  estimated_subjects = 50,
  data_categories = "HEALTH,IDENTIFICATION",
  reported_by = "security_team"
)

# Check 72-hour deadline
deadline <- check_72_hour_deadline(incident$incident_id)
# Returns hours remaining and notification status
```

# Part II: Good Clinical Practice (GCP) Features

Good Clinical Practice guidelines, established by the International Council
for Harmonisation (ICH E6), define standards for clinical trial design,
conduct, and reporting. ZZedc implements the following GCP-aligned features.

## 12. Protocol Compliance Monitoring

### Regulatory Requirement

ICH E6(R2) Section 4.5 requires that clinical trials be conducted in accordance
with the protocol. Section 5.18 mandates monitoring to verify protocol
adherence.

### Technical Implementation

The protocol compliance module provides systematic monitoring capabilities:

- **Protocol Definition**: Structured capture of protocol elements including
  visits, procedures, and eligibility criteria
- **Visit Scheduling**: Automated scheduling with window calculations
- **Deviation Detection**: Real-time identification of protocol deviations
- **Deviation Classification**: Categorization by severity and type
- **Corrective Action Tracking**: Documentation of responses to deviations

```r
# Example: Define protocol and monitor compliance
protocol <- create_protocol(
  protocol_code = "STUDY-001",
  protocol_version = "2.0",
  study_phase = "PHASE_3",
  therapeutic_area = "ONCOLOGY",
  created_by = "medical_director"
)

# Add visit schedule
add_protocol_visit(
  protocol_id = protocol$protocol_id,
  visit_code = "SCREENING",
  visit_name = "Screening Visit",
  visit_day = -14,
  window_before = 7,
  window_after = 0,
  is_required = TRUE
)

# Record protocol deviation
create_protocol_deviation(
  protocol_id = protocol$protocol_id,
  subject_id = "SUBJ-001",
  deviation_type = "VISIT_WINDOW",
  deviation_description = "Screening visit outside window",
  deviation_date = Sys.Date(),
  reported_by = "coordinator"
)
```

### References

International Council for Harmonisation. (2016). ICH E6(R2) Guideline for Good
Clinical Practice.

## 13. Adverse Event Management

### Regulatory Requirement

ICH E6(R2) Section 4.11 requires documentation and reporting of adverse events.
Serious adverse events require expedited reporting to sponsors and regulatory
authorities.

### Technical Implementation

The adverse event module provides comprehensive AE/SAE management:

- **Event Capture**: Structured recording of adverse events with MedDRA coding
  support
- **Severity Grading**: CTCAE-aligned severity grading (Grade 1-5)
- **Causality Assessment**: Systematic causality evaluation workflow
- **SAE Escalation**: Automatic identification of serious adverse event
  criteria
- **Expedited Reporting**: Timeline tracking for regulatory reporting
  requirements (24-hour, 7-day, 15-day)
- **Follow-up Tracking**: Documentation of event evolution and resolution

```r
# Example: Record adverse event with SAE upgrade
ae <- create_adverse_event(
  subject_id = "SUBJ-001",
  ae_term = "Headache",
  ae_description = "Moderate headache after dose",
  onset_date = Sys.Date(),
  severity_grade = "GRADE_2",
  reported_by = "investigator"
)

# Upgrade to SAE if hospitalization required
upgrade_to_sae(
  ae_id = ae$ae_id,
  sae_criteria = "HOSPITALIZATION",
  upgraded_by = "investigator"
)
```

## 14. CRF Completion Guidelines

### Regulatory Requirement

ICH E6(R2) Section 4.9 requires that data be recorded accurately and in a
manner that allows verification. CRF completion guidelines support consistent
data collection.

### Technical Implementation

The CCG module generates comprehensive field-level completion instructions:

- **Form Documentation**: Automatic generation of form completion guides
- **Field Instructions**: Detailed guidance for each data field including
  valid values, units, and collection procedures
- **Version Control**: CCG versioning aligned with CRF versions
- **Approval Workflow**: Review and approval process for CCG documents

## 15. CRF Version Control

### Regulatory Requirement

ICH E6(R2) Section 5.5.3 requires documentation of CRF modifications. FDA
21 CFR 312.62 requires maintenance of accurate case histories.

### Technical Implementation

The version control module provides complete CRF lifecycle management:

- **Version History**: Immutable record of all CRF versions
- **Change Documentation**: Detailed change logs with rationale
- **Comparison Tools**: Version-to-version comparison capabilities
- **Deployment Tracking**: Record of version deployment to sites

## 16. CRF Design Review

### Regulatory Requirement

ICH E6(R2) Section 5.5.3 specifies sponsor responsibilities for CRF design.
Industry best practices recommend formal design review processes.

### Technical Implementation

The review module implements structured CRF design review:

- **Review Cycles**: Multi-stage review with configurable reviewers
- **Comment Tracking**: Structured capture and resolution of review comments
- **Approval Workflow**: Formal sign-off requirements by stage
- **Design History**: Complete record of review activities

# Part III: FDA 21 CFR Part 11 Features

FDA 21 CFR Part 11 establishes requirements for electronic records and
electronic signatures. ZZedc implements the following Part 11 features.

## 17. Electronic Signatures

### Regulatory Requirement

21 CFR Part 11 Subpart C establishes requirements for electronic signatures
including unique identification, signature manifestation, and signature/record
linking.

### Technical Implementation

The electronic signature module implements compliant e-signatures:

- **Signature Components**: Username plus password combination as required by
  11.200(a)
- **Signature Meaning**: Capture of signature meaning (authorship, review,
  approval) as required by 11.50
- **Signature Manifestation**: Display of signer name, date/time, and meaning
  as required by 11.50
- **Record Linking**: Cryptographic binding of signature to record content
- **Non-Repudiation**: Hash-based verification preventing signature forgery

```r
# Example: Apply electronic signature
signature <- apply_electronic_signature(
  record_type = "CRF",
  record_id = "CRF-001-V1",
  signer_id = "dr_smith",
  signer_password = "****",
  signature_meaning = "APPROVAL",
  signature_reason = "Approving completed case report form"
)

# Verify signature integrity
verify_electronic_signature(
  signature_id = signature$signature_id
)
```

### References

U.S. Food and Drug Administration. (2003). 21 CFR Part 11: Electronic Records;
Electronic Signatures.

## 18. Audit Trail System

### Regulatory Requirement

21 CFR 11.10(e) requires computer-generated, time-stamped audit trails that
independently record the date and time of operator entries and actions that
create, modify, or delete electronic records.

### Technical Implementation

The audit trail module provides comprehensive activity logging:

- **Automatic Capture**: All record operations automatically logged without
  operator intervention
- **Immutability**: Hash-chained audit records preventing undetected tampering
- **Timestamp Accuracy**: System-generated timestamps independent of operator
  input
- **Record Linkage**: Direct association between audit entries and affected
  records
- **Retention**: Audit trail retention for the lifetime of the associated
  record plus regulatory retention period

```r
# Example: Audit trail verification
# All operations automatically logged
create_correction_request(
  table_name = "subjects",
  record_id = "SUBJ-001",
  field_name = "birth_date",
  current_value = "1990-01-01",
  proposed_value = "1990-01-02",
  correction_reason = "Transcription error",
  requested_by = "coordinator"
)

# Verify audit trail integrity
verify_correction_integrity(request_id)
```

## 19. Data Correction Workflow

### Regulatory Requirement

21 CFR 11.10(e) requires that changes to records not obscure previously
recorded information. Original entries must remain visible with the date/time
and identity of the person making the change.

### Technical Implementation

The data correction module implements controlled change management:

- **Correction Requests**: Formal request process for data changes
- **Reason Documentation**: Mandatory capture of correction rationale
- **Dual Control**: Configurable approval requirements for corrections
- **Original Preservation**: Original values retained with full history
- **Override Procedures**: Documented escalation for urgent corrections

## 20. System Validation Framework

### Regulatory Requirement

21 CFR 11.10(a) requires that systems be validated to ensure accuracy,
reliability, consistent intended performance, and the ability to discern
invalid or altered records.

### Technical Implementation

The validation framework implements a complete IQ/OQ/PQ approach:

- **Installation Qualification (IQ)**: Verification of correct installation
  and configuration
- **Operational Qualification (OQ)**: Testing of system functions under
  normal conditions
- **Performance Qualification (PQ)**: Verification under actual use
  conditions
- **Validation Documentation**: Automated generation of validation reports

```r
# Example: Run validation suite
validation <- run_validation_suite(
  validation_type = "FULL",
  environment = "PRODUCTION"
)

# Generate validation report
generate_validation_report(
  validation_id = validation$validation_id,
  format = "PDF"
)
```

## 21. Change Control System

### Regulatory Requirement

21 CFR 11.10(k)(2) requires controls over the revision and change of system
documentation and operational systems.

### Technical Implementation

The change control module manages system modifications:

- **Change Requests**: Formal request process with categorization
- **Impact Assessment**: Systematic evaluation of change impacts
- **Approval Workflow**: Multi-level approval for significant changes
- **Implementation Tracking**: Documentation of change implementation
- **Validation Integration**: Triggering of revalidation as appropriate

## 22. Access Controls

### Regulatory Requirement

21 CFR 11.10(d) requires limiting system access to authorized individuals.
11.10(g) requires authority checks to ensure authorized operations.

### Technical Implementation

The access control system implements role-based security:

- **Role Definitions**: Configurable roles with specific permissions
- **Function-Level Control**: Permissions at the operation level
- **Session Management**: Secure session handling with timeout
- **Access Logging**: Complete record of access attempts

# Part IV: Integrated Compliance Features

The following features support compliance across multiple regulatory frameworks.

## 23. Protocol-CRF Linkage

### Regulatory Requirement

ICH E6(R2) requires traceability between protocol requirements and data
collection. FDA expects clear documentation of data collection rationale.

### Technical Implementation

The linkage module connects protocol elements to CRF fields:

- **Objective Mapping**: Links between protocol objectives and data endpoints
- **Visit-Form Mapping**: Association of forms with protocol visits
- **Traceability Matrix**: Automated generation of requirement traceability

## 24. Study Closeout Management

### Regulatory Requirement

ICH E6(R2) Section 4.13 and FDA regulations require systematic study closeout
with complete documentation.

### Technical Implementation

The closeout module provides structured study completion:

- **Closeout Checklist**: Comprehensive checklist of required activities
- **Data Reconciliation**: Systematic data verification
- **Database Lock**: Controlled database locking procedures
- **Archive Preparation**: Documentation of archive contents

## 25. Master Field Library

### Regulatory Requirement

CDISC standards promote standardized data collection. Regulatory submissions
benefit from consistent variable definitions.

### Technical Implementation

The field library provides standardized field definitions:

- **CDISC Alignment**: Pre-defined fields aligned with SDTM domains
- **Controlled Terminology**: Integration with CDISC controlled terminology
- **Reusability**: Centralized definitions for consistency across studies

## 26. Conditional Logic System

### Regulatory Requirement

GCP requires relevant data collection. Conditional logic ensures collection of
applicable data while reducing participant burden.

### Technical Implementation

The conditional logic module enables dynamic form behavior:

- **Show/Hide Rules**: Field visibility based on prior responses
- **Enable/Disable Rules**: Field editability controls
- **Validation Rules**: Conditional validation requirements
- **Skip Logic**: Structured navigation based on responses

## 27. Calculated Fields

### Regulatory Requirement

FDA and ICH require accurate derived data. Standard calculations reduce
transcription errors and ensure consistency.

### Technical Implementation

The calculation module supports standard clinical calculations:

- **BMI Calculation**: Automatic BMI with category assignment
- **Age Calculation**: Precise age from dates
- **BSA Formulas**: Multiple body surface area formulas (DuBois, Mosteller)
- **Score Summation**: Composite score calculation with missing data handling

## 28. CRF Designer

### Regulatory Requirement

Effective CRF design supports data quality and regulatory compliance. Visual
design tools improve form usability.

### Technical Implementation

The designer module provides form creation capabilities:

- **Visual Layout**: Grid-based field positioning
- **Section Management**: Logical grouping of related fields
- **Field Properties**: Comprehensive field configuration
- **Preview Generation**: Form preview before deployment

# Conclusion

ZZedc provides a comprehensive regulatory compliance framework addressing the
requirements of GDPR, GCP, and FDA 21 CFR Part 11. The system's modular
architecture enables organizations to implement compliance features appropriate
to their specific regulatory context while maintaining flexibility for diverse
clinical research applications.

The implementation emphasizes automation of compliance activities, reducing
manual burden while improving consistency and reliability. Comprehensive audit
trails support regulatory inspections and demonstrate organizational commitment
to data protection and research integrity.

# References

European Parliament and Council of the European Union. (2016). Regulation (EU)
2016/679 of the European Parliament and of the Council of 27 April 2016 on the
protection of natural persons with regard to the processing of personal data
and on the free movement of such data (General Data Protection Regulation).
Official Journal of the European Union, L119, 1-88.

International Council for Harmonisation of Technical Requirements for
Pharmaceuticals for Human Use. (2016). Integrated Addendum to ICH E6(R1):
Guideline for Good Clinical Practice E6(R2).

U.S. Food and Drug Administration. (2003). 21 CFR Part 11: Electronic Records;
Electronic Signatures. Code of Federal Regulations, Title 21.

U.S. Food and Drug Administration. (2017). Part 11, Electronic Records;
Electronic Signatures - Scope and Application: Guidance for Industry.

Article 29 Data Protection Working Party. (2017). Guidelines on Data Protection
Impact Assessment (DPIA) (wp248rev.01).

Article 29 Data Protection Working Party. (2018). Guidelines on Personal data
breach notification under Regulation 2016/679 (wp250rev.01).

Clinical Data Interchange Standards Consortium. (2021). CDISC Standards.
https://www.cdisc.org/standards

European Medicines Agency. (2010). Reflection paper on expectations for
electronic source data and data transcribed to electronic data collection
tools in clinical trials. EMA/INS/GCP/454280/2010.

```{r session-info, eval=FALSE}
sessionInfo()
```
