---
title: "Technical-Lead Guide: Deploying ZZedc for a Clinical Trial"
subtitle: "End-to-end runbook for the statistician/IT-adjacent team member"
author: "ZZedc Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical-Lead Guide for Deployment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# About this guide

This is the canonical guide for the **technical lead** of a
clinical trial running ZZedc: the statistician, biomedical
informaticist, or IT-adjacent colleague who is responsible for
installation, server configuration, encryption, backups, and
day-2 operations. You work alongside a **content author** (a
study coordinator, project manager, or PI's research assistant)
who provides the case-report forms, validation rules, and user
roster. That person follows
[Content-Author Guide](content-author-guide.html).

The guide is written for a first small-to-medium clinical
trial (10 to 200 participants, 1 to 5 sites). It assumes you
have moderate R familiarity (you can `install.packages()`,
read a function signature, debug a stack trace) and basic
sysadmin skills (you can SSH to a Linux VM, edit a config
file, manage TLS certificates). If this describes your
intended role exactly, this guide will take you from a fresh
CRAN install to a production-ready deployment.

By the end of this guide:

- a server is provisioned with the chosen database backend
- ZZedc is installed, configured, and reachable over HTTPS
- the content author's workbook has been imported
- multi-site role-based access works
- the audit log validates end-to-end
- daily backups run automatically and a restore drill has
  succeeded
- the trial is ready for go-live

This document absorbs three previously separate vignettes
(disaster recovery, encryption at rest, AWS quick-start). The
material is now folded into Chapters 5, 8, 9, and Appendix C
respectively. There are no other vignette dependencies for
deployment; the per-backend quickstarts remain as separate
references.

---

# Chapter 1. Architecture overview

ZZedc separates concerns into five layers; each is documented
in its own reference vignette, but you should understand the
whole picture before deploying.

## The five-layer model

```
+------------------------------------------------------------+
| 1. Shiny application (UI + reactive logic)                |
+------------------------------------------------------------+
| 2. Module layer (auth, EDC, reports, exports, compliance) |
+------------------------------------------------------------+
| 3. R6 DatabaseAdapter (dialect-aware queries, pooling)    |
+------------------------------------------------------------+
| 4. SQLCipher / DBI driver (encryption + transport)        |
+------------------------------------------------------------+
| 5. Storage backend                                        |
|    SQLite | DuckDB | PostgreSQL | MySQL | ClickHouse      |
+------------------------------------------------------------+
```

The application interacts only with the adapter; backend
choice is configured per-project in `config.yml` and is
otherwise transparent.

## Three foundational guarantees

ZZedc commits to three things and the rest of this guide is
about preserving them:

1. **Encryption at rest.** Subject data is stored on disk
   under AES-256 when the package is run against a
   SQLCipher-enabled RSQLite (for example via
   SQLite3MultipleCiphers or a custom build), or against a
   server-side encrypted backend (TDE for
   PostgreSQL/MySQL/ClickHouse). The default stock RSQLite
   binary on CRAN does not include SQLCipher, so encryption
   on SQLite-backed deployments is opt-in and depends on the
   driver build present on the production host. Verify on
   the deployment host with `verify_database_encryption()`
   (Chapter 5).
2. **Hash-chained audit log.** Every database mutation appends
   a row to `audit_log` whose `audit_hash` depends on the
   previous row's hash. Tampering with any row breaks the
   chain. `verify_audit_log_integrity()` validates the chain
   end-to-end. The chain is correct by construction; an
   integrity claim against a particular live database
   requires running the verifier against that database.
   This is what the trial's 21 CFR Part 11
   continuity-of-record argument rests on.
3. **No side channels.** The application never writes subject
   data to disk except through the adapter; never to
   stdout/stderr, never to log files, never to caches. If you
   see subject data on disk outside the encrypted database,
   something is wrong.

The audit chain in particular is the differentiator vs. less
formal capture systems. Treat it as the system's most
load-bearing invariant. Any operation that bypasses the
adapter is a bug.

## Multi-site model

ZZedc treats *site* as a tag on every user, every subject, and
every audit-log entry. The `site_id` column is filterable in
every query. Coordinator-class roles are scoped to their site;
PI-class roles can be scoped to a site or to all sites; the
data-manager and central monitor roles are scoped to all sites
by convention.

This model is **logical, not physical**: there is one server,
one database, and the application enforces site-based access
at the adapter layer. There is no federated multi-database
mode and no cross-server replication for site isolation.
Trials whose threat model requires physical isolation per site
should run a separate ZZedc instance per site.

---

# Chapter 2. Picking the right backend

ZZedc supports five backends. For a first small-to-medium
trial, only two are likely candidates: **SQLite** (if you
expect under ~50 concurrent users and the database fits on
one disk) or **PostgreSQL** (if you need more than that, or
have institutional infrastructure already). The other three
are for specific situations.

## Decision tree

```
Are you running locally on one machine or one VM?
+-- Yes: SQLite.
+-- No: continue.

Is institutional MySQL/MariaDB already provisioned and managed
by your IT team?
+-- Yes: MySQL/MariaDB.
+-- No: continue.

Are you migrating from REDCap?
+-- Yes: MySQL/MariaDB (matches REDCap's native backend; the
|         migration importer is fastest against this).
+-- No: continue.

Will analysts run heavy ad-hoc OLAP queries against the EDC
database directly, or do you ship to Parquet/SAS/S3 for
analysis anyway?
+-- Heavy OLAP: DuckDB.
+-- Ship for analysis: continue.

Will you have more than 1000 subjects with frequent
high-cardinality time-series data (mobile-health, sensor
streams)?
+-- Yes: ClickHouse.
+-- No: PostgreSQL.
```

For 90% of academic clinical trials, the answer is **SQLite**
for solo or small-site work and **PostgreSQL** for
institutional multi-site deployments. Per-backend setup
mechanics are in
[`vignette('backend-quickstart')`](backend-quickstart.html)
(decision tree plus an appendix per backend: A SQLite,
B DuckDB, C MySQL/MariaDB, D PostgreSQL, E ClickHouse).

The rest of this guide assumes SQLite for examples (because
it is the simplest and the path most readers will follow);
where the answer differs for PostgreSQL/MySQL we call out the
delta.

## Backend cheat sheet

| Backend | When to use | Concurrent users | Encryption | Backup model |
|---|---|---|---|---|
| SQLite + SQLCipher | Solo, small site, on-prem | up to ~50 | SQLCipher (driver-dependent) | File copy |
| DuckDB | Analytics-heavy, columnar | up to ~50 | At-rest via OS | File copy |
| PostgreSQL | Institutional, multi-site | hundreds | TDE / pgcrypto | pg_dump or WAL archive |
| MySQL/MariaDB | REDCap shops, institutional | hundreds | TDE | mysqldump or binlog |
| ClickHouse | High-volume analytics | thousands | TDE | clickhouse-backup |

---

# Chapter 3. Installing ZZedc

## 3.1 Prerequisites

- R 4.1.0 or later. R 4.4 is recommended.
- A Linux VM (Ubuntu 22.04 LTS or Debian 12 are tested), a
  macOS machine, or a Windows server.
- ~2 GB disk space (R, dependencies, encrypted database).
- Outbound HTTPS to CRAN to install R packages.
- For a server deployment: a domain name and a TLS certificate.

For a Docker-based deployment (recommended for any deployment
above one user), also: Docker 24+ and Docker Compose 2+.

## 3.2 Installing from CRAN

Once the package is on CRAN, installation is one line:

```{r}
install.packages("zzedc")
```

This pulls all required dependencies, including `shiny`,
`bslib`, `DBI`, `RSQLite`, `R6`, `digest`, `openssl`,
`jsonlite`, and `googlesheets4`.

For backends other than SQLite, install the optional driver:

```{r}
install.packages("RPostgres")   # PostgreSQL
install.packages("RMariaDB")    # MySQL / MariaDB
install.packages("duckdb")      # DuckDB
# ClickHouse: see vignette('backend-quickstart') Appendix E
```

For SQLCipher-enabled SQLite (required if you want at-rest
encryption on a SQLite-backed deployment), install a
SQLCipher-aware build of RSQLite. The stock CRAN build of
`RSQLite` does not include SQLCipher; consult the
SQLite3MultipleCiphers project or your distribution for a
suitable build.

## 3.3 Reproducibility: pin dependencies with renv

For any deployment that will outlive a single session, pin
dependencies with `renv`. This guarantees that the exact
version of every package on the production server matches
the version your tests passed against.

```{r}
install.packages("renv")
renv::init()
renv::snapshot()
```

Commit `renv.lock` and `renv/activate.R` to your trial's
configuration repository (separate from the package source).
On the production server, `renv::restore()` rebuilds the
exact library.

## 3.4 Verify the install

```{r}
library(zzedc)
packageVersion("zzedc")
```

If you see the package version without errors, the install is
working. Continue to Chapter 4 for configuration.

---

# Chapter 4. Configuring the trial

ZZedc reads configuration from two places: a YAML file
(`config.yml`) and environment variables. The YAML file holds
the per-trial settings; environment variables hold secrets.

## 4.1 The trial's `config.yml`

Create `config.yml` in your trial's working directory. A
minimal configuration for a SQLite-backed solo trial:

```yaml
default:
  study_name: "Memory Study 001"
  protocol_id: "MEMORY-001"
  pi_name: "Jane Smith, MD"
  pi_email: "jsmith@example.org"
  study_phase: "II"
  target_enrollment: 60

  db_backend: "sqlite"
  db_path: "/srv/zzedc/MEMORY-001/study.db"

  encryption_at_rest: true
  audit_logging: true
  session_timeout_minutes: 30
```

For PostgreSQL replace the `db_path` block with:

```yaml
  db_backend: "postgresql"
  db_host: "postgres.internal"
  db_port: 5432
  db_name: "zzedc_memory001"
  db_user: "zzedc_app"
  db_sslmode: "require"
```

(The password comes from an environment variable, not the
file; see §4.2.)

For MySQL/MariaDB:

```yaml
  db_backend: "mysql"
  db_host: "mariadb.internal"
  db_port: 3306
  db_name: "zzedc_memory001"
  db_user: "zzedc_app"
```

The full list of options is documented in
`?zzedc::read_config`.

## 4.2 Environment variables (secrets)

Never put secrets in `config.yml`. Use environment variables:

```sh
# /etc/zzedc/MEMORY-001.env  (chmod 600, owned by zzedc user)
DB_ENCRYPTION_KEY="..."         # SQLite/DuckDB master key
DB_PASSWORD="..."               # Postgres/MySQL/ClickHouse
ZZEDC_SALT="..."                # Random 32-byte hex string
ZZEDC_DB_PATH="/srv/zzedc/MEMORY-001/study.db"
GOOGLE_APPLICATION_CREDENTIALS="/etc/zzedc/gcs-svc.json"
```

Source this file in the launch script. For a systemd-managed
deployment, use `EnvironmentFile=`. For Docker, use
`--env-file`.

## 4.3 Encryption-key custody

For SQLite/DuckDB, the encryption key is generated at database
init. *Where you store it determines the trial's threat model*.
Three patterns:

| Pattern | Custodian | When to use |
|---|---|---|
| **Server env var** | The server's local environment | Solo work, low-risk trials |
| **AWS KMS / GCP KMS** | A cloud key-management service | Most institutional deployments |
| **HSM / Yubikey** | Hardware security module | Pharma trials, FDA-submitted studies |

For institutional deployments use AWS KMS (or your cloud's
equivalent). The `R/aws_kms_utils.R` helpers expose the
encrypt/decrypt operations the application needs; see
`?zzedc::aws_kms_encrypt`. The custody scenarios common in
practice are:

- *Pharmaceutical trial, sponsor-held key.* Key in the
  sponsor's AWS Secrets Manager; sites access via a
  sponsor-controlled interface; rotation managed by the
  sponsor's IT security team; unblinding requires sponsor
  authorisation.
- *Academic trial, DCC-held key.* Key in the Data
  Coordinating Center's institutional secret store;
  investigators receive encrypted exports; the DCC controls
  the rotation schedule.
- *Single-site trial, site-held key.* Key in local
  institutional secrets management; the PI holds ultimate
  responsibility; consider a dual-key arrangement for
  critical studies and document recovery procedures.

## 4.4 Per-trial filesystem layout

Adopt this layout per trial; it makes backups, audits, and
restores predictable:

```
/srv/zzedc/
+-- MEMORY-001/
|   +-- study.db                 # encrypted SQLite database
|   +-- config.yml               # per-trial config
|   +-- renv.lock                # pinned dependencies
|   +-- exports/                 # signed exports
|   +-- backups/                 # daily snapshots
|   +-- logs/                    # application logs (no PHI)
+-- MEMORY-002/
+-- ...
```

For PostgreSQL/MySQL deployments, `study.db` is replaced by
the database service's data directory; the rest of the
layout is identical.

---

# Chapter 5. Initialising the database

## 5.1 Encryption-at-rest architecture

Database encryption operates at the connection layer. When
the SQLite driver in use is SQLCipher-aware, all data
written to disk is automatically encrypted and queries
decrypt transparently. Application code is unchanged.
Characteristics of the SQLCipher integration:

- *Algorithm.* AES-256-CBC with HMAC-SHA256 authentication.
- *Key length.* 256-bit (64 hexadecimal characters).
- *Page size.* 4096 bytes (SQLCipher default).
- *Key derivation.* PBKDF2 with SHA256.

For PostgreSQL/MySQL/ClickHouse, encryption-at-rest is the
backend's transparent data encryption (TDE) feature; ZZedc
contributes only the application-side key custody and audit
logic.

Note on the partial nature of this guarantee for
SQLite-backed deployments: the stock RSQLite binary
distributed via CRAN is built without SQLCipher. ZZedc will
still create and use a SQLite database, but at-rest
encryption requires a SQLCipher-enabled RSQLite (for example
via SQLite3MultipleCiphers or a custom build). Run
`verify_database_encryption()` on the deployment host
(§5.5) before claiming encryption-at-rest in any DMP or IRB
submission.

## 5.2 Generate and verify the encryption key

Generate a 256-bit key:

```{r}
library(zzedc)
key <- generate_db_key()
verify_db_key(key)               # checks 64-hex-char format
```

Store this key immediately in your secrets manager or in the
trial's env file under `DB_ENCRYPTION_KEY`. If you lose this
key, the database is unrecoverable. There is no backdoor.

For production, retrieve the key from a managed service:

```{r}
key <- get_encryption_key(
  aws_kms_key_id = "zzedc/db-encryption-key"
)
```

This pattern centralises custody, supports rotation, and
captures access in CloudTrail.

## 5.3 Create the encrypted database

For SQLite/DuckDB:

```{r}
init <- initialize_encrypted_database(
  db_path   = "/srv/zzedc/MEMORY-001/study.db",
  overwrite = FALSE
)
init$success           # TRUE
init$key               # SAVE THIS; treat as a password
```

The returned `init$key` is the master encryption key. Store
it as in §5.2.

For PostgreSQL/MySQL the equivalent is to provision an empty
database and grant the application user `CREATE` and `INSERT`
on it. See the per-backend quickstart vignettes for SQL.

## 5.4 Connect to an existing encrypted database

```{r}
conn <- connect_encrypted_db(
  db_path = "/srv/zzedc/MEMORY-001/study.db"
)
res <- DBI::dbGetQuery(conn, "SELECT COUNT(*) FROM subjects")
DBI::dbDisconnect(conn)
```

`connect_encrypted_db()` retrieves the key from
`DB_ENCRYPTION_KEY` or from the configured KMS automatically.

## 5.5 Verify encryption is actually in effect

```{r}
verification <- verify_database_encryption(
  db_path = "/srv/zzedc/MEMORY-001/study.db"
)
verification$encrypted     # TRUE only if SQLCipher is active
verification$message
```

This check opens the file with a stock SQLite reader and
confirms the bytes are not interpretable as plaintext SQLite.
If `encrypted` is `FALSE`, the running RSQLite is not
SQLCipher-enabled; subject data on this host is at rest in
the clear. Stop and rebuild against a SQLCipher driver
before proceeding.

## 5.6 Migrate an unencrypted database to encrypted

If the trial accumulated data on a stock RSQLite before a
SQLCipher driver was available, convert the database in place:

```{r}
result <- migrate_to_encrypted(
  old_db_path = "/srv/zzedc/MEMORY-001/legacy.db",
  backup_dir  = "/srv/zzedc/MEMORY-001/backups"
)
result$success
result$records_migrated
result$new_path
```

The function takes a backup before migration and verifies
row-counts after. Validate independently:

```{r}
verify_migration(
  old_db_path = "/srv/zzedc/MEMORY-001/legacy.db",
  new_db_path = "/srv/zzedc/MEMORY-001/legacy_encrypted.db"
)
```

If migration fails, roll back from the backup:

```{r}
rollback_migration(
  backup_path = "/srv/zzedc/MEMORY-001/backups/legacy_*.db",
  restore_to  = "/srv/zzedc/MEMORY-001/legacy.db"
)
```

## 5.7 Initialise the audit log

```{r}
audit_res <- init_audit_logging(
  db_path = "/srv/zzedc/MEMORY-001/study.db"
)
audit_res$success
```

This creates the `audit_log`, `audit_events`, and
`audit_chain` tables. Every subsequent database write
appends to them. Valid event types written by the
application: INSERT, UPDATE, DELETE, SELECT, EXPORT, LOGIN,
LOGOUT, ACCESS.

## 5.8 Create the canonical core tables

The setup wizard normally does this. For a programmatic setup,
provision the core tables (study_info, edc_users, edc_roles,
subjects, form_data, etc.) via:

```{r}
result <- create_wizard_database(
  config_list = list(
    study_name        = "Memory Study 001",
    protocol_id       = "MEMORY-001",
    pi_name           = "Jane Smith, MD",
    pi_email          = "jsmith@example.org",
    study_phase       = "II",
    target_enrollment = 60,
    admin_username    = "admin",
    admin_password    = "TempPass!2026-CHANGEME",
    admin_fullname    = "Trial Admin",
    admin_email       = "admin@example.org",
    security_salt     = Sys.getenv("ZZEDC_SALT")
  ),
  db_path = "/srv/zzedc/MEMORY-001/study.db",
  backend = "sqlite"
)
```

This produces an admin user (whose password the admin must
change on first login) and the empty `study_info` /
`edc_roles` rows the application expects.

## 5.9 Encryption-key rotation

Rotate the encryption key once a year, or sooner if you
suspect compromise. Plan a 30-minute maintenance window;
the trial is unavailable during rotation. The procedure is:

1. Stop the application (`docker compose stop zzedc`).
2. Take a fresh backup (Chapter 9).
3. Generate a new key with `generate_db_key()`.
4. Use SQLCipher's `PRAGMA rekey = 'new-key';` (via a
   privileged R session) to re-encrypt the database in
   place.
5. Update `DB_ENCRYPTION_KEY` in the secrets store.
6. Restart and run `verify_audit_log_integrity()` (Chapter
   10) before re-opening to users.

Document the rotation in the trial's DMP. Keep the prior
key in escrow long enough to decrypt any backups taken
before rotation.

## 5.10 Performance and security notes

Encryption overhead in practice: connection establishment
adds <10ms; query execution adds <5% for typical clinical
data volumes; export operations scale linearly with data
volume. The overhead is invisible to end users.

Operational hygiene:

- Never store keys in source code, in `config.yml`, or in
  Git history. Use environment variables in development
  and a managed secrets service in production.
- Restrict database file access to the application user;
  on Linux, `chmod 640` and a dedicated `zzedc` group.
- Encrypt backups at rest (the daily-backup recipe in
  Chapter 9 produces gzip + SHA-256-verified files; for
  off-site mirrors use S3 server-side encryption or
  equivalent).
- Encrypt backup transfers in transit (TLS 1.2+).
- Conduct a periodic security review of who holds the key
  and who can read the secrets store.

---

# Chapter 6. Loading the content author's workbook

The content author follows
[Content-Author Guide](content-author-guide.html) to produce
a Google Sheets workbook with three tabs: `Study_Users`,
`Data_Dictionary`, `validation_rules`. They share the workbook
URL with you. Your job is one R command.

## 6.1 Authentication to Google Sheets

Authenticate `googlesheets4` once per server:

```{r}
library(googlesheets4)
gs4_auth(email = "svc-zzedc@example.iam.gserviceaccount.com")
```

Use a service account, not a personal Google account, so the
import does not depend on any one human's credentials. The
service-account JSON goes into
`GOOGLE_APPLICATION_CREDENTIALS` (§4.2).

## 6.2 Run the import

```{r}
library(zzedc)
result <- setup_zzedc_from_gsheets(
  sheet_id = "https://docs.google.com/spreadsheets/d/SHEET_ID/edit",
  db_path  = "/srv/zzedc/MEMORY-001/study.db",
  imported_by = "tech_lead"
)

result$success                 # TRUE
result$users$imported          # number of users created
result$forms$imported_forms    # number of CRFs created
result$forms$imported_fields   # number of fields across CRFs
result$validation_rules$imported  # number of rules
length(result$skipped_rules)   # rules that need manual review
```

The import is idempotent: re-running it after the content
author updates the workbook re-applies the deltas. The audit
log captures every change.

## 6.3 Resolve skipped rules

If `length(result$skipped_rules) > 0`, the importer found a
validation rule it could not parse. Print the list:

```{r}
str(result$skipped_rules)
```

Each entry is a named list with `field`, `reason`, and the
original `rule_dsl`. Send this list to the content author with
a note: rewrite each rule using one of the six patterns in
[Content-Author Guide](content-author-guide.html) §3.6, or
flag it for manual implementation by you.

## 6.4 CSV authoring path (alternative to Google Sheets)

For sites that block Google Drive, the content author can
deliver three CSV files (`Study_Users.csv`,
`Data_Dictionary.csv`, `validation_rules.csv`) instead. Use:

```{r}
result <- setup_zzedc_from_csv(
  csv_dir = "/srv/zzedc/MEMORY-001/incoming/",
  db_path = "/srv/zzedc/MEMORY-001/study.db",
  imported_by = "tech_lead"
)
```

The contract is otherwise identical. See
[`vignette('demonstration-trial-setup')`](demonstration-trial-setup.html)
for a full worked example.

---

# Chapter 7. Multi-site setup

Multi-site setup is enforced at the database adapter layer
through the `site_id` column on `edc_users`, `subjects`,
and `form_data`. The content author has already populated
`site_id` on the user roster (Chapter 6 of their guide).
Your job is to confirm:

1. Every site appears in the application's *Admin -> Sites*
   view.
2. Coordinator-class users see only their site.
3. PI-class users with `site_id = "ALL"` see all sites.

## 7.1 Verifying site-based access

The cleanest verification is to log in as a coordinator at
each site and confirm they cannot list subjects from another
site. Do this with synthetic subjects (the content author
will create one per site during the dry run; see their
Chapter 4).

You can also verify directly against the database:

```{r}
library(DBI)
conn <- connect_encrypted_db(
  db_path = "/srv/zzedc/MEMORY-001/study.db",
  key     = Sys.getenv("DB_ENCRYPTION_KEY")
)
on.exit(DBI::dbDisconnect(conn), add = TRUE)

# How many sites are configured
DBI::dbGetQuery(conn, "
  SELECT site_id, COUNT(*) AS n_users
    FROM edc_users
   GROUP BY site_id
")

# Per-site user list
DBI::dbGetQuery(conn, "
  SELECT site_id, username, role
    FROM edc_users
   ORDER BY site_id, username
")
```

## 7.2 Adding a site after go-live

To add a site after the trial is live, ask the content author
to append rows to the `Study_Users` workbook for the new
site, then re-run `setup_zzedc_from_gsheets()`. The importer
treats new rows as additions and leaves existing rows alone
unless their data has changed.

## 7.3 Renaming a site

**Don't.** A `site_id` is referenced by every subject, every
form-data row, and every audit-log entry for that site.
Renaming would either break those references or require a
data-migration step that mutates the audit log (which would
break the hash chain). If you must, treat it as a new site
(append) and freeze the old site (set `active = FALSE`).

---

# Chapter 8. Deploying the application

## 8.1 Quick local launch (development only)

For a quick test on your laptop:

```{r}
library(zzedc)
launch_zzedc()
# Open http://localhost:3838
```

This is appropriate for the content author's dry-run protocol
(their Chapter 4) but not for any real-site use. For
production, use one of the patterns below.

## 8.2 Production pattern: Docker on a single VM

Recommended for any multi-site deployment. The Dockerfile is
included in the package; here is a minimal launch flow.

### Build the image

```sh
docker build -t zzedc:1.0.0 .
```

### Compose file

```yaml
# /srv/zzedc/MEMORY-001/docker-compose.yml
services:
  zzedc:
    image: zzedc:1.0.0
    container_name: zzedc-memory001
    restart: unless-stopped
    env_file: /etc/zzedc/MEMORY-001.env
    volumes:
      - /srv/zzedc/MEMORY-001:/srv/zzedc/MEMORY-001
    ports:
      - "127.0.0.1:3838:3838"

  caddy:
    image: caddy:2
    container_name: zzedc-tls
    restart: unless-stopped
    ports:
      - "80:80"
      - "443:443"
    volumes:
      - /etc/zzedc/Caddyfile:/etc/caddy/Caddyfile
      - caddy_data:/data
      - caddy_config:/config

volumes:
  caddy_data:
  caddy_config:
```

### Caddyfile (TLS termination)

```
edc.example.org {
    reverse_proxy 127.0.0.1:3838
    encode gzip
    log {
        output file /var/log/caddy/edc.access.log
    }
}
```

### Bring it up

```sh
cd /srv/zzedc/MEMORY-001
docker compose up -d
docker compose ps
```

Caddy obtains a TLS certificate automatically via Let's
Encrypt. Confirm `https://edc.example.org` returns the ZZedc
login page.

## 8.3 AWS deployment pattern (config-file-driven)

For an AWS-hosted deployment, the recommended pattern is a
config-file-driven, non-interactive initialisation: the
deployment is declarative, reproducible, and friendly to
CloudFormation, ECS task definitions, or Kubernetes manifests.
Time estimate after infrastructure provisioning: 5 to 10
minutes.

### Prerequisites

- An AWS EC2 instance (or ECS/EKS task host) with R 4.4+ and
  ZZedc installed (§3.2, §3.3).
- SSH access to the instance.
- Optional: Docker / container orchestration available.

### Step 1: write `zzedc_config.yml`

```yaml
study:
  name: "Multi-Site Stroke Trial"
  protocol_id: "STROKE-2024-100"
  pi_name: "Dr. James Wilson"
  pi_email: "jwilson@ucsd.edu"
  target_enrollment: 500
  phase: "Phase3"

admin:
  username: "trial_admin"
  fullname: "Trial Administrator"
  email: "admin@trial-site.org"
  password: "SecureAWSPassword123!@#"

security:
  session_timeout_minutes: 30
  enforce_https: true
  max_failed_login_attempts: 3

compliance:
  gdpr_enabled: true
  cfr_part11_enabled: true
  audit_logging_enabled: true

database:
  backend: "sqlite"        # or duckdb / postgresql / clickhouse
  sqlite:
    path: "data/zzedc.db"
  pool_size: 5
```

The five sections (`study`, `admin`, `security`,
`compliance`, `database`) are each documented in
`?zzedc::init`. For PostgreSQL replace the `sqlite` block
with a `postgresql:` block carrying `host`, `port`,
`database`, `user`, and a `${DB_PASSWORD}` reference.

Backend selection guide for AWS:

| Deployment context | Recommended backend |
|---|---|
| Single EC2, dev | SQLite |
| Single EC2, production | DuckDB or SQLite |
| AWS RDS | PostgreSQL |
| ECS / EKS, multi-node | PostgreSQL |
| High-volume analytics | DuckDB or ClickHouse |

### Step 2: run the initialiser

```sh
cd /srv/zzedc

Rscript -e "
  library(zzedc)
  zzedc::init(mode = 'config', config_file = 'zzedc_config.yml')
"
```

`zzedc::init()` creates the encrypted database, initialises
the audit log, creates the admin user, and writes the
generated `ZZEDC_SALT` to the local `.env`. Capture that
value and move it into your secrets manager.

### Step 3: set the salt and launch

```sh
export ZZEDC_SALT="<generated-salt-from-init-output>"
nohup Rscript launch_app.R > zzedc.log 2>&1 &
```

For container-based deployment, the equivalent
`docker-compose.yml` with an optional PostgreSQL sidecar:

```yaml
services:
  zzedc:
    build: .
    ports:
      - "3838:3838"
    volumes:
      - ./data:/app/data
      - ./backups:/app/backups
      - ./logs:/app/logs
    environment:
      ZZEDC_SALT: ${ZZEDC_SALT}
      DB_BACKEND: sqlite
    restart: always
    networks: [zzedc-network]

  postgres:           # optional
    image: postgres:15
    environment:
      POSTGRES_DB: zzedc
      POSTGRES_USER: zzedc_app
      POSTGRES_PASSWORD: ${DB_PASSWORD}
    volumes:
      - postgres_data:/var/lib/postgresql/data
    networks: [zzedc-network]

networks:
  zzedc-network:
    driver: bridge

volumes:
  postgres_data:
```

The container entrypoint should run `zzedc::init(mode =
'config', ...)` once on first start, then `exec` into
`launch_app.R`.

### CloudFormation, Kubernetes, and TLS

For full Infrastructure-as-Code recipes (CloudFormation
template provisioning the EC2 instance and security group;
Kubernetes Deployment + Service + ConfigMap + PVC; nginx as
a reverse proxy or AWS Certificate Manager + Application
Load Balancer for TLS), see the deployment manifests under
`inst/deployment/`. The architecture in this guide (Docker
on a VM) is appropriate up to several hundred concurrent
users; AWS multi-AZ becomes worth the operational overhead
beyond that or for trials with FDA submission paths that
require documented redundancy.

### AWS-specific operational notes

- Use IAM roles for the application's AWS access; do not
  embed access keys.
- Use AWS Secrets Manager (via `get_encryption_key()`) for
  `DB_ENCRYPTION_KEY`, `DB_PASSWORD`, and `ZZEDC_SALT`.
- Restrict the security group to the IPs that need port 22
  (admin) and the load-balancer subnet for port 3838.
- Forward CloudWatch logs by setting
  `logging.format: json` in `zzedc_config.yml` and
  configuring the CloudWatch agent on the host.
- Use AWS Certificate Manager rather than self-signed
  certificates; the Application Load Balancer handles
  termination.

## 8.4 Other patterns

- **Posit Connect / shinyapps.io.** Possible for dev/test, but
  inappropriate for production storage of subject data unless
  you have a BAA with the host (shinyapps.io does not offer
  one at the time of writing).
- **Kubernetes.** Apply the manifests under
  `inst/deployment/k8s/` after building and pushing the
  container image. Overkill for a small trial; appropriate
  for an organisation running many trials.
- **On-prem bare metal.** Same Docker/Compose pattern, no
  cloud. Adjust backups (Chapter 9) to write to local NAS
  instead of S3.

---

# Chapter 9. Backups and disaster recovery

## 9.1 Recovery objectives

Your trial's protocol or DMP should specify two numbers:

- **RPO** (Recovery Point Objective): how much data loss is
  acceptable.
- **RTO** (Recovery Time Objective): how long to restore
  service.

Indicative targets by environment class:

| Environment | RPO | RTO | Strategy |
|---|---|---|---|
| Development | 24 h | 24-48 h | Manual restore |
| Production, non-regulated | 1-4 h | 1-4 h | Automated with verification |
| Production, regulated/clinical | 15-60 min | 15-60 min | Automated failover + hot standby |

A daily SQLite backup with on-site retention satisfies a
24-hour RPO and a 4-hour RTO for a SQLite-backed trial.
PostgreSQL with continuous WAL archiving can reach a
5-minute RPO. Most academic trials commit to a 24-hour RPO
and a 4-hour RTO; document whichever you actually achieve,
not the aspirational target.

## 9.2 Daily backup recipes

### SQLite (online backup API)

The SQLite online-backup API produces a consistent snapshot
without requiring application downtime:

```sh
#!/bin/bash
# /srv/zzedc/MEMORY-001/bin/daily-backup.sh
set -euo pipefail

TRIAL=MEMORY-001
SRC=/srv/zzedc/${TRIAL}/study.db
DST=/srv/zzedc/${TRIAL}/backups/study-$(date +%F).db

sqlite3 "$SRC" ".backup '$DST'"
gzip "$DST"
sha256sum "${DST}.gz" > "${DST}.gz.sha256"

# Retention: keep 30 daily, 12 monthly
find /srv/zzedc/${TRIAL}/backups/ -name 'study-*.db.gz' \
    -mtime +30 -delete

# Off-site mirror
aws s3 cp "${DST}.gz" "s3://zzedc-backups/${TRIAL}/daily/"
```

Schedule via cron at 02:00:

```
0 2 * * * /srv/zzedc/MEMORY-001/bin/daily-backup.sh \
    >> /var/log/zzedc/backup.log 2>&1
```

### DuckDB

Use Parquet export plus a SQL dump for cross-platform
portability:

```sh
duckdb "$DB_PATH" "EXPORT DATABASE '$PARQUET_DIR' (FORMAT PARQUET);"
tar -czf "$BACKUP_DIR/zzedc_parquet_$(date +%F).tar.gz" \
    -C "$BACKUP_DIR" parquet
duckdb "$DB_PATH" ".dump" | gzip \
    > "$BACKUP_DIR/zzedc_$(date +%F).sql.gz"
```

### PostgreSQL

Combine `pg_basebackup` for daily base backups with
continuous WAL archiving:

```sh
# In postgresql.conf:
# archive_mode = on
# archive_command = 'cp %p /var/backups/zzedc/postgresql/wal/%f'

pg_basebackup \
    -h "$DB_HOST" \
    -U "$DB_USER" \
    -D "$BACKUP_DIR/base_$(date +%F)" \
    -Ft -z -P

pg_dump \
    -h "$DB_HOST" \
    -U "$DB_USER" \
    -Fc \
    -f "$BACKUP_DIR/zzedc_$(date +%F).dump" \
    "$DB_NAME"
```

WAL archiving is the mechanism that brings PostgreSQL's RPO
below an hour.

## 9.3 Backup verification

An untested backup is not a backup. Verify daily via a
shell script and weekly via a full restore drill.

### Daily file-integrity check

```sh
#!/bin/bash
set -e
BACKUP_DIR="/srv/zzedc/MEMORY-001/backups"
LOG="/var/log/zzedc/backup_verify.log"

LATEST=$(ls -t "$BACKUP_DIR"/*.db.gz 2>/dev/null | head -1)
[ -n "$LATEST" ] || { echo "$(date) ERROR: no backups" >> "$LOG"; exit 1; }
gzip -t "$LATEST" || { echo "$(date) ERROR: corrupt $LATEST" >> "$LOG"; exit 1; }

AGE_H=$(( ($(date +%s) - $(stat -f%m "$LATEST")) / 3600 ))
[ "$AGE_H" -le 25 ] || echo "$(date) WARN: latest backup ${AGE_H}h old" >> "$LOG"

echo "$(date) OK $LATEST" >> "$LOG"
```

### Programmatic R-side verification

```{r}
verify_backup <- function(backup_path, config) {
  temp_dir <- tempfile("backup_verify_")
  dir.create(temp_dir)
  on.exit(unlink(temp_dir, recursive = TRUE))

  if (grepl("\\.gz$", backup_path)) {
    out <- file.path(temp_dir, "backup.sql")
    system2("gunzip", c("-c", backup_path), stdout = out)
    backup_path <- out
  }

  test_config <- config
  test_config$sqlite$path <- file.path(temp_dir, "verify.db")
  adapter <- create_db_adapter(test_config)
  conn <- adapter$connect()
  on.exit(adapter$disconnect(conn), add = TRUE)

  tryCatch({
    sql <- readLines(backup_path)
    for (stmt in split_sql_statements(sql)) {
      DBI::dbExecute(conn, stmt)
    }
    for (tbl in c("subjects", "forms", "form_data", "audit_log")) {
      n <- DBI::dbGetQuery(
        conn, sprintf("SELECT COUNT(*) AS n FROM %s", tbl)
      )$n
      message(sprintf("Table %s: %d rows", tbl, n))
    }
    list(success = TRUE, message = "Backup verification passed")
  }, error = function(e) {
    list(success = FALSE, message = e$message)
  })
}
```

### Backup health monitor

A second R helper, suitable for cron, raises an alert when
the most recent backup is older than expected, suspiciously
small, or the backup volume is filling up:

```{r}
check_backup_health <- function(backup_dir, max_age_hours = 25) {
  issues <- character()
  files <- list.files(
    backup_dir,
    pattern = "\\.(sql\\.gz|db\\.gz|dump)$",
    full.names = TRUE
  )
  if (length(files) == 0) {
    issues <- c(issues, "No backup files found")
  } else {
    latest <- files[which.max(file.mtime(files))]
    age_h <- as.numeric(
      difftime(Sys.time(), file.mtime(latest), units = "hours")
    )
    if (age_h > max_age_hours) {
      issues <- c(issues, sprintf("Latest backup %.1fh old", age_h))
    }
    if (file.size(latest) < 1000) {
      issues <- c(issues, "Latest backup suspiciously small")
    }
  }
  list(
    healthy    = length(issues) == 0,
    issues     = issues,
    checked_at = Sys.time()
  )
}
```

## 9.4 Quarterly restore drill

```sh
# Stop the application
docker compose -f /srv/zzedc/MEMORY-001/docker-compose.yml stop zzedc

# Restore the latest backup to a scratch path
TRIAL=MEMORY-001
LATEST=$(ls -1t /srv/zzedc/${TRIAL}/backups/study-*.db.gz | head -1)
gunzip -c "$LATEST" > /tmp/restored-${TRIAL}.db

# Verify the restored DB opens and the audit chain validates
Rscript -e '
  library(zzedc)
  res <- verify_audit_log_integrity(
    db_path = "/tmp/restored-MEMORY-001.db",
    key     = Sys.getenv("DB_ENCRYPTION_KEY")
  )
  stopifnot(isTRUE(res$valid))
'

docker compose -f /srv/zzedc/MEMORY-001/docker-compose.yml start zzedc
rm /tmp/restored-MEMORY-001.db
```

Document each drill in the trial's DMP (drill date, RTO
achieved, audit-chain result, any anomalies).

## 9.5 Recovery scenarios

The four scenarios below cover the recovery situations a
small clinical trial actually encounters. Each scenario
assumes you have a recent valid backup; if not, the immediate
priority is to preserve any partial state and call the
project maintainer.

### Scenario 1: SQLite database corruption

Symptoms: application errors, *database disk image is
malformed*, unexpected query results.

```sh
#!/bin/bash
CORRUPT=/srv/zzedc/MEMORY-001/study.db
BACKUP_DIR=/srv/zzedc/MEMORY-001/backups
TS=$(date +%Y%m%d_%H%M%S)

systemctl stop zzedc-shiny
mv "$CORRUPT" "${CORRUPT}.corrupted.${TS}"

# Try SQLite's .recover first
sqlite3 "${CORRUPT}.corrupted.${TS}" ".recover" \
  | sqlite3 "${CORRUPT}.recovered"

if sqlite3 "${CORRUPT}.recovered" "PRAGMA integrity_check;" \
     | grep -q '^ok$'; then
  mv "${CORRUPT}.recovered" "$CORRUPT"
else
  LATEST=$(ls -t "$BACKUP_DIR"/*.db.gz | head -1)
  gunzip -c "$LATEST" | sqlite3 "$CORRUPT"
fi

chown shiny:shiny "$CORRUPT" && chmod 640 "$CORRUPT"
systemctl start zzedc-shiny
sleep 5
curl -sf http://localhost:3838/health || echo "WARN: health check failed"
```

After recovery, run `verify_audit_log_integrity()` (Chapter
10). Any chain break is a serious incident; preserve the
corrupt-database file (`.corrupted.TS`) for forensic
analysis.

### Scenario 2: PostgreSQL failure

Symptoms: connection errors, transaction failures,
replication lag.

Self-hosted, with WAL archive available:

```sh
PG_DATA=/var/lib/postgresql/14/main
BACKUP_DIR=/var/backups/zzedc/postgresql

systemctl stop postgresql
# WAL recovery on restart
systemctl start postgresql
sleep 10
pg_isready && exit 0

# Otherwise, full restore from base + WAL
systemctl stop postgresql
mv "$PG_DATA" "${PG_DATA}.failed.$(date +%F)"
LATEST_BASE=$(ls -td "$BACKUP_DIR"/base_* | head -1)
cp -r "$LATEST_BASE" "$PG_DATA"
chown -R postgres:postgres "$PG_DATA"

cat > "$PG_DATA/recovery.signal" <<EOF
EOF
cat >> "$PG_DATA/postgresql.conf" <<EOF
restore_command = 'cp $BACKUP_DIR/wal/%f %p'
recovery_target_time = '$(date -d "5 minutes ago" "+%Y-%m-%d %H:%M:%S")'
EOF

systemctl start postgresql
until pg_isready; do sleep 5; done
```

AWS RDS, restore from snapshot:

```sh
DB_INSTANCE=zzedc-production
REGION=us-east-1
NEW=${DB_INSTANCE}-restored-$(date +%F)

aws rds describe-db-snapshots \
  --db-instance-identifier "$DB_INSTANCE" \
  --region "$REGION" \
  --query 'DBSnapshots[*].[DBSnapshotIdentifier,SnapshotCreateTime]' \
  --output table

# Choose a snapshot, then:
aws rds restore-db-instance-from-db-snapshot \
  --db-instance-identifier "$NEW" \
  --db-snapshot-identifier "$SNAPSHOT_ID" \
  --db-instance-class db.t3.medium \
  --region "$REGION"

aws rds wait db-instance-available \
  --db-instance-identifier "$NEW" --region "$REGION"

NEW_ENDPOINT=$(aws rds describe-db-instances \
  --db-instance-identifier "$NEW" --region "$REGION" \
  --query 'DBInstances[0].Endpoint.Address' --output text)
# Update config.yml db_host to $NEW_ENDPOINT, restart application.
```

### Scenario 3: Complete server failure

Symptoms: server unreachable, hardware failure, cloud
instance termination.

Procedure (Docker-based deployment, AWS example):

1. Launch a replacement EC2 instance from the trial's
   golden AMI or a fresh base image.
2. Install Docker, docker-compose, and AWS CLI.
3. Pull the latest backup from the off-site bucket
   (`aws s3 cp s3://zzedc-backups/MEMORY-001/daily/ ./
   --recursive`).
4. Restore the database (`gunzip -c latest.db.gz | sqlite3
   data/zzedc.db` for SQLite, or the RDS-restore flow
   above for PostgreSQL).
5. Bring up the compose stack (`docker compose up -d`).
6. Update DNS or load-balancer target to the new host.
7. Run `verify_audit_log_integrity()` and confirm the
   application's `/health` endpoint.
8. Communicate the new URL to sites if it changed.

Capture the elapsed time as the realised RTO and record it
in the trial DMP.

### Scenario 4: Accidental or malicious data deletion

Symptoms: missing records, unexpected data changes, audit
log entries indicating DELETE actions outside expected
workflow.

Recover deleted rows from the audit log itself; the audit
trail captures the `old_values` JSON for every DELETE:

```{r}
recover_deleted_records <- function(config, table_name,
                                    deletion_time) {
  adapter <- create_db_adapter(config)
  conn <- adapter$connect()
  on.exit(adapter$disconnect(conn))

  sql <- "
    SELECT record_id, old_values, changed_at, changed_by
      FROM audit_log
     WHERE table_name = ?
       AND action = 'DELETE'
       AND changed_at >= ?
     ORDER BY changed_at
  "
  deleted <- DBI::dbGetQuery(
    conn, sql,
    params = list(table_name, as.character(deletion_time))
  )
  if (nrow(deleted) == 0) {
    message("No deleted records found in the time range")
    return(invisible(NULL))
  }
  deleted
}

restore_from_audit <- function(config, audit_records,
                               table_name) {
  adapter <- create_db_adapter(config)
  conn <- adapter$connect()
  on.exit(adapter$disconnect(conn))

  restored <- 0L
  for (i in seq_len(nrow(audit_records))) {
    old <- jsonlite::fromJSON(audit_records$old_values[i])
    cols <- names(old)
    sql <- sprintf(
      "INSERT INTO %s (%s) VALUES (%s)",
      table_name,
      paste(cols, collapse = ", "),
      paste(rep("?", length(cols)), collapse = ", ")
    )
    tryCatch({
      DBI::dbExecute(conn, sql, params = as.list(old))
      restored <- restored + 1L
    }, error = function(e) {
      warning(sprintf(
        "Failed to restore record %s: %s",
        audit_records$record_id[i], e$message
      ))
    })
  }
  message(sprintf(
    "Restored %d of %d records", restored, nrow(audit_records)
  ))
}
```

Always restore into a staging database, review with the
content author, and only then promote into production.
Each insertion is itself an audit event; the chain
remains intact.

## 9.6 Recovery checklist

```
Immediate (0-15 min)
- [ ] Confirm incident and assess scope
- [ ] Notify content author and PI
- [ ] Preserve evidence (corrupted file, logs, error output)
- [ ] Identify the most recent valid backup

Recovery (15-60 min)
- [ ] Stop affected services
- [ ] Restore from backup
- [ ] Verify data integrity (verify_audit_log_integrity)
- [ ] Test core application flows

Post-recovery (1-4 h)
- [ ] Monitor for follow-on errors
- [ ] Document the timeline
- [ ] Identify the root cause
- [ ] Update the runbook if a procedure failed
- [ ] Schedule a post-mortem
```

## 9.7 Regulatory expectations for backups

- *21 CFR Part 11.* Maintain backup logs as part of the
  audit trail; secure and access-control backup media;
  document verification in validation records; test
  recovery at least annually; retain backups for the
  trial's full retention period (typically study duration
  plus 2 years).
- *HIPAA.* Encrypt backups at rest (AES-256); encrypt
  transfers in transit (TLS 1.2+); access-control the
  backup storage; log all backup access and restoration;
  cover backup procedures in any Business Associate
  Agreement.

Document at minimum: backup schedule and retention,
recovery runbooks per scenario, DR test results,
incident reports, and the change log of the procedures
themselves. The encryption-key rotation procedure
(§5.9) belongs in the same document set.

---

# Chapter 10. Audit-log verification

The audit log is the trial's regulatory backbone. Verify it
weekly and after every restore.

## 10.1 Run the integrity check

```{r}
library(zzedc)
res <- verify_audit_log_integrity(
  db_path = "/srv/zzedc/MEMORY-001/study.db",
  key     = Sys.getenv("DB_ENCRYPTION_KEY")
)

res$valid                  # TRUE if chain validates end-to-end
res$first_audit_id         # earliest audit_id in the chain
res$last_audit_id          # latest
res$chain_length           # total entries
res$tampered_entries       # empty if chain valid
```

If `valid` is `TRUE`, every row in `audit_log` chains to its
predecessor and no entry has been tampered with since it was
written. If `valid` is `FALSE`, `tampered_entries` lists the
rows where the chain breaks. Treat any chain break as a
serious incident: stop new writes, capture the database state,
and contact your compliance lead before continuing.

## 10.2 The MIGRATION_AUDIT_GAP marker

If you ran a REDCap migration via
`import_redcap_to_zzedc_db(source = "api")` and the source
REDCap instance returned a partial or empty audit log
(because the API token lacked the `Logging` right), the
importer prepends a `MIGRATION_AUDIT_GAP` marker to the
chain. This marker is intentional: the chain still validates
end-to-end, but continuity-of-record is preserved only from
the marker forward. Inspect:

```{r}
library(DBI)
conn <- connect_encrypted_db(
  db_path = "/srv/zzedc/MEMORY-001/study.db",
  key     = Sys.getenv("DB_ENCRYPTION_KEY")
)
DBI::dbGetQuery(conn,
  "SELECT timestamp, details FROM audit_log
     WHERE event_type = 'MIGRATION_AUDIT_GAP'")
DBI::dbDisconnect(conn)
```

If the marker is present, document it in the DMP: the trial's
21 CFR Part 11 continuity argument is anchored at the marker,
not at the source-side genesis.

---

# Chapter 11. Pre-go-live checklist

Use this checklist 7 to 14 days before the planned go-live
date.

## Infrastructure

- [ ] Server provisioned with the chosen backend (Chapter 2)
- [ ] R 4.4+ installed
- [ ] zzedc installed from CRAN (or a pinned tarball)
- [ ] renv.lock committed to the trial's config repository
- [ ] config.yml configured for the trial (Chapter 4)
- [ ] Environment variables set (`DB_ENCRYPTION_KEY`,
      `DB_PASSWORD` if applicable, `ZZEDC_SALT`)
- [ ] TLS certificate in place; HTTPS redirects HTTP

## Database

- [ ] Encrypted database initialised
- [ ] `verify_database_encryption()` returns `encrypted =
      TRUE` on the production host
- [ ] Audit log initialised
- [ ] Core tables created via `create_wizard_database()` or the
      web wizard
- [ ] Admin user created; admin password changed at first
      login
- [ ] Encryption key stored in the agreed custodian (§4.3)

## Content

- [ ] Workbook (or CSV folder) imported via
      `setup_zzedc_from_gsheets()` /
      `setup_zzedc_from_csv()`
- [ ] No skipped rules outstanding (or each skipped rule
      manually addressed)
- [ ] Site list verified (Chapter 7)
- [ ] Site-based access manually verified per site

## Operations

- [ ] Daily backup cron scheduled and running
- [ ] Restore drill executed successfully
- [ ] Audit-log integrity check returns `valid = TRUE`
- [ ] Monitoring / alerting in place (disk space, container
      health, backup success/failure)
- [ ] Trial DMP updated with: server location, custodian,
      RPO/RTO, encryption-key custody, escalation path

## People

- [ ] Content author has run the dry-run protocol (their
      Chapter 4)
- [ ] Each site has been trained
- [ ] Each coordinator has logged in and changed their
      password
- [ ] Escalation path documented and shared with sites
      (coordinator -> content author -> you -> PI)

## Trial-specific

- [ ] IRB approval references the deployed CRF version
- [ ] Sponsor / DCC sign-off recorded

When every box is checked, schedule a 30-minute go-live call
with the content author and the site PIs. On the call:
confirm the URL works for everyone, confirm the credentials,
ask each site PI for go/no-go, and switch the trial banner to
*Live*.

---

# Chapter 12. Day-2 operations

The first two weeks after go-live are the highest-risk window.
After that, the operational rhythm settles.

## Daily (week 1)

- Morning: check container health (`docker compose ps`).
- Morning: check overnight backup (`ls -la backups/ | head`).
- Throughout the day: respond to escalations from the content
  author within 1 hour.
- Evening: spot-check the application URL is reachable.

## Weekly (week 2 onward)

- Run `verify_audit_log_integrity()` (§10.1). Log result.
- Check disk space on the VM (`df -h`).
- Check backup retention (most recent 30 days present).
- Review application logs for errors (`docker compose logs zzedc`).

## Monthly

- Review user roster with content author for stale accounts.
- Review the validation-rule list for additions / removals
  the trial has accumulated.
- Sample-export one subject's data and confirm it round-trips
  through the *Export* tab cleanly.
- Run a full DR test (the four-test routine of file
  integrity, restore, consistency, and recovery-time
  measurement) and append the result to the DR log.

## Quarterly

- Restore drill (§9.4).
- Key rotation review (rotate annually, but check quarterly
  for compromise indicators).
- Update R and zzedc to the latest patch versions; re-run the
  full test suite (`Rscript -e 'tinytest::run_test_dir(...)'`)
  before pushing to production.

## When to call the project maintainer

Open a GitHub issue at the project's bug tracker for: chain
breaks, application crashes, data corruption, security
findings. Include the redacted log output and the package
version. Do not include subject data.

---

# Appendix A. Migrating from REDCap

If your trial is moving from REDCap, ZZedc has a one-call
importer. The full guide is at
[`vignette('mysql-redcap-migration-roadmap')`](mysql-redcap-migration-roadmap.html).
The short version:

```{r}
# REST API path (no MySQL access needed):
result <- import_redcap_to_zzedc_db(
  source    = "api",
  api_url   = "https://redcap.example.org/api/",
  api_token = Sys.getenv("REDCAP_API_TOKEN"),
  db_path   = "/srv/zzedc/MEMORY-001/study.db",
  overwrite = TRUE
)

result$success                              # TRUE
result$audit_replay$completeness            # full / partial / empty
result$audit_replay$marker_inserted         # TRUE if not full
result$audit_replay$chain_validates         # TRUE
length(result$skipped_rules)                # rules to manually port
```

```{r}
# Live MySQL path (faster, audit log always full):
library(RMariaDB)
conn <- DBI::dbConnect(
  MariaDB(),
  host     = "redcap.example.org",
  dbname   = "redcap",
  user     = "redcap_reader",
  password = Sys.getenv("REDCAP_DB_PASSWORD"))

result <- import_redcap_to_zzedc_db(
  source    = "db",
  conn      = conn,
  pid       = 42L,
  db_path   = "/srv/zzedc/MEMORY-001/study.db",
  overwrite = TRUE
)
DBI::dbDisconnect(conn)
```

After migration, share the resulting database with the content
author and run their Chapter 4 dry-run protocol against the
imported study.

---

# Appendix B. Per-backend cheat sheets

Per-backend setup mechanics live in
[`vignette('backend-quickstart')`](backend-quickstart.html);
this appendix is the one-line `config.yml` snippet for each
plus a pointer to the relevant appendix in that vignette.

| Backend | One-line config | Setup details |
|---|---|---|
| SQLite | `db_backend: sqlite` + `db_path: ...` | `backend-quickstart` Appendix A |
| DuckDB | `db_backend: duckdb` + `db_path: ...` | `backend-quickstart` Appendix B |
| MySQL/MariaDB | `db_backend: mysql` + host/port/name/user | `backend-quickstart` Appendix C |
| PostgreSQL | `db_backend: postgresql` + host/port/name/user | `backend-quickstart` Appendix D |
| ClickHouse | `db_backend: clickhouse` + host/port/name/user | `backend-quickstart` Appendix E |

---

# Appendix C. Where the regulatory claims live in the code

If you are responding to an audit, here is a map from claim to
file:

| Claim | Implemented in |
|---|---|
| Encryption at rest (AES-256) | `R/encryption_utils.R`, `R/db_connection.R` |
| Hash-chained audit log | `R/audit_logging.R`, `R/cfr_part11_extensions.R` |
| Audit-log integrity validation | `R/audit_logging.R::verify_audit_log_integrity()` |
| Electronic signatures (21 CFR Part 11) | `R/electronic_signatures.R` |
| GDPR data-subject rights | `R/dsar.R`, `R/erasure.R`, `R/portability.R`, `R/rectification.R` |
| GDPR consent management | `R/consent.R` |
| Data retention enforcement | `R/retention.R` |
| Breach notification | `R/breach_notification.R` |
| Privacy impact assessment | `R/privacy_impact.R` |
| REDCap migration with audit replay | `R/redcap_import.R` |
| Multi-backend portability | `R/portability.R`, `R/db_adapter_*.R` |

## Mapping regulatory clauses to ZZedc behaviour

The two clauses most often cited in DMP and IRB submissions
are GDPR Article 32 and FDA 21 CFR Part 11. ZZedc's
contribution to each:

- *GDPR Article 32 (security of processing).* Encryption of
  personal data at rest (when SQLCipher-enabled RSQLite or
  TDE is in effect; see Chapter 5); access control via
  role-based authentication and per-site scoping; audit
  trail for accountability; documented operational
  procedures. The 'partial' caveat applies because
  encryption-at-rest depends on the deployed driver.
- *FDA 21 CFR Part 11 (electronic records and electronic
  signatures).* Hash-chained audit trail (immutable by
  construction; verifiable end-to-end with
  `verify_audit_log_integrity()`); electronic-signature
  framework (`R/electronic_signatures.R`); user
  accountability through audit logging tied to the
  authenticated user; data-integrity verification on
  exports (`export_encrypted_data()` produces a
  companion `.sha256` file consumed by
  `verify_exported_data()`).

Neither claim is self-attesting. The audit chain is correct
by construction but a per-database integrity claim requires
running the verifier; encryption-at-rest is correct by
construction but a per-host claim requires running
`verify_database_encryption()` against the production
binary.

The canonical compliance write-up is
`docs/compliance/gdpr-and-cfr-part11.md`. That document is
the authoritative source for the partial vs. full
classification of each control.

---

# Companion guides and references

- [Content-Author Guide](content-author-guide.html): the
  parallel guide for the non-technical team member
- [REDCap Migration Roadmap](mysql-redcap-migration-roadmap.html)
- [Demonstration Trial Setup](demonstration-trial-setup.html):
  a worked example with synthetic data
- [Advanced Features](advanced-features.html): custom
  modules, REST API, integration examples
