---
title: 'Prazosin Sleep Study (Large): Site Coordinator User Guide'
subtitle: 'Day-to-day data entry reference for site coordinators -- Protocol PRAZ-L-2026'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prazosin Sleep Study (Large): Site Coordinator User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*2026-05-26 16:46 PDT*

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for site coordinators of the Prazosin Sleep
Study (Large), protocol PRAZ-L-2026. One coordinator is assigned
per site (UCSD, UCLA, UCSF, Stanford). The study is a randomised
double-blind placebo-controlled parallel-group trial evaluating
prazosin versus placebo for sleep quality in 200 participants (50
per site). The primary outcome is change in Pittsburgh Sleep Quality
Index (PSQI) global score from baseline to week 12.

Each coordinator can view and enter data only for their assigned
site. Cross-site data are visible only to the principal investigator
(Professor Smith, username `psmith`). No knowledge of R,
databases, or ZZedc internals is required. A lab data scientist
has installed and configured the system; this guide covers all
tasks a coordinator will perform day to day.

---

## 1. Launching the Application

Open a terminal or RStudio console and run:

```{r}
library(zzedc)
launch_zzedc()
```

The application opens in your default browser at
`http://localhost:3838`. Keep the console open while you work;
closing it stops the server.

If the study is deployed on a shared server, your data scientist
will provide a URL in the form `https://<hostname>/praz-large/`.
Bookmark that URL; no R session is required on your local machine.

---

## 2. Logging In

Enter your username and password on the login screen. Usernames
are assigned by site:

| Site | Username |
|---|---|
| UCSD | `ucsd_coord` |
| UCLA | `ucla_coord` |
| UCSF | `ucsf_coord` |
| Stanford | `stanford_coord` |

On your first login you will be prompted to change the initial
password. Choose a password you will remember; it cannot be
recovered without the lab data scientist's help.

**Important:** your account is scoped to your site. Records from
other sites will not appear in your views, and the system will
reject any attempt to save a record whose `site_id` does not match
your assigned site.

If the application shows a session-timeout warning, save your
current entry before clicking away. Unsaved form fields are lost
on timeout.

---

## 3. Application Layout

The left sidebar contains seven tabs, one per data collection form:

| Tab | What it tracks |
|---|---|
| Enrollment | Screening, eligibility, and consent |
| Randomization | Treatment arm assignment |
| Study Drug | Dispensing, returns, and compliance |
| Sleep PSQI | PSQI assessments at each visit |
| Adverse Events | All AEs reported by your site's subjects |
| Concomitant Meds | Medications taken alongside study drug |
| Visits | Visit completion and subject study status |

The top navigation bar provides access to the Data Explorer
(consolidated view of all records for your site), the Reports
tab, and the Export Center.

---

## 4. Subject ID and Tracking ID Conventions

All identifiers in this study carry a site prefix to prevent
confusion across sites. Use the following conventions for your
site.

**Subject IDs**

| Site | Format | Example |
|---|---|---|
| UCSD | `PRAZ-L-S1-NNN` | `PRAZ-L-S1-001` |
| UCLA | `PRAZ-L-S2-NNN` | `PRAZ-L-S2-001` |
| UCSF | `PRAZ-L-S3-NNN` | `PRAZ-L-S3-001` |
| Stanford | `PRAZ-L-S4-NNN` | `PRAZ-L-S4-001` |

Number sequentially within your site starting from `001`. Do not
reuse numbers, even if a subject screen-fails or withdraws.

**Adverse Event IDs**

| Site | Format | Example |
|---|---|---|
| UCSD | `AE-S1-NNN` | `AE-S1-001` |
| UCLA | `AE-S2-NNN` | `AE-S2-001` |
| UCSF | `AE-S3-NNN` | `AE-S3-001` |
| Stanford | `AE-S4-NNN` | `AE-S4-001` |

**Concomitant Medication IDs**

| Site | Format | Example |
|---|---|---|
| UCSD | `MED-S1-NNN` | `MED-S1-001` |
| UCLA | `MED-S2-NNN` | `MED-S2-001` |
| UCSF | `MED-S3-NNN` | `MED-S3-001` |
| Stanford | `MED-S4-NNN` | `MED-S4-001` |

Maintain a paper or local log of the next available number for
each ID series. The system does not auto-generate these values.

---

## 5. Entering Data

### How to create a new record

1. Click the relevant tab in the left sidebar.
2. Click the **New Record** button (top right of the form panel).
3. Enter the subject ID and set `site_id` to your site before
   filling in any other fields.
4. Fill in the remaining fields. Required fields are marked with
   a red asterisk and will block the save if left empty.
5. Click **Save**. A green confirmation banner appears if the
   save succeeded. A red banner describes any validation errors.

### How to edit an existing record

Click the record in the table at the top of the tab, edit the
fields that have changed, and click **Save**.

### The site_id field

Every form requires `site_id`. Always select your site from the
dropdown; the system will reject records where `site_id` does not
match your account. This is a data integrity safeguard -- it
prevents cross-site data entry errors that are difficult to
correct after the fact.

---

### 5.1 Enrollment

Create one enrollment record when a subject signs informed
consent. Update the eligibility fields as the screening
assessment proceeds.

**Subject ID convention:** assign the next available number for
your site (e.g., `PRAZ-L-S1-007` for UCSD's seventh subject).

| Field | What to enter |
|---|---|
| Subject ID | Site-prefixed ID (see Section 4) |
| Site ID | Your assigned site |
| Consent Date | Date informed consent was signed |
| Date of Birth | Subject's date of birth |
| Sex | `male`, `female`, or `other_or_prefer_not` |
| Diagnosis | Primary sleep-related diagnosis |
| IC1: Age 18 or Older | `yes` if subject is 18 or older |
| IC2: Sleep Complaint | `yes` if subject reports a clinically significant sleep complaint |
| IC3: Stable Medications | `yes` if medications stable for at least 4 weeks |
| EC1: Alpha Blocker Use | `yes` if subject is currently taking an alpha-blocker |
| EC2: Pregnancy | `yes` if subject is currently or plans to become pregnant |
| Eligibility Status | `yes` (eligible), `no` (ineligible), or `screen_failure` |
| Screen Failure Reason | Required if Eligibility Status = `screen_failure` |
| Notes | Additional notes |

**Inclusion criteria (IC):** all three must be `yes` for
eligibility. **Exclusion criteria (EC):** both must be `no` for
eligibility. If any IC is `no` or any EC is `yes`, set Eligibility
Status to `screen_failure` and document the reason.

---

### 5.2 Randomization

Create one randomization record per eligible subject on the day
of randomization. Randomization must not precede consent.

| Field | What to enter |
|---|---|
| Subject ID | Must match an enrolled subject ID at your site |
| Site ID | Your assigned site |
| Randomization Date | Date of randomization (must be >= consent date) |
| Treatment Arm | `prazosin` or `placebo` (from the sealed allocation list) |
| Randomization Code | The code on the allocation envelope or IRT system |
| Randomized By | Your username |
| Notes | Additional notes |

**Validation:** the system will flag a randomization date that
precedes the consent date recorded in the enrollment form.

---

### 5.3 Study Drug

Create one study drug record at each dispensing visit. A single
subject will have multiple records across the study duration.

| Field | What to enter |
|---|---|
| Subject ID | Must match a randomized subject ID at your site |
| Site ID | Your assigned site |
| Visit Date | Date of the dispensing visit |
| Dose (mg) | `1`, `2`, `4`, or `6` -- the dispensed dose |
| Doses Dispensed | Number of doses given to the subject at this visit |
| Doses Returned | Number of unused doses returned (leave blank if none) |
| Compliance (%) | Calculated compliance; between 0 and 100 |
| Dose Change | `yes` if the dose was changed at this visit, `no` otherwise |
| Dose Change Reason | Required if Dose Change = `yes` |
| Notes | Additional notes |

**Compliance calculation:** `(doses_dispensed - doses_returned) /
doses_dispensed * 100`, adjusted for the number of days between
visits. Your data scientist can provide a reference worksheet if
needed.

---

### 5.4 Sleep PSQI

Create one PSQI record per assessment visit. A subject enrolled in
the full protocol will have assessments at baseline, week 4, week
8, and week 12, yielding four PSQI records.

| Field | What to enter |
|---|---|
| Subject ID | Must match a randomized subject ID at your site |
| Site ID | Your assigned site |
| Visit Date | Date the PSQI was administered |
| Visit Number | 1 (baseline) through 6 |
| PSQI 1 -- PSQI 7 | Component scores 0--3 per the PSQI scoring guide |
| PSQI Global Score | Sum of the seven component scores; must be 0--21 |
| Notes | Additional notes |

**PSQI scoring reminder:** each component is scored 0 (best)
to 3 (worst). The global score is the sum of all seven components.
A global score greater than 5 indicates poor sleep quality.
Confirm the component scores and their sum before saving; the
system validates the global range but does not recompute the sum.

---

### 5.5 Adverse Events

Create one adverse event record per reported event. The same
subject may have multiple AE records.

**AE ID convention:** assign the next available number for your
site's AE series (e.g., `AE-S1-003` for UCSD's third AE).

| Field | What to enter |
|---|---|
| Subject ID | Must match an enrolled subject ID at your site |
| Site ID | Your assigned site |
| AE ID | Site-prefixed AE ID (see Section 4) |
| Onset Date | Date the event began |
| Resolution Date | Date the event resolved; leave blank if ongoing |
| AE Term | Standardised adverse event term |
| Severity | `mild`, `moderate`, `severe`, or `life_threatening` |
| Serious | `yes` if the event meets SAE criteria, `no` otherwise |
| Relatedness | `yes`, `no`, `possibly`, or `probably` |
| Action Taken | Action taken in response to the event |
| Outcome | Current outcome of the event |
| Notes | Additional notes |

**Serious adverse events (SAEs):** if Serious = `yes`, notify
the PI (Professor Smith) immediately by telephone and submit the
required regulatory report within 24 hours. Do not wait for the
next data entry session.

**Validation:** the resolution date, if entered, must be on or
after the onset date.

---

### 5.6 Concomitant Medications

Create one record per concomitant medication per subject. A new
record is required if the dose or indication changes materially.

**Medication ID convention:** assign the next available number
for your site's medication series (e.g., `MED-S2-005` for UCLA's
fifth medication record).

| Field | What to enter |
|---|---|
| Subject ID | Must match an enrolled subject ID at your site |
| Site ID | Your assigned site |
| Medication ID | Site-prefixed medication ID (see Section 4) |
| Medication | Generic or brand name |
| Indication | Medical reason for the medication |
| Start Date | Date the medication was started |
| Stop Date | Date the medication was stopped; leave blank if ongoing |
| Dose | Dose and unit (e.g., `10 mg`) |
| Route | `oral`, `topical`, `iv`, `im`, or `other` |
| Ongoing | `yes` if still being taken, `no` if discontinued |
| Notes | Additional notes |

**Validation:** the stop date, if entered, must be on or after
the start date.

---

### 5.7 Visits

Create one visit record per scheduled visit per subject. Update
the record if the visit status changes.

| Field | What to enter |
|---|---|
| Subject ID | Must match an enrolled subject ID at your site |
| Site ID | Your assigned site |
| Visit Number | 1 through 6 |
| Visit Date | Date the visit occurred or was scheduled |
| Visit Type | `screening`, `baseline`, `week4`, `week8`, `week12`, `end_of_study`, or `early_termination` |
| Completed | `yes` if the visit occurred as scheduled, `no` otherwise |
| Reason Not Completed | Required if Completed = `no` |
| Study Status | Subject's overall status at the time of this record |
| Withdrawal Reason | Required if Study Status = `withdrawn` |
| Notes | Additional notes |

**Study status values:**

- `ongoing` -- subject is actively participating
- `completed` -- subject has completed all protocol visits
- `withdrawn` -- subject withdrew consent or was withdrawn by
  the investigator
- `lost_to_followup` -- subject cannot be reached and has missed
  two or more consecutive visits

---

## 6. Reviewing Your Site's Data

The **Data Explorer** tab provides a consolidated view of all
records for your site across all seven forms.

Useful features:

- **Filter by form** using the form-code selector to focus on a
  single domain.
- **Filter by subject** to review a single subject's complete
  record set.
- **Sort any column** by clicking the column header.
- **Search** for a keyword using the search box.

To identify subjects with missing or incomplete records, use
the Data Explorer to filter by subject ID and review which
forms have entries. A subject who has been randomized should
have at least one study drug record and one PSQI record per
completed visit.

---

## 7. Reports

The **Reports** tab provides summary statistics and listings
for your site's data. The following reports are most useful
for ongoing monitoring:

| Report | Use |
|---|---|
| Enrollment Summary | Screen-failure rate and eligibility counts |
| Randomization Balance | Treatment arm assignment counts |
| Visit Completion | Completed vs. missed visits by visit type |
| PSQI Profile | Baseline and follow-up score distributions |
| AE Summary | AE counts by severity and relatedness |
| Compliance Summary | Mean compliance percentage by visit |

Run the Enrollment Summary and Visit Completion reports at the
end of each week to confirm your site is on track with the
enrollment target of 50 subjects.

---

## 8. Exporting Data

When requested by the PI or data manager:

1. Open the **Export Center** tab.
2. Select the date range and forms to export.
3. Choose Excel or CSV format.
4. Click **Export**. The file downloads to your browser's
   default download folder.

Your export will contain only your site's records. The PI can
export a cross-site file through the Admin interface. The audit
log records every export.

---

## 9. Site Configuration Sheet

Before the study opened, your data scientist used a Google
Sheets workbook to define the study's configuration: the
**data dictionary** (one row per field, specifying each
field's name, label, and allowed values across all seven
forms) and the **validation rules** (the business rules that
ZZedc enforces when you save a record -- for example, that
a randomization date cannot precede the consent date). This
workbook is the authoritative seed source from which the
ZZedc database was initially populated.

Your data scientist has shared a copy of this workbook with
your Google account and granted you editor access. This
allows you to review and propose changes to your site's
validation rules without needing to contact the data
scientist for every small correction.

### Accessing the workbook

1. Locate the workbook link your data scientist sent. It is a
   Google Sheets URL of the form
   `https://docs.google.com/spreadsheets/d/...`.
2. Open the link in a browser and sign in with the Google
   account your data scientist registered.
3. If the browser shows a "Request access" prompt, your
   Google account has not yet been added as an editor.
   Contact your data scientist and confirm they have the
   correct address for your Google account.

### What you can edit

You have editor access to the `validation_rules` tab only.
Each row in this tab defines one business rule: the field it
applies to, the condition that must be satisfied, and the
error message shown to you if the condition is violated.
You may review these rows and, if you identify a rule that
is incorrect or missing, edit the relevant row or add a new
row at the bottom.

The workbook also contains two other tabs:

- `Data_Dictionary`: the field definitions that control which
  fields appear on each form. **Do not edit this tab.**
- `Study_Users`: the user account roster (usernames, roles,
  site assignments). **Do not edit this tab.** It may contain
  initial passwords in plain text and must remain private.

Both of those tabs are managed exclusively by the data
scientist. Any edits you make there will be overwritten
during the next import.

### When your changes take effect

Edits to the workbook do not appear in ZZedc immediately.
The workbook is a configuration source, not a live interface
to the database. A monitoring script on the study server
checks the workbook at regular intervals. When it detects a
change, it places your edit in a review queue and sends an
email notification to the coordinating centre study manager.

The manager reviews the proposed change and either approves
or rejects it:

- **Approved**: the validation rule is rebuilt in the
  database from your proposed version and takes effect at
  the next data entry session. No action is required from
  you.
- **Rejected**: the existing rule remains in place. The
  manager's rejection reason is recorded in the audit log.
  Your data scientist will contact you with details.

If you have not received any response within one working day
of making a change, contact the study manager or your data
scientist to confirm the notification was received.

### What the workbook does not contain

The workbook holds configuration only. Do not enter subject
records, visit data, PSQI scores, or any other clinical data
in the sheet. All clinical data entry occurs exclusively
through the ZZedc application (Sections 1--5).

---

## 10. Recommended Data Entry Cadence

Entering data at the time of the activity is strongly preferred.
Batch reconstruction from paper notes is error-prone and
increases the risk of query generation.

| Activity | When to enter |
|---|---|
| New screen | Same day as consent |
| Eligibility determination | Same day |
| Randomization | Same day |
| Each visit | Within 24 hours of the visit |
| Study drug dispensing | Same day |
| PSQI administration | Same day |
| Adverse event reported | Same day; SAEs within 2 hours |
| Concomitant medication change | Same day |
| Subject withdrawal | Same day |

A weekly 15-minute review of open records (subjects with
`study_status = ongoing`, AEs with blank resolution dates,
concomitant medications with `ongoing = yes`) is sufficient
to identify any entries missed during the week.

---

## 11. Troubleshooting

**The application will not launch.** Close the browser and
re-run `launch_zzedc()` in a fresh R session. If port 3838
is occupied, your data scientist can configure an alternate
port.

**A save fails with a 'site_id mismatch' error.** Confirm that
the `site_id` dropdown is set to your assigned site. The system
rejects records where `site_id` differs from your account's
site scope.

**A save fails with a validation error.** Read the red banner
message; it identifies the field and the issue. Common causes
include a date out of order (randomization before consent, AE
resolution before onset), a PSQI global score outside 0--21, or
a required field left empty.

**A subject ID does not appear in a form's subject dropdown.**
The subject must first have an enrollment record. Create the
enrollment record before entering data on other forms.

**A record is missing from the Data Explorer.** Records are not
deleted from the database; if a record was saved, it will appear
in the Data Explorer. Confirm that the form-code filter is not
hiding the relevant form.

**The site configuration sheet shows "Request access" when you
open the link.** Your Google account has not been added as an
editor. Send your data scientist the exact email address of
your Google account and ask them to re-share the workbook.

**You edited a validation rule in the sheet but ZZedc still
rejects (or accepts) the same data as before.** Edits to the
sheet take effect only after the data scientist runs a
synchronisation step against the database. Notify them and
allow one working day for the update to propagate.

**You forgot your password.** Contact the lab data scientist.
Passwords are hashed and cannot be recovered, but an
administrator can reset the account.

---

## 12. Glossary

- **Subject ID**: the unique identifier for a study participant,
  carrying a site prefix (see Section 4).
- **AE ID**: the unique identifier for an adverse event record,
  carrying a site prefix.
- **Medication ID**: the unique identifier for a concomitant
  medication record, carrying a site prefix.
- **site_id**: the site to which a record belongs. Must match
  your assigned site on every form.
- **PSQI global score**: the sum of seven component scores
  (range 0--21); higher scores indicate worse sleep quality.
- **SAE**: serious adverse event. Requires immediate notification
  of the PI and regulatory reporting within 24 hours.
- **Study status**: the subject's overall participation status
  at a given point in time (`ongoing`, `completed`, `withdrawn`,
  or `lost_to_followup`).
- **Audit log**: a system-maintained record of every change to
  the database. Runs automatically; no coordinator action required.
- **Configuration workbook**: the Google Sheets workbook that
  served as the seed source when the ZZedc database was
  initialised. It holds the data dictionary and validation rules.
  You have editor access to the `validation_rules` tab; the
  other tabs are read-only for coordinators.
- **Data dictionary**: the master list of every field in the
  study, one row per field. It specifies each field's name, label,
  type, and allowed values. Managed by the data scientist;
  visible but not editable in your configuration workbook.
- **Validation rules**: the business rules that ZZedc evaluates
  when you save a record (for example, date ordering and range
  checks). You may propose changes to these rules by editing the
  `validation_rules` tab of your configuration workbook.
