---
title: 'Prazosin Sleep Study (Large): Technical Setup Guide'
subtitle: >
  Multi-site installation, configuration, and maintenance
  for the lab data scientist -- Protocol PRAZ-L-2026
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prazosin Sleep Study (Large): Technical Setup Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

*2026-05-26 16:46 PDT*

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This guide is written for the lab data scientist responsible for
installing, configuring, and maintaining the Prazosin Sleep Study
(Large) EDC system, protocol PRAZ-L-2026. The system is a ZZedc
installation tracking a 200-subject, four-site, randomised
double-blind placebo-controlled trial evaluating prazosin versus
placebo for sleep quality. The primary outcome is change in PSQI
global score from baseline to week 12.

This deployment differs materially from a single-site
installation. Four site coordinators require separate accounts
scoped to their respective sites. The principal investigator
(Professor Smith) requires cross-site read access. Site-scoped
access is enforced at the record level through the `site_id` field,
which is required on all seven forms and is validated against the
coordinator's account `site_id`. Architectural decisions specific
to multi-site deployments are discussed throughout.

The companion document for site coordinators is
`vignette('prazosin-large-user-guide')`. After setup, distribute
that guide to each coordinator along with their credentials and
the application URL.

---

## 1. Prerequisites

- R 4.4 or later
- RStudio (recommended) or any terminal with R on `PATH`
- The `pak` package: `install.packages('pak')`
- A local working directory for the project, hereafter called
  `~/prj/praz-large/`
- For shared-server deployment (recommended for multi-site
  trials): a Linux host with R 4.4+, Shiny Server or
  Posit Connect, and HTTPS configured (see Section 3.2)

---

## 2. Installation

```{r}
pak::pak('zzedc')
```

Verify the installation:

```{r}
library(zzedc)
packageVersion('zzedc')
```

---

## 3. Deployment Architecture

### 3.1 Local deployment (single workstation)

A local SQLite deployment on one workstation is viable for
single-site pilots but is not recommended for a four-site trial.
Coordinators at remote sites would need either VPN access to the
workstation or a separately deployed instance per site. Neither
arrangement is advisable at scale.

Use local deployment only during development and testing before
the shared server is ready.

### 3.2 Shared server deployment (recommended)

Deploy a single ZZedc instance on a shared server accessible
to all four sites over HTTPS. This architecture provides:

- A single database with complete cross-site visibility for
  the PI.
- Site-scoped views for each coordinator enforced by the
  application layer.
- A centralised audit log covering all sites.
- One backup target rather than four.

Minimum server requirements: 2 CPU cores, 4 GB RAM, 20 GB
disk. Shiny Server Open Source is sufficient; Posit Connect
provides additional authentication options if your institution
has a licence.

Set `enforce_https: true` in `zzedc_config.yml` for any
deployment accessible over a network. Never transmit credentials
over HTTP.

**URL convention:** deploy at a path that identifies the
protocol, for example `https://<hostname>/praz-large/`.
Communicate this URL to all five users before the first
data entry session.

---

## 4. Project Directory

Create the project directory:

```{r}
dir.create('~/prj/praz-large', recursive = TRUE)
setwd('~/prj/praz-large')
```

On a shared server, this directory should reside on a volume
included in your institution's backup infrastructure. Do not
place it on a user's home directory if that directory is not
backed up.

---

## 5. Configuration File

Create `~/prj/praz-large/zzedc_config.yml`. Replace all
placeholder values before first use.

```yaml
study:
  name: "Prazosin Sleep Study (Large)"
  protocol_id: "PRAZ-L-2026"
  pi_name: "Professor Smith"
  pi_email: "psmith@university.edu"
  target_enrollment: 200
  phase: "Phase II"

admin:
  username: "psmith"
  fullname: "Professor Smith"
  email: "psmith@university.edu"
  password: "ChangeMeToSecurePassword123!"

security:
  session_timeout_minutes: 60
  enforce_https: true
  max_failed_login_attempts: 10

compliance:
  gdpr_enabled: false
  cfr_part11_enabled: false
  audit_logging_enabled: true

database:
  type: "sqlite"
  pool_size: 5
```

Notes on these settings:

- `pool_size: 5` accommodates five concurrent users
  (one per account). Increase if additional read-only accounts
  are added later.
- `enforce_https: true` is required for any network-accessible
  deployment. Set to `false` only during local development.
- `target_enrollment: 200` reflects the protocol-specified
  ceiling of 50 subjects per site across four sites.
- GDPR and 21 CFR Part 11 modules are disabled. The audit log
  remains active regardless.

Initialise the database:

```{r}
library(zzedc)
setwd('~/prj/praz-large')
zzedc::init(mode = 'config', config_file = 'zzedc_config.yml')
```

---

## 6. User Accounts

The study requires five accounts. Account creation is the
critical multi-site distinction from a single-site setup:
each coordinator account must have `site_id` set to its
assigned site, not `ALL`. The PI account uses `ALL` to
enable cross-site read access.

### 6.1 Study_Users.csv

The template CSV ships with the package at
`vignettes/prazosin-large/Study_Users.csv`. Copy it to your
project directory and replace the placeholder passwords before
importing:

```{r}
pkg_dir <- system.file(
  'vignettes/prazosin-large',
  package = 'zzedc'
)
file.copy(
  list.files(pkg_dir, pattern = '\\.csv$', full.names = TRUE),
  '~/prj/praz-large/'
)
```

The five accounts and their site scopes are:

| Username | Role | Site | Access scope |
|---|---|---|---|
| `psmith` | Admin | ALL | All sites; admin functions |
| `ucsd_coord` | Coordinator | UCSD | UCSD records only |
| `ucla_coord` | Coordinator | UCLA | UCLA records only |
| `ucsf_coord` | Coordinator | UCSF | UCSF records only |
| `stanford_coord` | Coordinator | Stanford | Stanford records only |

### 6.2 Site-scoped access enforcement

ZZedc enforces site scope as follows:

- On **read**: the Data Explorer and all form tables filter
  records to those whose `site_id` matches the logged-in
  user's `site_id`. The PI (`site_id = ALL`) sees all records.
- On **write**: before saving any record, the application
  validates that the record's `site_id` matches the user's
  account `site_id`. Records that fail this check are rejected
  with a descriptive error message, regardless of whether the
  validation DSL rules pass.

This dual enforcement means that a coordinator cannot accidentally
enter data under another site's ID, and cannot view another
site's data even by direct URL manipulation.

### 6.3 Distributing credentials

Distribute initial passwords to coordinators by a channel
separate from the application URL (e.g., encrypted email or
institutional password manager). Coordinators will be forced
to change their initial password on first login.

Document which coordinator account belongs to which person in
a local access log maintained outside the database.

---

## 7. Data Dictionary and Validation Rules

### 7.1 CSV import path

With the three CSV files in the project directory, import:

```{r}
library(zzedc)
setwd('~/prj/praz-large')

setup_zzedc_from_csv(
  users_csv      = 'Study_Users.csv',
  dictionary_csv = 'Data_Dictionary.csv',
  rules_csv      = 'validation_rules.csv',
  db_path        = 'data/praz-large.db'
)
```

### 7.2 Google Sheets import path

Alternatively, share a Google Sheet with the three tabs
(`Study_Users`, `Data_Dictionary`, `validation_rules`) and run:

```{r}
library(zzedc)
setwd('~/prj/praz-large')

setup_zzedc_from_gsheets(
  sheet_url = 'https://docs.google.com/spreadsheets/d/YOUR_SHEET_ID',
  db_path   = 'data/praz-large.db'
)
```

Run `googlesheets4::gs4_auth()` interactively before the
first import if you have not previously authenticated.

### 7.3 Granting coordinators write access to the workbook

After the initial import, each coordinator requires editor
access to their site's `validation_rules` tab so they can
review and propose rule changes without routing every
correction through you. The `Study_Users` and
`Data_Dictionary` tabs are import-only and should remain
under your control exclusively.

#### Per-site workbook (recommended)

Create one Google Sheets workbook per site, each containing
that site's `Study_Users`, `Data_Dictionary`, and
`validation_rules` tabs. This is the correct approach for
this trial: it constrains each coordinator's write access to
their own site's rows and prevents accidental cross-site
edits. It also keeps the `Study_Users` tab (which may carry
plain-text initial passwords) out of other coordinators'
view.

Share each workbook with the corresponding coordinator:

1. Open the site workbook in Google Sheets.
2. Click **Share**.
3. Enter the coordinator's institutional Google account
   email address.
4. Set the permission level to **Editor**.
5. Communicate the link by the same out-of-band channel
   used for initial credentials (Section 6.3). Do not
   embed the link in the ZZedc application itself.

Import each workbook individually into the shared database:

```{r}
setup_zzedc_from_gsheets(
  sheet_url = 'https://docs.google.com/spreadsheets/d/UCSD_SHEET_ID',
  db_path   = 'data/praz-large.db'
)
setup_zzedc_from_gsheets(
  sheet_url = 'https://docs.google.com/spreadsheets/d/UCLA_SHEET_ID',
  db_path   = 'data/praz-large.db'
)
# repeat for UCSF and Stanford
```

#### Single shared workbook (simpler, fewer access controls)

A single workbook containing all four sites' rows in each
tab is acceptable if the configuration content is not
sensitive. Do not use this arrangement when `Study_Users`
contains plain-text initial passwords.

If you use a shared workbook, protect the `Study_Users` and
`Data_Dictionary` tabs to prevent coordinator edits: in
Google Sheets, choose **Data > Protect sheets and ranges**,
select the tab, and restrict editing to your account only.
Leave `validation_rules` unprotected so coordinators can
edit it.

#### Which tabs coordinators may edit

| Tab | Coordinator access | Notes |
|---|---|---|
| `Study_Users` | None | Contains credentials; data scientist only |
| `Data_Dictionary` | None | Schema changes require data scientist review |
| `validation_rules` | Editor | Coordinators may propose or update rules |

#### Synchronising coordinator edits into the database

The `validation_rules` tab supports bidirectional sync.
After a coordinator notifies you of an edit, pull the
changes:

```{r}
library(zzedc)
setwd('~/prj/praz-large')

sync_dsl_rules_from_gsheets(
  sheet_id       = 'YOUR_SHEET_ID',
  db_path        = 'data/praz-large.db',
  imported_by    = 'psmith',
  delete_missing = FALSE
)
```

Set `delete_missing = TRUE` only if the sheet is the
canonical authority and rows absent from the sheet should
be removed from the database. For this trial, leave it
`FALSE` to avoid inadvertently deleting rules that a
coordinator has not yet re-entered.

`Study_Users` and `Data_Dictionary` are import-only.
Coordinator edits to those tabs (if any slip through
despite tab protection) require a full re-import via
`setup_zzedc_from_gsheets()`.

---

### 7.4 Data dictionary summary

The data dictionary defines seven forms across 79 field
definitions. Every form includes `site_id` as a required
select field with allowed values `UCSD`, `UCLA`, `UCSF`,
`Stanford`. This field is the linchpin of the multi-site
isolation mechanism and must not be removed or made optional.

### 7.5 Validation rules summary

Nineteen validation rules are defined:

- One cross-field rule: randomization date >= consent date.
- Nine PSQI range rules: global score 0--21; each of the seven
  components 0--3.
- Two conditional date-ordering rules: AE resolution >=
  onset; medication stop >= start.
- One conditional range rule: compliance percentage 0--100.
- Seven site_id membership rules: one per form, enforcing
  `site_id in(UCSD, UCLA, UCSF, Stanford)`.

The site_id rules are redundant with the application-layer
enforcement described in Section 6.2, but provide a second
layer of protection against data entry errors and any
client-side bypass.

### 7.6 Checking the import log

Review the import output before proceeding. Common issues:

| Message | Cause | Fix |
|---|---|---|
| `Skipped rule: <rule_id>` | DSL syntax error | Re-check the `expression` cell against DSL patterns in the Content-Author Guide |
| `Unknown field type: <type>` | Typo in `field_type` | Allowed types: `text`, `numeric`, `date`, `select`, `radio`, `checkbox` |
| `Tab not found: Study_Users` | Tab name mismatch | Rename to exactly `Study_Users` |
| `No rows in Data_Dictionary` | Empty row before end of data | Delete the empty row and re-import |

---

## 7a. Change Monitor: Alerting and Approval Workflow

When coordinators have editor access to the `validation_rules` tab
(Section 7.3), their changes must pass through a review gate before
they take effect in ZZedc. The change monitor:

1. Detects edits by diffing the live sheet against the database.
2. Stages proposals in a dedicated `dsl_rule_proposals` table without
   touching the live rules.
3. Notifies the study manager.
4. Rebuilds the affected rule atomically when the manager approves.

### 7a.1 How the staging table works

The staging table (`dsl_rule_proposals`) is separate from the live
`dsl_validation_rules` table. A coordinator's edit is stored as a
proposal with `status = 'PENDING'` alongside a copy of the rule's
current DSL. The live rule remains active and unchanged until the
manager calls `approve_proposals()`. On approval:

- The live `dsl_validation_rules` row is updated with the proposed
  DSL.
- The compiled R and SQL expressions are recompiled from the new DSL.
- `is_active` is set to `1` and `approval_status` to `'APPROVED'`.
- The audit log records the change, the reviewer, and the timestamp.

On rejection, the proposal is marked `'REJECTED'` and the live rule
is not modified.

### 7a.2 Polling cron job

Add a cron entry on the server that calls the poll function every
15 minutes (adjust the interval to match your trial's operational
tempo):

```
*/15 * * * * Rscript ~/prj/praz-large/poll_rules.R \
  >> ~/prj/praz-large/logs/monitor.log 2>&1
```

Create `~/prj/praz-large/poll_rules.R`:

```{r}
library(zzedc)

poll_gsheets_and_stage(
  sheet_id  = Sys.getenv('PRAZ_SHEET_ID'),
  db_path   = '~/prj/praz-large/data/praz-large.db',
  staged_by = 'system_monitor'
)
```

Set `PRAZ_SHEET_ID` in the server environment (e.g., in
`/etc/environment` or the user's `.Renviron`). Never hard-code
sheet IDs in scripts under version control.

By default the monitor writes a plain-text summary to
`~/zzedc_pending_proposals.log`. This is sufficient for
development; replace the notifier for production (see Section 7a.3).

### 7a.3 Email notification (production)

Supply a custom notifier function to `poll_gsheets_and_stage()`.
The function receives a data frame of pending proposals and is
responsible for delivery. A minimal example using `blastula`
(add to project `Suggests` or install separately):

```{r}
library(zzedc)
library(blastula)

email_notifier <- function(proposals) {
  body <- paste0(
    '**', nrow(proposals), ' validation rule proposal(s) ',
    'require your review.**\n\n',
    paste(apply(proposals, 1, function(r) {
      paste0(
        '- **[', r['change_type'], ']** `', r['rule_id'], '`  \n',
        '  Field: `', r['field_code'], '`  \n',
        '  Current: `', r['current_dsl'], '`  \n',
        '  Proposed: `', r['proposed_dsl'], '`  \n',
        '  Staged: ', r['staged_at']
      )
    }), collapse = '\n\n'),
    '\n\nLog in to ZZedc as `psmith` and call ',
    '`approve_proposals()` or `reject_proposals()` to action.'
  )

  email <- compose_email(body = md(body))

  smtp_send(
    email,
    from    = Sys.getenv('ZZEDC_FROM_EMAIL'),
    to      = Sys.getenv('ZZEDC_MANAGER_EMAIL'),
    subject = paste0('[PRAZ-L-2026] ', nrow(proposals),
                     ' rule proposal(s) pending review'),
    credentials = creds_envvar(
      user     = 'ZZEDC_SMTP_USER',
      pass_envvar = 'ZZEDC_SMTP_PASS',
      host     = Sys.getenv('ZZEDC_SMTP_HOST'),
      port     = as.integer(Sys.getenv('ZZEDC_SMTP_PORT', '587')),
      use_ssl  = FALSE
    )
  )
}

poll_gsheets_and_stage(
  sheet_id  = Sys.getenv('PRAZ_SHEET_ID'),
  db_path   = '~/prj/praz-large/data/praz-large.db',
  staged_by = 'system_monitor',
  notifier  = email_notifier
)
```

Set the five SMTP environment variables in `.Renviron` or the
server environment before the first poll run. Do not store
credentials in the script file.

### 7a.4 Manager approval

When notified, the manager reviews pending proposals:

```{r}
library(zzedc)

pending <- get_pending_proposals(
  db_path = '~/prj/praz-large/data/praz-large.db'
)
print(pending[, c('proposal_id', 'rule_id', 'change_type',
                  'current_dsl', 'proposed_dsl', 'staged_at')])
```

To approve one or more proposals:

```{r}
approve_proposals(
  ids         = c('MMSE_TOTAL_RANGE_20260601120000',
                  'AGE_RANGE_20260601120000'),
  reviewer_id = 'psmith',
  db_path     = '~/prj/praz-large/data/praz-large.db',
  comments    = 'Reviewed; consistent with protocol amendment 3'
)
```

To reject:

```{r}
reject_proposals(
  ids         = 'VISIT_WINDOW_20260601120001',
  reviewer_id = 'psmith',
  db_path     = '~/prj/praz-large/data/praz-large.db',
  comments    = 'Window tightened incorrectly; reverted to protocol spec'
)
```

Each call writes to the audit log with the reviewer's identity,
timestamp, and comments. Approved rules become active immediately;
rejected rules are flagged for the coordinator to revise in the sheet.

### 7a.5 Environment variables reference

| Variable | Purpose |
|---|---|
| `PRAZ_SHEET_ID` | Google Sheet ID for the study workbook |
| `ZZEDC_PROPOSAL_LOG` | Path for the default file-based notifier |
| `ZZEDC_MANAGER_EMAIL` | Recipient address for email notifications |
| `ZZEDC_FROM_EMAIL` | Sender address |
| `ZZEDC_SMTP_USER` | SMTP username |
| `ZZEDC_SMTP_PASS` | SMTP password |
| `ZZEDC_SMTP_HOST` | SMTP server hostname |
| `ZZEDC_SMTP_PORT` | SMTP port (default 587) |

---

## 8. Verification

### 8.1 Launch and inspect

```{r}
launch_zzedc()
```

Log in as `psmith` with the initial password from
`zzedc_config.yml`. You will be prompted to change it on
first login.

Navigate to **Admin > Forms** and confirm that all seven
forms appear: `enrollment`, `randomization`, `study_drug`,
`sleep_psqi`, `adverse_events`, `concomitant_meds`, `visits`.

Click into each form and verify:

- `site_id` is a required select field with exactly four
  options: `UCSD`, `UCLA`, `UCSF`, `Stanford`.
- All other required fields display a red asterisk.
- Select and radio fields show the correct options.

### 8.2 Site-scope dry run

Log out and log back in as each coordinator account in turn.
For each account:

1. Confirm the Data Explorer shows no records (the database
   is empty at this point).
2. Create one synthetic enrollment record with `site_id`
   set to the correct site for that account. Confirm the
   save succeeds.
3. Attempt to create a second synthetic record with
   `site_id` set to a different site. Confirm the system
   rejects the save with a site-mismatch error.

Delete all synthetic records after the dry run. Log back in
as `psmith` to confirm the admin account can see records
from all sites.

### 8.3 Validation dry run

While logged in as a coordinator, attempt to save:

- A randomization date before the subject's consent date
  (expect `RAND_AFTER_CONSENT` error).
- A PSQI global score of 22 (expect `PSQI_GLOBAL_RANGE`
  error).
- An AE resolution date before the onset date (expect
  `AE_RESOLUTION_AFTER_ONSET` error).
- A compliance percentage of 110 (expect `COMPLIANCE_RANGE`
  error).

Confirm that each produces the expected error message from
`validation_rules.csv`.

### 8.4 Audit log integrity check

```{r}
library(zzedc)
setwd('~/prj/praz-large')

verify_audit_log_integrity(db_path = 'data/praz-large.db')
```

Expected output: `Audit chain intact. N entries verified.`
Run this check after the dry run and again after any bulk
data migration.

---

## 9. Backup Strategy

A multi-site trial introduces additional backup obligations.
All five user accounts interact with a single database, so the
backup strategy is the same as for a single-site deployment
(one database file), but the consequences of data loss are
proportionally greater.

### 9.1 Manual backup

```{r}
file.copy(
  'data/praz-large.db',
  paste0(
    'backups/praz-large_',
    format(Sys.time(), '%Y%m%d'),
    '.db'
  )
)
```

### 9.2 Automated daily backup via cron (Linux server)

On the server hosting the application, add a cron entry:

```
0 2 * * * Rscript -e "
  file.copy(
    '~/prj/praz-large/data/praz-large.db',
    paste0(
      '~/prj/praz-large/backups/praz-large_',
      format(Sys.time(), '%Y%m%d'),
      '.db'
    )
  )
"
```

This runs at 02:00 server time each night. Adjust the time
to a low-traffic window for your sites' time zones. With four
sites potentially across different time zones, 02:00 UTC is
a reasonable default.

### 9.3 Off-site backup

Copy daily backup files to off-site storage (institutional
S3 bucket, institution-managed cloud backup service, or
encrypted Dropbox). A failure on the server must not destroy
the only copy of the data.

Retain backups for at least the duration of the trial plus
the regulatory retention period specified in the protocol
(typically 15 years for clinical trial data under FDA
regulations). Confirm the retention requirement with your
institution's regulatory affairs office.

### 9.4 Restore from backup

Stop the Shiny session before restoring:

```{r}
file.copy(
  'backups/praz-large_20260515.db',
  'data/praz-large.db',
  overwrite = TRUE
)
```

Restart `launch_zzedc()` after the restore completes.
Notify all five users of the restore and the data cutoff
date so they can re-enter any records lost since the last
backup.

---

## 10. Updating the Data Dictionary After Go-Live

**Adding or changing validation rules:** edit
`validation_rules.csv` and re-run the import command. The
import merges new rules and updates changed rules without
deleting existing subject data. Inform all coordinators of
any rule changes that affect their data entry workflow.

**Adding a new field:** add the row to `Data_Dictionary.csv`
and re-run the import. New fields appear as optional by
default. Existing records will show the new field as blank
until updated.

**Changing a field type** after data have been entered is not
supported by the importer. Contact the ZZedc maintainer for
a migration path.

**Modifying site_id options:** changing the allowed values in
the `site_id` validation requires a data migration for any
records already saved with the old values. Do not add or
remove sites mid-trial without a formal protocol amendment
and a documented data migration plan.

---

## 11. Directory Structure (Final State)

After setup and first use, the project directory should
contain:

```
~/prj/praz-large/
  zzedc_config.yml        # Study configuration
  Study_Users.csv         # User roster (CSV path)
  Data_Dictionary.csv     # Field definitions (CSV path)
  validation_rules.csv    # Validation rules (CSV path)
  data/
    praz-large.db         # Live SQLite database
  backups/
    praz-large_YYYYMMDD.db  # Rolling daily backups
  logs/                   # Application logs (if enabled)
```

Keep `zzedc_config.yml` in version control with the password
field replaced by a placeholder. Do not commit the `.db` file
to any repository; it contains identifiable subject data.

---

## 12. Troubleshooting

**Import fails with `googlesheets4` authentication error.**
Run `googlesheets4::gs4_auth()` interactively to authorise
access to your Google account, then re-run the import.

**The poll cron job runs but no proposals appear in
`get_pending_proposals()`.** Confirm that `PRAZ_SHEET_ID` is
set in the cron environment (cron jobs do not inherit the
user's shell environment). Add an explicit
`Sys.setenv(PRAZ_SHEET_ID = '...')` at the top of
`poll_rules.R` temporarily to isolate the issue. Also
confirm `googlesheets4::gs4_auth()` has been run
interactively at least once so the cached token is
available to the non-interactive cron process.

**`approve_proposals()` returns an error for a `proposal_id`
that appears in `get_pending_proposals()`.** The most common
cause is that the `dsl_rule_proposals` table does not yet
exist because `init_proposals_table()` has not been called.
`poll_gsheets_and_stage()` calls it automatically, but if
`stage_gsheets_proposals()` was invoked directly without
first calling `poll_gsheets_and_stage()`, run
`init_proposals_table(db_path = 'data/praz-large.db')`
once and retry.

**A coordinator reports a "Request access" prompt when
opening their workbook link.** Their Google account was not
added as an editor, or they opened the link while signed
into a different Google account. Verify the email address
they provided, re-share from the workbook's **Share**
dialog with editor access, and confirm they are using the
registered account in their browser.

**A coordinator edited the `validation_rules` tab but the
ZZedc database still enforces the old rule.** The sheet and
the database are decoupled; edits do not propagate
automatically. Run `sync_dsl_rules_from_gsheets()` (see
Section 7.3) to pull the update into the database.

**Port 3838 already in use.** Pass an alternate port:
`launch_zzedc(port = 3939)`. Update the bookmarked URL for
all coordinators.

**The database is locked.** A previous R session did not
close cleanly and left a write lock. Restart R, then check
whether any other R process is running against the same
database file before relaunching.

**A coordinator can see records from another site.** Verify
that the coordinator's `site_id` in `Study_Users.csv` is
set to exactly one of `UCSD`, `UCLA`, `UCSF`, or `Stanford`
(not `ALL`). Re-import the user table and restart the
application.

**A coordinator cannot save a record with a valid site_id.**
Confirm that the `site_id` value in the form matches the
coordinator's account `site_id` exactly, including case.
The field is case-sensitive.

**A validation rule fires on every save, even correct data.**
The most common cause is a typo in the `form_code` or
`expression` column that references a field that does not
exist. Re-check those columns against `Data_Dictionary.csv`.

**User cannot log in.** Confirm the `active` column is `TRUE`
in `Study_Users.csv`. If the account is locked after repeated
failed attempts, reset it under **Admin > Users**.

**A coordinator is replaced mid-trial.** Set the departing
coordinator's `active` to `FALSE` and create a new account
for the replacement with the same `site_id`. Do not reuse
the old username. Re-import the user table. All existing
records created under the old account remain intact; the
audit log preserves attribution.

---

## 13. Reference

For deeper detail on any of the following, consult the
corresponding vignette:

| Topic | Vignette |
|---|---|
| Google Sheets authoring in full | `vignette('content-author-guide')` |
| Server deployment, TLS, reverse proxy | `vignette('technical-lead-guide')` |
| Database backend selection | `vignette('backend-quickstart')` |
| Setup wizard (GUI alternative to config YAML) | `vignette('setup-wizard-guide')` |
| Regulatory compliance modules | `vignette('zzedc-whitepaper')` |
